---
title: "testing multivariate analyses on cities data"
subtitle: "Land use proportional area data for 40 cities on Earth"
documentclass: article
classoption: a4paper
geometry: margin=0.787in
output: 
  pdf_document: 
  highlight: espresso
  fig_height: "50%"
  fig_width: "50%"
  fig_caption: TRUE
  number_sections: TRUE
  toc: TRUE
---

# load packages

```{r load packages etc., include=FALSE}
library(stringr)
library(RcmdrMisc)
library(MASS)
library(rgr)
library(car)
library(cluster)
library(factoextra)
library(DataExplorer)

palette(c("black","blue","green4","red2","purple",
          "darkcyan","firebrick","grey","grey40","white",
          "transparent"))
```

# load data

**Data are from**: Hu, J., Wang, Y., Taubenböck, H., Zhu, X.X., 2021. Land consumption in cities: A comparative study across the globe. *Cities*, **113**: 103163, https://doi.org/10.1016/j.cities.2021.103163. (numeric from Fig. 11)

```{r}
cities <- read.csv("cities_Hu_etal_2021.csv", stringsAsFactors = TRUE)
# str(cities)
row.names(cities) <- as.character(cities$City)
cities$sType <- as.character(cities$Type)
cities$sType <- gsub("Compact-Open","CO",cities$sType)
cities$sType <- gsub("Open-Lightweight","OL",cities$sType)
cities$sType <- gsub("Compact","C",cities$sType)
cities$sType <- gsub("Open","O",cities$sType)
cities$sType <- gsub("Industrial","I",cities$sType)
cities$sType <- as.factor(cities$sType)
```

# make CLR-transformed dataset
```{r}
cities_clr <- cities
cities_clr[,c("Compact","Open","Lightweight","Industry")] <- 
  clr(cities_clr[,c("Compact","Open","Lightweight","Industry")])
names(cities);names(cities_clr)
```

# correlation matrices
```{r correlation matrix closed, fig.height=4, fig.width=3.5, fig.cap="Correlation matrix for closed variables."}
require(DataExplorer)
plot_correlation(cities[,c("Compact","Open","Lightweight","Industry")],
                 cor_args = list("use" = "pairwise.complete.obs"))
```

# correlation matrices (CLR)
```{r correlation matrix clr, fig.height=4, fig.width=3.5, fig.cap="Correlation matrix for open (CLR-transformed) variables."}
require(DataExplorer)
plot_correlation(cities_clr[,c("Compact","Open","Lightweight","Industry")],
                 cor_args = list("use" = "pairwise.complete.obs"))
```

# principal components analysis

```{r run and summarise PCA}
data0 <- na.omit(cities[,c("Compact","Open","Lightweight","Industry")])
pca_cities_clos <- prcomp(data0, scale. = TRUE)
pca_cities_clos$rot
cat("...\n\nComponent Variances\n")
pca_cities_clos$sdev^2
cat("\n--------------------\n")

data0 <- na.omit(cities_clr[,c("Compact","Open","Lightweight","Industry")])
pca_cities_open <- prcomp(data0, scale. = TRUE)
pca_cities_open$rot
cat("...\n\nComponent Variances\n")
pca_cities_open$sdev^2
cat("\n--------------------\n")
rm(data0)
```

```{r visualise PCA, fig.height=5, fig.width=10}
require(car)
palette(c("black","blue","green4","red2","purple",
          "darkcyan","firebrick","grey","grey40","white",
          "transparent"))
par(mfrow=c(1,2), mar = c(3.5,3.5,3.5,3.5), oma = c(0,0,0,0), 
    mgp=c(1.7,0.3,0), tcl = 0.25, font.lab=2,
    lend = "square", ljoin = "mitre")
# choose components and set scaling factor (sf)
v1 <- 1
v2 <- 2
sf <- 0.2

biplot(pca_cities_clos, choices = c(v1,v2), col = c(11,9), cex=c(1,0.65),
       pc.biplot = FALSE, scale = 0.4, arrow.len = 0.08,
       xlab = paste0("Scaled PC",v1," Component Loadings"),
       ylab = paste0("Scaled PC",v2," Component Loadings"))
mtext(paste0("Scaled PC",v1," Observation Scores"), 3, 1.6, font = 2)
mtext(paste0("Scaled PC",v2," Observation Scores"), 4, 1.6, font = 2)
mtext("Untransformed data\n(compositionally closed)",
      side = 3, line = -2, font = 2, adj = 0.98)
data0 <- na.omit(cities[,c("Type","Global","Region","Compact","Open","Lightweight","Industry")])
points(pca_cities_clos$x[,v1]*sf, pca_cities_clos$x[,v2]*sf*1.5,
       pch = c(22,21,24,0,1)[data0$Type],
       bg = c(2,3,4,5,6)[data0$Type],
       col = c(2,3,4,5,6)[data0$Type],
       cex = c(1.2,1.4,1.2,1.2,1.4)[data0$Type])
dataEllipse(x=pca_cities_clos$x[,v1]*sf, y=pca_cities_clos$x[,v2]*sf*1.5,
            groups = data0$Type, add = TRUE,
            plot.points = FALSE, levels = c(0.9),
            center.pch = 3, col = c(2,3,4,5,6,7),
            lty = 2, lwd = 1, center.cex = 2.4, group.labels = "")
legend("bottomright", bty = "o", inset = 0.03,
       box.col = "gray", box.lwd = 2, bg = 10,
       legend = levels(data0$Type),
       pch = c(22,21,24,0,1),
       col = c(2,3,4,5,6), pt.bg = c(2,3,4,5,6),
       pt.cex = c(1.2, 1.4, 1.2,1.2,1.4),
       cex = 0.9, y.intersp = 0.9)

# v1 <- 1
# v2 <- 3
sf <- 0.4

biplot(pca_cities_open, choices = c(v1,v2), col = c(11,9), cex=c(1,0.65),
       pc.biplot = FALSE, scale = 0.5, arrow.len = 0.08,
xlab = paste0("Scaled PC",v1," Component Loadings"),
       ylab = paste0("Scaled PC",v2," Component Loadings"))
mtext(paste0("Scaled PC",v1," Observation Scores"), 3, 1.6, font = 2)
mtext(paste0("Scaled PC",v2," Observation Scores"), 4, 1.6, font = 2)
mtext("CLR\ntransformed\ndata\n(no closure)",
      side = 3, line = -4, font = 2, adj = 0.02)
data0 <- na.omit(cities_clr[,c("Type","Compact","Open","Lightweight","Industry")])
points(pca_cities_open$x[,v1]*sf, pca_cities_open$x[,v2]*sf*1.5,
       pch = c(22,21,24,0,1)[data0$Type],
       bg = c(2,3,4,5,6)[data0$Type],
       col = c(2,3,4,5,6)[data0$Type],
       cex = c(1.2,1.4,1.2,1.2,1.4)[data0$Type])
dataEllipse(x=pca_cities_open$x[,v1]*sf, y=pca_cities_open$x[,v2]*sf*1.5,
            groups = data0$Type, add = TRUE,
            plot.points = FALSE, levels = c(0.9),
            center.pch = 3, col = c(2,3,4,5,6),
            lty = 2, lwd = 1, center.cex = 2.4, group.labels = "")
legend("bottomright", bty = "o", inset = 0.03,
       box.col = "gray", box.lwd = 2, bg = 10,
       legend = levels(data0$Type),
       pch = c(22,21,24,0,1),
       col = c(2,3,4,5,6), pt.bg = c(2,3,4,5,6),
       pt.cex = c(1.2, 1.4, 1.2,1.2,1.4),
       cex = 0.9, y.intersp = 0.9)
rm(list = c("v1","v2","sf","data0"))
```

# K-means clustering

```{r assess optimum clusters for closed data, fig.height=3, fig.width=4.}
require(factoextra)
data0 <- na.omit(cities[,c("Compact","Open","Lightweight","Industry")])
fviz_nbclust(data0, kmeans, method = "wss") +
  geom_vline(xintercept = 4, linetype = 2)
```

```{r assess optimum clusters for CLR data, fig.height=3, fig.width=4}
require(factoextra)
data0 <- na.omit(cities_clr[,c("Compact","Open","Lightweight","Industry")])
fviz_nbclust(data0, kmeans, method = "wss") +
  geom_vline(xintercept = 4, linetype = 2)
```

# compute K-means closed

```{r}
data0 <- na.omit(cities[,c("sType","Compact","Open","Lightweight","Industry")])
data0[,c("Compact","Open","Lightweight","Industry")] <- scale(data0[,c("Compact","Open","Lightweight","Industry")])
set.seed(123)
cities_clos_kmeans <- kmeans(data0[,2:NCOL(data0)], 4, nstart = 25)
cat("K-means clustering with",length(cities_clos_kmeans$size),"clusters of sizes",cities_clos_kmeans$size,"\n")
cat("\ncomponents of output object are:\n")
ls(cities_clos_kmeans)
```

# plot kmeans clusters closed

```{r visualize kmeans clusters closed, fig.height=5, fig.width=6, message=FALSE, warning=FALSE}
row.names(data0) <- paste0(data0$sType,seq(1,NROW(data0)))
fviz_cluster(cities_clos_kmeans, data = data0[,2:NCOL(data0)],
             palette = c("#00AFBB", "#E7B800", "#FC4E07","purple2"),
             labelsize=10,
             ellipse.type = "euclid", # Concentration ellipse
             star.plot = TRUE, # Add segments from centroids to items
             repel = F, # if true avoids label overplotting (slow)
             ggtheme = theme_minimal()
             )
```

# compute K-means open

```{r}
data0 <- na.omit(cities_clr[,c("sType","Compact","Open","Lightweight","Industry")])
data0[,c("Compact","Open","Lightweight","Industry")] <- scale(data0[,c("Compact","Open","Lightweight","Industry")])
set.seed(123)
cities_open_kmeans <- kmeans(data0[,2:NCOL(data0)], 4, nstart = 25)
cat("K-means clustering with",length(cities_open_kmeans$size),"clusters of sizes",cities_open_kmeans$size,"\n")
cat("\ncomponents of output object are:\n")
ls(cities_open_kmeans)
```

# plot kmeans clusters open

```{r visualize kmeans clusters open, fig.height=6, fig.width=6, message=FALSE, warning=FALSE}
row.names(data0) <- paste0(data0$sType,seq(1,NROW(data0)))
fviz_cluster(cities_open_kmeans, data = data0[,2:NCOL(data0)],
             palette = c("#00AFBB", "#E7B800", "#FC4E07","purple2"),
             labelsize=10,
             ellipse.type = "euclid", # Concentration ellipse
             star.plot = TRUE, # Add segments from centroids to items
             repel = F, # if true avoids label overplotting (slow)
             ggtheme = theme_minimal()
             )
```

# hierarchical clustering

## create dissimilarity (distance) matrix

```{r distance matrix closed, message=FALSE, warning=FALSE}
dataHC <- na.omit(cities[,c("sType","Compact","Open","Lightweight","Industry")])
row.names(dataHC) <- paste0(dataHC$sType, seq(1,NROW(dataHC)))
dataHC$sType <- NULL
cities_clos_diss <- get_dist(dataHC, method = "euclidean")
as.matrix(cities_clos_diss)[1:8, 1:8]
```

## perform hierarchical clustering

```{r hierarchical clustering closed, fig.height=4.5, fig.width=10, message=FALSE, warning=FALSE}
cities_clos_hc <- hclust(cities_clos_diss, method = "ward.D2")
require(factoextra)
fviz_dend(cities_clos_hc, cex = 0.5)
```

## assess cluster tree

```{r verify cluster tree closed}
cities_clos_coph <- cophenetic(cities_clos_hc)
cor(cities_clos_diss,cities_clos_coph)
cat("\nRule-of-thumb: cluster tree represents actual distance matrix accurately enough if r>0.75\n")
```

## cut dendrogram into different groups

```{r cut dendrogram closed}
cities_clos_grp <- cutree(cities_clos_hc, k = 5)
cities_clos_grp ; cat("\n")
table(cities_clos_grp)
```

## plot dendrogram with cuts

```{r plot cut dendrogram closed, fig.height=4.5, fig.width=10, message=FALSE, warning=FALSE}
fviz_dend(cities_clos_hc, k = 5, # Cut in five groups
          cex = 0.75, # label size
          k_colors = c("red3", "blue2", "purple","sienna","grey40"),
          color_labels_by_k = TRUE, # color labels by groups
          rect = TRUE # Add rectangle around groups
          )
```

# hierarchical clustering (CLR-open data)

## create dissimilarity (distance) matrix

```{r distance matrix open, message=FALSE, warning=FALSE}
dataHC <- na.omit(cities_clr[,c("sType","Compact","Open","Lightweight","Industry")])
row.names(dataHC) <- paste0(dataHC$sType, seq(1,NROW(dataHC)))
dataHC$Type <- NULL
cities_open_diss <- get_dist(dataHC, method = "euclidean")
as.matrix(cities_open_diss)[1:8, 1:8]
```

## perform hierarchical clustering

```{r hierarchical clustering open, fig.height=4.5, fig.width=10, message=FALSE, warning=FALSE}
cities_open_hc <- hclust(cities_open_diss, method = "ward.D2")
require(factoextra)
fviz_dend(cities_open_hc, cex = 0.5)
```

## assess cluster tree

```{r verify cluster tree open}
cities_open_coph <- cophenetic(cities_open_hc)
cor(cities_open_diss,cities_open_coph)
cat("\nRule-of-thumb: cluster tree represents actual distance matrix accurately enough if r>0.75\n")
```

## cut dendrogram into different groups

```{r cut dendrogram open}
cities_open_grp <- cutree(cities_open_hc, k = 5)
cities_open_grp ; cat("\n")
table(cities_open_grp)
```

## plot dendrogram with cuts

```{r plot cut dendrogram open, fig.height=4.5, fig.width=10, message=FALSE, warning=FALSE}
fviz_dend(cities_open_hc, k = 5, # Cut in five groups
          cex = 0.75, # label size
          k_colors = c("red3", "blue2", "purple"),
          color_labels_by_k = TRUE, # color labels by groups
          rect = TRUE # Add rectangle around groups
)
```

# LDA linear discriminant analysis

```{r LDA closed cities data}
data0 <- cities
data0[,c("Compact", "Open", "Lightweight", "Industry")] <- scale(data0[,c("Compact", "Open", "Lightweight", "Industry")])
lda_cities_clos <- lda(formula = Type ~ Compact + Open + Lightweight + Industry, 
                    data=data0,
                    prior=as.numeric(summary(cities$Type))/
                      nrow(cities)) 
print(lda_cities_clos)
```

```{r plot LDA closed, fig.height=5, fig.width=10}
par(mfrow = c(1,2), mar = c(3.5,3.5,1,1), mgp = c(1.5,0.3,0), tcl = 0.25,
    lend = "square", ljoin = "mitre", cex.main = 0.9, font.lab=2)
plot(lda_cities_clos$scaling[,1], lda_cities_clos$scaling[,2], 
     xlim = c(-0.5,3), ylim=c(-2,2),
     xlab="Linear Discriminant [1]", ylab="Linear Discriminant [2]", main="(a) Variable Coefficients")
abline(v=0,col="grey",lty=2)
abline(h=0,col="grey",lty=2)
text(lda_cities_clos$scaling[,1],lda_cities_clos$scaling[,2],
     labels=c("Compact", "Open", "Lightweight","Industry"),
     pos=1,cex=0.95,col="blue2",offset=0.2)

ldaPred_cities_clos <- predict(lda_cities_clos)
for(i in 1:NROW(lda_cities_clos$scaling)){
  arrows(0,0,lda_cities_clos$scaling[i,1],lda_cities_clos$scaling[i,2],
         length = 0.15, col = 8)
}
plot(ldaPred_cities_clos$x[,1], ldaPred_cities_clos$x[,2], 
     col=c(2,"#00666680",8,"#99660080","grey")[cities$Type], 
     pch=c(0,15,1,19,3)[cities$Type], xlim = c(-3,14),
  lwd=c(2,1,2,1,2)[cities$Type],
  cex=c(1.8,1.8,2,2,2)[cities$Type], 
  xlab="Linear Discriminant [1]", ylab="Linear Discriminant [2]", 
  main="(b) Predictions for Observations")
abline(v=0,col="grey",lty=2)
abline(h=0,col="grey",lty=2)
text(ldaPred_cities_clos$x[,1], ldaPred_cities_clos$x[,2], 
     labels=cities$Type, col=c(2,"blue4","#991100","#663300",8)[cities$Type],
     pos=1, offset=0.15, cex=0.65)
legend("topright",legend=levels(cities$Type),
       col=c(2,"#00666680",8,"#99660080","grey"),
       pch=c(0,15,1,19,3), pt.lwd=c(2,1,2,1,2),
       title="City Type",bty="o", box.col="grey90", 
       box.lwd=2, inset=0.02, 
       pt.cex=c(1.8,1.8,2,2,2), cex=1.)

```

```{r plot all LDA dimensions closed, fig.height=4.5, fig.width=12}
par(mfrow = c(1,3), mar = c(3.5,3.5,1,1), mgp = c(1.5,0.3,0), tcl = 0.25,
    lend = "square", ljoin = "mitre", cex.main = 0.9, font.lab=2)

plot(ldaPred_cities_clos$x[,1], ldaPred_cities_clos$x[,2], col=c(2,4,6,8,1)[cities$Type],
     pch=c(0,15,1,19,3)[cities$Type], xlim = c(-4,14),lwd=c(2,1,2,1,2)[cities$Type],
     cex=c(1.8,1.8,2,2,2)[cities$Type], main="(b) Predictions for Observations", 
     xlab="Linear Discriminant [1]", ylab="Linear Discriminant [2]")
abline(v=0,col="grey",lty=2); abline(h=0,col="grey",lty=2)
text(ldaPred_cities_clos$x[,1], ldaPred_cities_clos$x[,2], labels=cities$Type, 
     col=c(2,4,6,8,1)[cities$Type], pos=1, offset=0.15, cex=0.65)
legend("topright", legend=levels(cities$Type), col=c(2,4,6,8,1),
       pch=c(0,15,1,19,3), pt.lwd=c(2,1,2,1,2),
       title="City Type",bty="o", box.col="grey90", 
       box.lwd=2, inset=0.02, pt.cex=c(1.8,1.8,2,2,2), cex=0.9)

plot(ldaPred_cities_clos$x[,1], ldaPred_cities_clos$x[,3], col=c(2,4,6,8,1)[cities$Type],
     pch=c(0,15,1,19,3)[cities$Type], xlim = c(-4,14),lwd=c(2,1,2,1,2)[cities$Type],
     cex=c(1.8,1.8,2,2,2)[cities$Type], main="(b) Predictions for Observations", 
     xlab="Linear Discriminant [1]", ylab="Linear Discriminant [3]")
abline(v=0,col="grey",lty=2); abline(h=0,col="grey",lty=2)
text(ldaPred_cities_clos$x[,1], ldaPred_cities_clos$x[,3], labels=cities$Type, 
     col=c(2,4,6,8,1)[cities$Type], pos=1, offset=0.15, cex=0.65)

plot(ldaPred_cities_clos$x[,2], ldaPred_cities_clos$x[,3], col=c(2,4,6,8,1)[cities$Type],
     pch=c(0,15,1,19,3)[cities$Type], xlim = c(-4,6),lwd=c(2,1,2,1,2)[cities$Type],
     cex=c(1.8,1.8,2,2,2)[cities$Type], main="(b) Predictions for Observations", 
     xlab="Linear Discriminant [2]", ylab="Linear Discriminant [3]")
abline(v=0,col="grey",lty=2); abline(h=0,col="grey",lty=2)
text(ldaPred_cities_clos$x[,2], ldaPred_cities_clos$x[,3], labels=cities$Type, 
     col=c(2,4,6,8,1)[cities$Type], pos=1, offset=0.15, cex=0.65)

par(mfrow=c(1,1))
```

```{r LDA open cities data}
data0 <- cities_clr
data0[,c("Compact", "Open", "Lightweight", "Industry")] <- scale(data0[,c("Compact", "Open", "Lightweight", "Industry")])
lda_cities_open <- lda(formula = Type ~ Compact + Open + Lightweight + Industry, 
                    data=data0,
                    prior=as.numeric(summary(data0$Type))/nrow(data0)) 
print(lda_cities_open)
cormat <- rcorr.adjust(data0[,c("Compact", "Open", "Lightweight", "Industry")], type="pearson")
cat("\nCorrelation matrix for predictors:\n"); print(cormat$R$r, digits = 3)
rm(cormat)
```

```{r plot LDA open, fig.height=5, fig.width=10}
par(mfrow = c(1,2), mar = c(3.5,3.5,1,1), mgp = c(1.5,0.3,0), tcl = 0.25,
    lend = "square", ljoin = "mitre", cex.main = 0.9, font.lab=2)
plot(lda_cities_open$scaling[,1], lda_cities_open$scaling[,2], xlim = c(-4,2), ylim = c(-1.2,2), 
     xlab="Linear Discriminant [1]", ylab="Linear Discriminant [2]", main="(a) Variable Coefficients")
abline(v=0,col="grey",lty=2); abline(h=0,col="grey",lty=2)
text(lda_cities_open$scaling[,1], lda_cities_open$scaling[,2], labels=names(cities)[2:5],
     pos=1,cex=0.95,col="blue2",offset=0.2)
for(i in 1:NROW(lda_cities_open$scaling)){
  arrows(0,0,lda_cities_open$scaling[i,1],lda_cities_open$scaling[i,2],
         length = 0.15, col = 8)
}
ldaPred_cities_open <- predict(lda_cities_open)

plot(ldaPred_cities_open$x[,1], ldaPred_cities_open$x[,2], col=c(2,4,6,8,1)[cities$Type],
     pch=c(0,15,1,19,3)[cities$Type], xlim = c(-14,7),lwd=c(2,1,2,1,2)[cities$Type],
     cex=c(1.8,1.8,2,2,2)[cities$Type], main="(b) Predictions for Observations", 
     xlab="Linear Discriminant [1]", ylab="Linear Discriminant [2]")
abline(v=0,col="grey",lty=2); abline(h=0,col="grey",lty=2)
text(ldaPred_cities_open$x[,1], ldaPred_cities_open$x[,2], labels=cities$Type, 
     col=c(2,4,6,8,1)[cities$Type], pos=1, offset=0.15, cex=0.65)
legend("topleft", legend=levels(cities$Type), col=c(2,4,6,8,1),
       pch=c(0,15,1,19,3), pt.lwd=c(2,1,2,1,2),
       title="City Type",bty="o", box.col="grey90", 
       box.lwd=2, inset=0.02, pt.cex=c(1.8,1.8,2,2,2), cex=0.9)
par(mfrow=c(1,1))
```

```{r plot all LDA dimensions open, fig.height=4.5, fig.width=12}
par(mfrow = c(1,3), mar = c(3.5,3.5,1,1), mgp = c(1.5,0.3,0), tcl = 0.25,
    lend = "square", ljoin = "mitre", cex.main = 0.9, font.lab=2)

plot(ldaPred_cities_open$x[,1], ldaPred_cities_open$x[,2], col=c(2,4,6,8,1)[cities$Type],
     pch=c(0,15,1,19,3)[cities$Type], xlim = c(-14,7),lwd=c(2,1,2,1,2)[cities$Type],
     cex=c(1.8,1.8,2,2,2)[cities$Type], main="(b) Predictions for Observations", 
     xlab="Linear Discriminant [1]", ylab="Linear Discriminant [2]")
abline(v=0,col="grey",lty=2); abline(h=0,col="grey",lty=2)
text(ldaPred_cities_open$x[,1], ldaPred_cities_open$x[,2], labels=cities$Type, 
     col=c(2,4,6,8,1)[cities$Type], pos=1, offset=0.15, cex=0.65)
legend("topleft", legend=levels(cities$Type), col=c(2,4,6,8,1),
       pch=c(0,15,1,19,3), pt.lwd=c(2,1,2,1,2),
       title="City Type",bty="o", box.col="grey90", 
       box.lwd=2, inset=0.02, pt.cex=c(1.8,1.8,2,2,2), cex=0.9)

plot(ldaPred_cities_open$x[,1], ldaPred_cities_open$x[,3], col=c(2,4,6,8,1)[cities$Type],
     pch=c(0,15,1,19,3)[cities$Type], xlim = c(-14,7),lwd=c(2,1,2,1,2)[cities$Type],
     cex=c(1.8,1.8,2,2,2)[cities$Type], main="(b) Predictions for Observations", 
     xlab="Linear Discriminant [1]", ylab="Linear Discriminant [3]")
abline(v=0,col="grey",lty=2); abline(h=0,col="grey",lty=2)
text(ldaPred_cities_open$x[,1], ldaPred_cities_open$x[,3], labels=cities$Type, 
     col=c(2,4,6,8,1)[cities$Type], pos=1, offset=0.15, cex=0.65)

plot(ldaPred_cities_open$x[,2], ldaPred_cities_open$x[,3], col=c(2,4,6,8,1)[cities$Type],
     pch=c(0,15,1,19,3)[cities$Type], xlim = c(-14,7),lwd=c(2,1,2,1,2)[cities$Type],
     cex=c(1.8,1.8,2,2,2)[cities$Type], main="(b) Predictions for Observations", 
     xlab="Linear Discriminant [2]", ylab="Linear Discriminant [3]")
abline(v=0,col="grey",lty=2); abline(h=0,col="grey",lty=2)
text(ldaPred_cities_open$x[,2], ldaPred_cities_open$x[,3], labels=cities$Type, 
     col=c(2,4,6,8,1)[cities$Type], pos=1, offset=0.15, cex=0.65)

par(mfrow=c(1,1))
```

```{r eval=FALSE, include=FALSE}
data3d <- as.data.frame(lda_cities_open$scaling)
scatter3d(ldaPred_cities_open$x[,1], ldaPred_cities_open$x[,2], ldaPred_cities_open$x[,3],
          groups = ldaPred_cities_open$class,
          point.col = c(2,3,4,5,6), surface = F,
          labels = cities_clr$City)
```

```{r compare models with reality, paged.print=FALSE}
closComp <- as.data.frame(cbind(as.character(cities_clr$Type), 
                                as.character(ldaPred_cities_clos$class)))
colnames(closComp) <- c("=-Actual-=","=-Predicted-=")
closComp$test <- as.character(cities_clr$Type) == as.character(ldaPred_cities_clos$class)
k = length(which(closComp$test == TRUE))
cat("Predictions by LDA using closed data:",k,"out of",NROW(cities_clr),"=",paste0(100*k/NROW(cities_clr),"% correct\n"))
closComp

openComp <- as.data.frame(cbind(as.character(cities_clr$Type), as.character(ldaPred_cities_open$class)))
colnames(openComp) <- c("=-Actual-=","=-Predicted-=")
openComp$test <- as.character(cities_clr$Type) == as.character(ldaPred_cities_open$class)
k = length(which(openComp$test == TRUE))
cat("\nPredictions by LDA using open data:",k,"out of",NROW(cities_clr),"=",paste0(100*k/NROW(cities_clr),"% correct\n"))
openComp
```

```{r LDA train and predict closed, message=FALSE, warning=FALSE, paged.print=FALSE}
library(MASS)
library(RcmdrMisc)
library(knitr)
library(beepr)
### MATCH this to read file below
# sink(file="cities_type_pred.csv", type="output")
n0 <- 1000 # number of iterations
ftrain <- 0.75 # proportion of observations in training set
results <- data.frame(
    Rep = rep(NA, n0),
    matches = rep(NA, n0),
    non_matches = rep(NA, n0),
    success = rep(NA, n0))
# train <- sample(1:NROW(cities), round(NROW(cities)-5,0))
train <- sample(1:NROW(cities), round(NROW(cities) * ftrain,0))
lda.cities.train <- lda(formula = Type ~  Compact + Open + Lightweight + Industry, 
                    data = cities[train,],
                    prior = as.numeric(summary(cities$Type[train]))/
                      nrow(cities[train,])) 
lda.cities.pred <- predict(lda.cities.train, cities[-train,])
# make vector of individual category non-matches
matchByClass <- data.frame(Matches = rep(0,nlevels(cities$Type))) 
rownames(matchByClass) <- levels(lda.cities.pred$class)
e0 <- 0
# colnames(matchByClass) <- c("Matches")
for (i in 1:n0) {
  # train <- sample(1:NROW(cities), round(NROW(cities)-5,0))
  train <- sample(1:NROW(cities), round(NROW(cities) * ftrain,0))
  if (is.na(match(NA,tapply(cities[train,]$Open, cities[train,]$Type, sd, na.rm=T))) == TRUE) {
    lda.cities.train <- lda(formula = Type ~ Compact + Open + Lightweight + Industry, 
                          data = cities[train,],
                          prior=as.numeric(summary(cities$Type[train]))/
                        nrow(cities[train,])) 
    lda.cities.pred <- predict(lda.cities.train, cities[-train,])
    e0 <- e0 + 1
  }

  k=0 # number of non-matches
  m0 <- as.matrix(rep(0,5)) # vector of individual category non-matches
  rownames(m0) <- levels(lda.cities.pred$class)
  rownames(matchByClass) <- levels(lda.cities.pred$class)
  colnames(matchByClass) <- c("Matches")
  for (jM in 1:NROW(cities[-train,])) {
    # cat("big loop #",jM,"\n")
    for (jS in 1:nlevels(lda.cities.pred$class)) {
      if((lda.cities.pred$class[jM] == levels(lda.cities.pred$class)[jS]) & 
         (cities$Type[-train][jM] == levels(lda.cities.pred$class)[jS]) ) 
        m0[jS] = m0[jS] + 1
      else  m0[jS] = m0[jS] 
      # cat("small loop iteration #",jS,"; matching",
      #     levels(lda.cities.pred$class)[jS],
      #     "; matches =",m0,"\n")
    }
    k = sum(m0)
    # cat("medium loop iteration ",jM,"; matches = ",k,"\n", sep="")
    # if(jM==NROW(cities[-train,])) 
  }
  # cat("GIANT LOOP #",i,"\n")
  matchByClass <- matchByClass + m0
  # output to results data frame: iteration, matches, non-matches, proportion matched
  results[i,] <- c(i, k, NROW(cities[-train,])-k, signif(k/NROW(cities[-train,]),3))
  # cbind(lda.cities.pred$class,cities$Type[-train])
}
matchByClass$Actual <- round(1000*as.numeric(summary(cities$Type))*
                               (NROW(cities[-train,])/NROW(cities)),0)
matchByClass$Proportion <- matchByClass$Matches/matchByClass$Actual

# sink() # close output file
### make sure read file is SAME AS SINK
beep(sound = 10)

# results <- read.csv("cities_type_pred.csv")
{cat("[Based on", n0, "random subsets of dataset to",
    "train LDA model to\npredict remaining observations]\n")
cat("Number of obs. in random subsets =",NROW(train),
    " (predicting",NROW(cities)-NROW(train),"samples)\n")
print(numSummary(results[,2:4], statistics=c("mean","sd"))$table)
ns0 <- numSummary(results$success)
t0 <- t.test(results$success)
cat(rep("-\u2013-",24),
    "\nStat. summary for 'success':\nMean = ",round(ns0$table[1],4),
    ", sd = ",round(ns0$table[2],4),", 95% confidence interval = (",
    signif(t0$conf.int[1],3),", ",signif(t0$conf.int[2],4),") (after ",i," reps)\n", sep="")
cat(n0-e0,"iterations failed due to random sampling missing a group\n\n")
print(matchByClass, digits=3)}
rm(list = c("n0","ftrain","i","e0","jS","jM","k","m0","matchByClass","results","train"))
```

```{r LDA train and predict open, message=FALSE, warning=FALSE, , paged.print=FALSE}
n0 <- 1000 # number of iterations
ftrain <- 0.75 # proportion of observations in training set
results <- data.frame(
  Rep = rep(NA, n0),
  matches = rep(NA, n0),
  non_matches = rep(NA, n0),
  success = rep(NA, n0))
# train <- sample(1:NROW(cities_clr), round(NROW(cities_clr)-5,0))
train <- sample(1:NROW(cities_clr), round(NROW(cities_clr) * ftrain,0))
lda.cities_clr.train <- lda(formula = Type ~  Compact + Open + Lightweight + Industry, 
                        data = cities_clr[train,],
                        prior = as.numeric(summary(cities_clr$Type[train]))/
                          nrow(cities_clr[train,])) 
lda.cities_clr.pred <- predict(lda.cities_clr.train, cities_clr[-train,])
# make vector of individual category non-matches
matchByClass <- data.frame(Matches = rep(0,nlevels(cities_clr$Type))) 
rownames(matchByClass) <- levels(lda.cities_clr.pred$class)
e0 <- 0
# colnames(matchByClass) <- c("Matches")
for (i in 1:n0) {
  # train <- sample(1:NROW(cities_clr), round(NROW(cities_clr)-5,0))
  train <- sample(1:NROW(cities_clr), round(NROW(cities_clr) * ftrain,0))
  if (is.na(match(NA,tapply(cities_clr[train,]$Open, cities_clr[train,]$Type, sd, na.rm=T))) == TRUE) {
    lda.cities_clr.train <- lda(formula = Type ~ Compact + Open + Lightweight + Industry, 
                            data = cities_clr[train,],
                            prior=as.numeric(summary(cities_clr$Type[train]))/
                              nrow(cities_clr[train,])) 
    lda.cities_clr.pred <- predict(lda.cities_clr.train, cities_clr[-train,])
    e0 <- e0 + 1
  }
  
  k=0 # number of non-matches
  m0 <- as.matrix(rep(0,5)) # vector of individual category non-matches
  rownames(m0) <- levels(lda.cities_clr.pred$class)
  rownames(matchByClass) <- levels(lda.cities_clr.pred$class)
  colnames(matchByClass) <- c("Matches")
  for (jM in 1:NROW(cities_clr[-train,])) {
    # cat("big loop #",jM,"\n")
    for (jS in 1:nlevels(lda.cities_clr.pred$class)) {
      if((lda.cities_clr.pred$class[jM] == levels(lda.cities_clr.pred$class)[jS]) & 
         (cities_clr$Type[-train][jM] == levels(lda.cities_clr.pred$class)[jS]) ) 
        m0[jS] = m0[jS] + 1
      else  m0[jS] = m0[jS] 
      # cat("small loop iteration #",jS,"; matching",
      #     levels(lda.cities_clr.pred$class)[jS],
      #     "; matches =",m0,"\n")
    }
    k = sum(m0)
    # cat("medium loop iteration ",jM,"; matches = ",k,"\n", sep="")
    # if(jM==NROW(cities_clr[-train,])) 
  }
  # cat("GIANT LOOP #",i,"\n")
  matchByClass <- matchByClass + m0
  # output to results data frame: iteration, matches, non-matches, proportion matched
  results[i,] <- c(i, k, NROW(cities_clr[-train,])-k, signif(k/NROW(cities_clr[-train,]),3))
  # cbind(lda.cities_clr.pred$class,cities_clr$Type[-train])
}
matchByClass$Actual <- round(n0*as.numeric(summary(cities_clr$Type))*
                               (NROW(cities_clr[-train,])/NROW(cities_clr)),0)
matchByClass$Proportion <- matchByClass$Matches/matchByClass$Actual

beep(sound = 10)

{cat("[Based on", n0, "random subsets of dataset to",
     "train LDA model to\npredict remaining observations]\n")
  cat("Number of obs. in random subsets =",NROW(train),
      " (predicting",NROW(cities_clr)-NROW(train),"samples)\n")
  print(numSummary(results[,2:4], statistics=c("mean","sd"))$table)
  ns0 <- numSummary(results$success)
  t0 <- t.test(results$success)
  cat(rep("-\u2013-",24),
      "\nStat. summary for 'success':\nMean = ",round(ns0$table[1],4),
      ", sd = ",round(ns0$table[2],4),", 95% confidence interval = (",
      signif(t0$conf.int[1],3),", ",signif(t0$conf.int[2],4),") (after ",i," reps)\n", sep="")
  cat(n0-e0,"iterations failed due to random sampling missing a group\n\n")
  print(matchByClass, digits=3)}
rm(list = c("n0","ftrain","i","e0","jS","jM","k","m0","matchByClass","results","train"))
```
