---
documentclass: article
classoption: a4paper
geometry: margin=0.6in
output: 
  pdf_document: 
    fig_caption: TRUE
    number_sections: TRUE
    toc: no
    highlight: tango
  html_document: 
    toc_depth: 2
    fig_caption: yes
    number_sections: yes
    self_contained: no
  word_document: 
    toc_depth: 2
    fig_caption: yes
fontsize: 12pt
header-includes:
  \usepackage{sourcesanspro}
  \usepackage[T1]{fontenc}
  \renewcommand{\familydefault}{\sfdefault}
  \renewcommand{\thefigure}{4.\arabic{figure}}
  \renewcommand{\thetable}{4.\arabic{table}}
---

\setcounter{section}{3}

```{r load knitr, include=FALSE}
library(flextable)
library(magrittr)
library(png)

set_flextable_defaults(theme_fun = "theme_zebra", 
                       font.size = 9, fonts_ignore = TRUE)
BorderDk <- officer::fp_border(color = "#B5C3DF", style = "solid", width = 1)
BorderLt <- officer::fp_border(color = "#FFFFFF", style = "solid", width = 1)

palette(c("black", "#003087", "#DAAA00", "#8F92C4", "#E5CF7E", 
          "#001D51", "#B7A99F", "#A51890", "#C5003E", "#FDC596", 
          "#AD5D1E", "gray40", "gray85", "#FFFFFF", "transparent"))

addImg <- function(obj, x = NULL, y = NULL, width = NULL, interpolate = TRUE){
  if(is.null(x) | is.null(y) | is.null(width)){stop("Must provide args 'x', 'y', and 'width'")}
  USR <- par()$usr ; PIN <- par()$pin ; DIM <- dim(obj) ; ARp <- DIM[1]/DIM[2]
  WIDi <- width/(USR[2]-USR[1])*PIN[1] ;   HEIi <- WIDi * ARp 
  HEIu <- HEIi/PIN[2]*(USR[4]-USR[3]) 
  rasterImage(image = obj, xleft = x-(width/2), xright = x+(width/2),
            ybottom = y-(HEIu/2), ytop = y+(HEIu/2), interpolate = interpolate)
}
```

```{r page 1 header hide code, fig.height=1.5, fig.width=10, echo=FALSE, out.width="100%", fig.align='right', results='hold'}
logo <- readPNG("UWA logo_text_V_wsL.png")
par(mar = c(0,0,0,0))
layout(matrix(c(1,1,1,1,2),nrow = 1))

plot(c(0,1),c(0,1), axes=F, type="n",xaxt="n", yaxt="n",ann=F)
text(-0.025,0.9, pos = 4, cex = 2.6, font = 2, 
     labels="ENVTM501 Data Analysis in R for Compositional Data")
text(-0.025,0.6, pos = 4, cex = 2.2, 
     labels="Multivariate analyses on whole-rock major element compositional data")
text(-0.025,0.4, pos = 4, font = 3, cex = 1.4, col = 12,
     labels="Geochemical data for the Yilgarn Craton, Western Australia by J.A. Hallberg")
text(1,0.1, pos = 2, font = 3, family = 'serif', cex = 1.5, col = 2,
     labels="Andrew Rate, School of Agriculture and Environment")
plot(1,1, axes=F, type="n",xaxt="n", yaxt="n",ann=F)
addImg(logo, x = 1.2, y = 1, width = 0.5)
par(mar = c(3.5,3.5,0.5,0.5))
```

```{r load packages etc., message=FALSE, warning=FALSE, include=FALSE, results='hide'}
library(rgr)
library(car)
library(cluster)
library(factoextra)
library(ggplot2)
library(reshape2)
library(ggpubr)
library(flextable)
library(magrittr)
library(RcmdrMisc)

set_flextable_defaults(theme_fun = "theme_zebra", 
                       font.size = 9, fonts_ignore = TRUE)
BorderDk <- officer::fp_border(color = "#B5C3DF", style = "solid", width = 1)
BorderLt <- officer::fp_border(color = "#FFFFFF", style = "solid", width = 1)

UWApal <- c("black", "#003087", "#DAAA00", "#8F92C4", "#E5CF7E", 
          "#001D51", "#B7A99F", "#A51890", "#C5003E", "#FDC596", 
          "#AD5D1E","gray40","gray85","#FFFFFF","transparent"); palette(UWApal)
```

```{r read file and show data, include=FALSE, results='hide'}
Hallberg <- read.csv("Hallberg.csv", stringsAsFactors = TRUE)
```

```{r make new Rock factor with abbreviated names, eval=FALSE, include=FALSE, results='hold'}
# row.names(Hallberg) <- paste0(as.character(Hallberg$Rock),seq(1:NROW(Hallberg)))
# Hallberg$sRock <- as.character(Hallberg$Rock)
# Hallberg$sRock <- gsub("Basaltic Trachyandesite","C",Hallberg$sRock)
# Hallberg$sRock <- gsub("Basaltic Andesite","L",Hallberg$sRock)
# Hallberg$sRock <- gsub("Andesite","A",Hallberg$sRock)
# Hallberg$sRock <- gsub("Trachybasalt","T",Hallberg$sRock)
# Hallberg$sRock <- gsub("Basalt","B",Hallberg$sRock)
# Hallberg$sRock <- gsub("Basanite","N",Hallberg$sRock)
# Hallberg$sRock <- gsub("Phonotephrite","P",Hallberg$sRock)
# Hallberg$sRock <- as.factor(Hallberg$sRock)
```

```{r clr-tranform data, include=FALSE, results='hold'}
Hallberg_clr <- Hallberg
Hallberg_clr[,11:24] <- 
  clr(Hallberg_clr[,11:24], ifwarn = FALSE) # recommend ifwarn = TRUE
```

We are using a curated version of a whole rock major element dataset from 
Hallberg (https://catalogue.data.wa.gov.au/dataset/hallberg-geochemistry)

# LDA -- Linear Discriminant Analysis

We use **linear discriminant analysis (LDA)** to create functions of the dataset
variables which maximise the separation between pre-defined categories. LDA is
also sensitive to compositional closure, as I hope we're starting to expect by
now! Predicting existing categories is not the most useful application of LDA 
-- ideally, we would like to measure some key variables so we can predict a
previously unknown category (a strategy for machine learning). Towards the end
of this section we will see if we can divide our dataset into two groups – one
to 'train' our LDA model to generate a set of linear discriminant functions,
which we can then apply to the remaining observations in our dataset to
*validate* our LDA model (a later session will cover this issue).

Linear discriminant analysis resembles principal components analysis (PCA), in
that it generates new sets of variables (dimensions) to reduce the complexity
(*i.e*. dimensionality) of multivariate data. Another important feature of LDA,
as mentioned above, is its ability to classify the observations in our data --
LDA is a **supervised classification** method, in that it requires us to use
pre-defined categories in our dataset. Key differences between LDA and PCA are
that:

- the components of PCA capture successively less multivariate *variance* (and are are not 'supervised' by pre-defined categories)
- the dimensions of LDA *maximise some measure of separation* between supervised categories. 

In LDA, this is achieved by generating new variables (the dimensions) which are 
linear functions of the existing variables in the dataset.

To implement LDA in R, we use the \texttt{lda()} function in the **MASS** package 
(Venables and Ripley, 2002). We specify the variables to be considered using a
formula similar to that used for multiple regression, and we set the prior 
(initial) probabilities of an observation being in a particular category at the 
actual frequencies at which they occur in the data.

In practice, we apply LDA to a scaled transformations of our variables (*i.e*. 
conversion to Z-scores with mean = 0 and standard deviation = 1). This avoids 
the variables with larger absolute values having an unbalanced effect on the 
results.

## LDA on closed whole rock major element data

We will use LDA to discriminate the rock type, contained in the column
'**Rock**' in the Hallberg dataset. The variables we will use are the major
element oxide contents, SiO<sub>2</sub>, TiO<sub>2</sub>,
Al<sub>2</sub>O<sub>3</sub>, Fe<sub>2</sub>O<sub>2</sub>, FeO, MnO, MgO, CaO,
Na<sub>2</sub>O, K<sub>2</sub>O, & P<sub>2</sub>O<sub>5</sub>. We're excluding
LOI, CO<sub>2</sub> and H<sub>2</sub>O, since these variables contain too many
similar low values which adversely affects the matrix algebra calculation of the
linear discriminant analysis procedure.

\scriptsize

```{r LDA closed whole rock data, results='hold'}
data0 <- Hallberg
data0[,c(11:24)] <- scale(data0[,11:24]) # scale just numeric variables
lda_rock_clos <- lda(formula = Rock ~ SiO2 + TiO2 + Al2O3 + Fe2O3 + FeO + MnO + 
                       MgO + CaO + Na2O + K2O + P2O5, 
                    data = data0,
                    prior = as.numeric(summary(Hallberg$Rock))/
                      nrow(Hallberg)) 
print(lda_rock_clos)
```

\scriptsize

## LDA on open (CLR) whole rock major element data

```{r LDA open whole rock data, message=FALSE, warning=FALSE, results='hold'}
data0 <- Hallberg_clr
data0[,11:24] <- scale(data0[,11:24]) # scale just numeric variables
lda_rock_open <- lda(formula = Rock ~ SiO2 + TiO2 + Al2O3 + Fe2O3 + FeO + MnO + 
                       MgO + CaO + Na2O + K2O + P2O5, 
                    data = data0,
                    prior = as.numeric(summary(data0$Rock))/nrow(data0)) 
print(lda_rock_open)
```

```{r correl matrix CLR not used in this version, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
cormat <- rcorr.adjust(data0[,11:24], 
                       type="pearson")
cat("\nCorrelation matrix for predictors (not part of LDA output):\n"); print(cormat$R$r, digits = 2)
rm(cormat)
```

\normalsize

The LDA model specified and the prior probabilities are reported back to us in
the output. The *Group means* sub-table in the output contains the centroids of
the categories (5 classes, in our whole rock data) in the n-dimensional space
defined by the n variables used for classification. The *Coefficients of linear*
*discriminants* sub-table essentially defines the linear discriminant functions 
that separate our categories, since each function (LD1, LD2, *etc*.) is a 
linear combination of all the variables. Finally, the *Proportion of trace* 
sub-table gives the *proportions of between-class variance* that are explained 
by successive discriminant functions (*e.g*. for the open Hallberg rock data,
LD1 explains 0.716 (72%) and LD2 explains 0.171 (17%) of variance between Rock 
type categories).

[**Note**: sometimes we cannot make an LDA model, because the predictors are
collinear (highly correlated). We may be able to fix this by inspecting a 
correlation matrix for all the predictor variables, and removing *one* variable 
at a time from correlated pairs, then re-running the LDA procedure. By analogy 
with multiple linear regression, this would mean a Pearson's r $\ge$ 0.8.]

## Visualising LDA separation

### LDA histograms

We can plot the separation achieved by each linear discriminant (LD) function 
by predicting the classification using the input data, then using the 
\texttt{ldahist} function (Venables and Ripley 2002). To see the separation in 
another LDA dimension, we change the subscript in the \texttt{predClos$x[,1]} 
option. Histograms (actually drawn with custom code, enabling a plot with a
side-by side comparison) are shown in \autoref{ldahistComp}.

\scriptsize

```{r lda hist compare, fig.align='center', fig.cap="\\label{ldahistComp}Histograms based on the first linear discriminant function for (a) closed and (b) open (CLR-transformed) whole-rock major element data.", fig.height=14, fig.width=10, out.width="80%", results='hold'}
predClos <- predict(lda_rock_clos, Hallberg[,11:21])
predOpen <- predict(lda_rock_open, Hallberg[,11:21])
LD1c <- data.frame(Rock=as.character(Hallberg$Rock),LD1=predClos$x[,1])
LD1c$Rock <- factor(LD1c$Rock, levels=levels(Hallberg$Rock))
par(mfcol = c(nlevels(LD1c$Rock),2), mar = c(1,2,1,1), oma = c(1,0,1,0), 
    mgp = c(0.75,0.2,0), tcl=0.15)
for (i in 1:nlevels(LD1c$Rock)){
  with(subset(LD1c, subset=LD1c$Rock==levels(LD1c$Rock)[i]),
       hist(LD1, main = "", breaks = pretty(LD1c$LD1, n=20), col=5,
       xlim = c(min(LD1c$LD1, na.rm=T),max(LD1c$LD1, na.rm=T))))
  box()
  mtext(levels(LD1c$Rock)[i],3,-1.55,adj=0.505, cex = 0.85, font = 2, col=14)
  mtext(levels(LD1c$Rock)[i],3,-1.5, cex = 0.85, font = 2, col = 11)
  if(i==1) mtext("(a) Closed data", 3, 0.5, font=2, cex = 1.4, col = 11)
}

LD1o <- data.frame(Rock=as.character(Hallberg$Rock),LD1=predOpen$x[,1])
LD1o$Rock <- factor(LD1o$Rock, levels=levels(Hallberg$Rock))
for (i in 1:nlevels(LD1o$Rock)){
  with(subset(LD1o, subset=LD1o$Rock==levels(LD1o$Rock)[i]),
       hist(LD1, main = "", breaks = pretty(LD1o$LD1, n=20), col=4,
            xlim = c(min(LD1o$LD1, na.rm=T),max(LD1o$LD1, na.rm=T))))
  box()
  mtext(levels(LD1o$Rock)[i],3,-1.55, adj=0.505, cex = 0.85, font = 2, col=14)
  mtext(levels(LD1o$Rock)[i],3,-1.5, cex = 0.85, font = 2, col = 2)
  if(i==1) mtext("(b) Open data", 3, 0.5, font=2, cex = 1.4, col = 2)
}
```

\normalsize

The sets of histograms for closed and open data in \autoref{ldahistComp} both
show some separation of categories, but with overlap. Of course this only shows
the separation in one dimension, and two or more dimensions may be needed to
achieve clear separation. We will make plots showing more than one LDA dimension
later.

### Partition plots

Another potentially useful way of showing separation of groups in LDA is to use
a *partition plot*, accessible using the \texttt{partimat()} function from the
**klaR** R package (Weihs et al. 2005).

\scriptsize

```{r pplot comparison, fig.height=5, fig.width=10, fig.align='center', out.width="100%", fig.cap="\\label{pplotComp}Partition plots for (a) closed and (b) open (CLR-transformed) whole rock major element data, based on the SiO2 and TiO2 contents. Filled symbols are means in each category, with red letters showing apparently mis-classified observations.", results='hold'}
require(klaR)
par(mfrow = c(1,2),mar = c(3,3,1,1), mgp= c(1.3,0.2,0), tcl=0.2, font.lab=2)
with(Hallberg, 
     drawparti(Rock, SiO2, TiO2, method="lda",image.colors = c(10,7,14,13,2:5),
               xlab = expression(bold(paste(SiO[2]," (%)"))), 
               ylab = expression(bold(paste(TiO[2]," (%)"))))
     )
mtext("(a)", 3, -1.5, adj=0.05, font=2, cex = 1.2)
with(Hallberg_clr, 
     drawparti(Rock, SiO2, TiO2, method="lda",image.colors = c(10,7,14,13,2:5),
               xlab = expression(bold(paste(SiO[2]," (CLR-transformed)"))), 
               ylab = expression(bold(paste(TiO[2]," (CLR-transformed)"))))
     )
mtext("(b)", 3, -1.5, adj=0.95, font=2, cex = 1.2)
```

\normalsize

Partition plots, such as those  in \autoref{pplotComp}, are presented for single
pairwise combinations of the variables (in this example SiO<sub>2</sub> and
TiO<sub>2</sub>) used to make the LDA model. We can make differnt plots by
specifying different variables in the \texttt{drawparti()} function. Each such plot
can be considered to be a different view of the data (which of course has 
multiple dimensions). Colored regions delineate each classification area. Any
observation that falls within a region is predicted to be from a specific
category, with apparent mis-classification in a different color (but we usually 
need more than two dimensions for correct classification). Each plot also
includes the apparent error rate for that view of the data.

## Scatter Plots resembling biplots

Scatter-plots showing each variable and observation in linear discriminant
dimensions, and grouped by category, are useful for visual assessment of how
well the LDA model separates the observations.

\scriptsize

```{r plot LDA closed, fig.height=8, fig.width=8, fig.align='center', out.width="80%", fig.cap="\\label{LDAclos}Linear discriminant analysis (LDA) plots for closed rock composition data: (a) variable coefficients in LD1-LD2 space, and predictions for observations in (b) LD1-LD2 space; (c) LD1-LD3 space; (d) LS2-LD4 space. Legend in (a) applies to plots in (b), (c), and (d).", results='hold'}
par(mfrow = c(2,2), mar = c(3.5,3.5,1,1), mgp = c(1.5,0.3,0), tcl = 0.25,
    lend = "square", ljoin = "mitre", cex.main = 0.9, font.lab=2)
plot(lda_rock_clos$scaling[,1], lda_rock_clos$scaling[,2],
     xlim = c(-4,1), ylim=c(-0.8,1.4), 
     xlab="Linear Discriminant [1]", ylab="Linear Discriminant [2]", 
     main="(a) Variable Coefficients [LD1, LD2]")
abline(v=0,col="grey",lty=2)
abline(h=0,col="grey",lty=2)
text(lda_rock_clos$scaling[,1],lda_rock_clos$scaling[,2],
     labels=names(Hallberg)[11:21], pos = c(3,1,1,4),
     cex = 0.95, col = 2, offset = 0.2)
mtext("(a)", 3, -1.5, adj = 0.45, cex = 1.2, font = 2)

ldaPred_rock_clos <- predict(lda_rock_clos)

for(i in 1:NROW(lda_rock_clos$scaling)){
  arrows(0,0,lda_rock_clos$scaling[i,1],lda_rock_clos$scaling[i,2],
         length = 0.1, col = 7)
}
legend("bottomleft",legend=levels(Hallberg$Rock), ncol = 1, bty="n", 
       inset=0.01, col=c(1:3,6,8,9,11,12), pch=c(0:7), pt.lwd = 2,
    title="Rock Type in (b) - (d)", pt.cex = 1.5, cex = 1.1, y.intersp = 1)

plot(ldaPred_rock_clos$x[,1], ldaPred_rock_clos$x[,2],
     col=c(1:3,6,8,9,11,12)[Hallberg_clr$Rock],
     pch=c(0:7)[Hallberg_clr$Rock], lwd = 2, cex = 1.5, 
  xlab="Linear Discriminant [1]", ylab="Linear Discriminant [2]", 
  main="Predictions for Observations [LD1, LD2]")
abline(v=0,col="grey",lty=2)
abline(h=0,col="grey",lty=2)
# text(ldaPred_rock_clos$x[,1], ldaPred_rock_clos$x[,2], 
#      labels=as.character(Hallberg$Rock), col=c(2,4,6,3,12)[Hallberg$Rock],
#      pos=1, offset=0.15, cex=0.65)
# legend("bottomright",legend=levels(Hallberg$Rock)[6:13],
#        ncol = 2, col=c(6:13), pch=c(5:12), pt.lwd = 2,
#        title=expression(bold("Rock Type")),
#        bty="n", inset=0.01, 
#        pt.cex=c(1.8,1.8,2,2,1.3), cex=0.9)
mtext("(b)", 3, -1.5, adj = 0.95, cex = 1.2, font = 2)

plot(ldaPred_rock_clos$x[,1], ldaPred_rock_clos$x[,3], 
     col=c(1:3,6,8,9,11,12)[Hallberg_clr$Rock],
     pch=c(0:7)[Hallberg_clr$Rock], lwd = 2, cex = 1.5, 
  xlab="Linear Discriminant [1]", ylab="Linear Discriminant [3]", 
  main="Predictions for Observations [LD1, LD3]")
abline(v=0,col="grey",lty=2)
abline(h=0,col="grey",lty=2)
# text(ldaPred_rock_clos$x[,1], ldaPred_rock_clos$x[,3], 
#      labels=Hallberg$Rock, col=c(1:13)[Hallberg$Rock],
#      pos=1, offset=0.15, cex=0.65)
mtext("(c)", 3, -1.5, adj = 0.05, cex = 1.2, font = 2)

plot(ldaPred_rock_clos$x[,2], ldaPred_rock_clos$x[,4],
     col=c(1:3,6,8,9,11,12)[Hallberg_clr$Rock],
     pch=c(0:7)[Hallberg_clr$Rock], lwd = 2, cex = 1.5, 
  xlab="Linear Discriminant [2]", ylab="Linear Discriminant [4]", 
  main="Predictions for Observations [LD2, LD4]")
abline(v=0,col="grey",lty=2)
abline(h=0,col="grey",lty=2)
# text(ldaPred_rock_clos$x[,2], ldaPred_rock_clos$x[,4], 
#      labels=Hallberg$Rock, col=c(2,4,6,3,12)[Hallberg$Rock],
#      pos=1, offset=0.15, cex=0.65)
mtext("(d)", 3, -1.5, adj = 0.55, cex = 1.2, font = 2)
```

```{r plot LDA open, fig.height=8, fig.width=8, out.width="70%", fig.align='center', fig.cap="\\label{LDAopen}Linear discriminant analysis (LDA) plots for open (CLR-transformed) rock composition data: (a) variable coefficients in LD1-LD2 space, and predictions for observations in (b) LD1-LD2 space; (c) LD1-LD3 space; (d) LS2-LD3 space. Legend in (a) applies to plots (b), (c), and (d).", results='hold'}
par(mfrow = c(2,2), mar = c(3.5,3.5,1,1), mgp = c(1.5,0.3,0), tcl = 0.25,
    lend = "square", ljoin = "mitre", cex.main = 0.9, font.lab=2)
plot(lda_rock_open$scaling[,1], lda_rock_open$scaling[,2], 
     xlim = c(-1,2.5),  ylim = c(-1.3,1.3), 
     xlab="Linear Discriminant [1]", ylab="Linear Discriminant [2]", 
     main="Variable Coefficients [LD1, LD2]")
abline(v=0,col="grey",lty=2); abline(h=0,col="grey",lty=2)
text(lda_rock_open$scaling[,1], lda_rock_open$scaling[,2], 
     labels=names(Hallberg_clr)[11:21],
     pos = 1, cex = 0.9, col = 2, offset=0.2)
for(i in 1:NROW(lda_rock_open$scaling)){
  arrows(0,0,lda_rock_open$scaling[i,1],lda_rock_open$scaling[i,2],
         length = 0.1, col = 7) }
mtext("(a)", 3, -1.5, adj = 0.05, cex = 1.2, font = 2)
legend("bottomright", ncol = 1, legend=levels(Hallberg$Rock), 
       col=c(1:3,6,8,9,11,12), pch=c(0:7), pt.lwd = 2,
       bty="n", box.col="grey90", y.intersp = 1, 
       title="Rock Type in (b) - (d)",
       box.lwd=2, inset=0.02, pt.cex=1.5, cex=1.1)

ldaPred_rock_open <- predict(lda_rock_open)

plot(ldaPred_rock_open$x[,1], ldaPred_rock_open$x[,2], 
     col=c(1:3,6,8,9,11,12)[Hallberg$Rock],
     pch=c(0:7)[Hallberg$Rock], lwd=2, 
     cex = 1.5, 
     main="Predictions for Observations [LD1, LD2]", 
     xlab="Linear Discriminant [1]", ylab="Linear Discriminant [2]")
abline(v=0,col="grey",lty=2); abline(h=0,col="grey",lty=2)
# text(ldaPred_rock_open$x[,1], ldaPred_rock_open$x[,2], labels=Hallberg$Rock, 
#      col=c(2,4,6,3,12)[Hallberg$Rock], pos=1, offset=0.15, cex=0.65)
mtext("(b)", 3, -1.5, adj = 0.05, cex = 1.2, font = 2)

plot(ldaPred_rock_open$x[,1], ldaPred_rock_open$x[,3], 
     col=c(1:3,6,8,9,11,12)[Hallberg$Rock],
     pch=c(0:7)[Hallberg$Rock], lwd=2, 
     cex = 1.5, 
     main="Predictions for Observations [LD1, LD3]", 
     xlab="Linear Discriminant [1]", ylab="Linear Discriminant [3]")
abline(v=0,col="grey",lty=2); abline(h=0,col="grey",lty=2)
# text(ldaPred_rock_open$x[,1], ldaPred_rock_open$x[,3], labels=Hallberg$Rock, 
#      col=c(2,4,6,3,12)[Hallberg$Rock], pos=1, offset=0.15, cex=0.65)
mtext("(c)", 3, -1.5, adj = 0.05, cex = 1.2, font = 2)

plot(ldaPred_rock_open$x[,2], ldaPred_rock_open$x[,4], 
     col=c(1:3,6,8,9,11,12)[Hallberg$Rock],
     pch=c(0:7)[Hallberg$Rock], lwd=2, 
     cex = 1.5, 
     main="Predictions for Observations [LD2, LD4]", 
     xlab="Linear Discriminant [2]", ylab="Linear Discriminant [4]")
abline(v=0,col="grey",lty=2); abline(h=0,col="grey",lty=2)
# text(ldaPred_rock_open$x[,2], ldaPred_rock_open$x[,3], labels=Hallberg$Rock, 
#      col=c(2,4,6,3,12)[Hallberg$Rock], pos=1, offset=0.15, cex=0.65)
mtext("(d)", 3, -1.5, adj = 0.05, cex = 1.2, font = 2)
```

```{r reset par to 1x1, message=FALSE, warning=FALSE, include=FALSE, results='hide'}
par(mfrow=c(1,1))
```

\normalsize

From the plots in \autoref{LDAclos} and \autoref{LDAopen}, we can see that the
LDA models obtained are cetainly able to separate observations by the selected
factor. Firstly, however, there is a lot of clustering of the predictor
variables in \autoref{LDAclos}(a), which may relate to spurious relationships
between variables caused by compositional closure. This clustering is not so
pronounced in \autoref{LDAopen}(a), since the closure has been removed by
CLR-transformation.

Out of the combinations of LDA dimensions selected, the best separation is with
LD2 *vs*. LD1, which is not surprising since these dimensions together account
for about 95% of the between-groups variance -- for both closed and open data.
There is a lot of apparent overlap, but we can not see the true separation in
only 2 dimensions. The LDA performed on open data may result in slightly clearer
separation of samples by Rock category than LDA using closed data, but without a
multidimensional view this is also hard to be sure about.

One way that we can get a better idea about the usefulness of our classification
models is to perform some validation. This involves 'training' the model on a
subset of our data, and trying to predict the category of a different subset
using the training model. A later session will look at some ways we can do
this.
