---
output: 
  html_document: 
    toc_depth: 2
    fig_caption: yes
    number_sections: yes
    self_contained: no
---

```{r initial settings, include=FALSE}
library(flextable)
library(magrittr)
library(png)

set_flextable_defaults(theme_fun = "theme_zebra", 
                       font.size = 9, fonts_ignore = TRUE)
BorderDk <- officer::fp_border(color = "#B5C3DF", style = "solid", width = 1)
BorderLt <- officer::fp_border(color = "#FFFFFF", style = "solid", width = 1)

addImg <- function(obj, x = NULL, y = NULL, width = NULL, interpolate = TRUE){
  if(is.null(x) | is.null(y) | is.null(width)){stop("Must provide args 'x', 'y', and 'width'")}
  USR <- par()$usr ; PIN <- par()$pin ; DIM <- dim(obj) ; ARp <- DIM[1]/DIM[2]
  WIDi <- width/(USR[2]-USR[1])*PIN[1] ;   HEIi <- WIDi * ARp 
  HEIu <- HEIi/PIN[2]*(USR[4]-USR[3]) 
  rasterImage(image = obj, xleft = x-(width/2), xright = x+(width/2),
            ybottom = y-(HEIu/2), ytop = y+(HEIu/2), interpolate = interpolate)
}
```

In this UWA Micro-Credential course, you will learn specialised methods which
allow you to analyse compositional data so that results can be interpreted with
confidence. Students will also gain skills in use of R for relevant multivariate
data analyses.

First we load the packages we need, and set a preferred colour palette.

```{r load packages etc., message=FALSE, warning=FALSE, results='hide'}
library(corrplot)
library(RcmdrMisc)
library(rgr)
library(car)
library(cluster)

palette(c("black", "#003087", "#DAAA00", "#8F92C4", "#E5CF7E", 
          "#001D51", "#B7A99F", "#A51890", "#C5003E", "#FDC596", 
          "#AD5D1E", "gray40", "gray85", "#FFFFFF", "transparent"))
```

## Load the cities data

**Data are from**: Fig. 11 *in* Hu, J., Wang, Y., Taubenböck, H., Zhu, X.X., 2021. Land consumption in cities: A comparative study across the globe. *Cities*, **113**: 103163, <https://doi.org/10.1016/j.cities.2021.103163>. (digitised from figure using <https://automeris.io/WebPlotDigitizer/>)

After reading the data we make a column containing abbreviated codes for the
land-use type factor, using the **gsub()** function to substitute character
strings.

```{r read file and show data, results='hold'}
cities <- 
  read.csv(paste0("https://raw.githubusercontent.com/Ratey-AtUWA/",
                  "compositional_data/main/cities_Hu_etal_2021.csv"), 
           stringsAsFactors = TRUE)
ft <- flextable(cities)
ft <- add_header_row(ft, values = c("","Land use categories","City categories"),
                     colwidths = c(1,4,3))
ft <- set_caption(ft, caption = "Cities land use data used in this document, showing cities 
      categorized by land-use Type, Global socioeconomic zone, and geographic Region.")
ft <- theme_zebra(ft,odd_header = "#D0E0FF", even_header = "#D0E0FF")
ft <- align(ft,align = "center", part = "header")
ft <- height_all(ft, height=0.39, unit = "cm"); ft <- hrule(ft, rule = "exact")
ft <- width(ft, j=1:8, width=c(2.7,1.5,1.5,1.8,1.5,3,2,3), unit = "cm")
ft <- border_outer(ft, border = BorderDk, part = "all")
ft %>% border_inner_v(border=BorderLt, part="header")
```

The original data categorised the cities by **Type**, based on the land use
categories occupying the greatest proportion(s) of urban area. We have added two
other categories, **Global** identifying each city as being in the socioeconomic
"global south" or "global north" zones, and also assigning each city to a
geographic **Region**, a category having eight levels. All categories are stored
as factors in the R data frame.

```{r make new Type factor with abbreviated names, results='hide', echo=FALSE}
row.names(cities) <- as.character(cities$City)
cities$sType <- as.character(cities$Type)
cities$sType <- gsub("Compact-Open","CO",cities$sType)
cities$sType <- gsub("Open-Lightweight","OL",cities$sType)
cities$sType <- gsub("Compact","C",cities$sType)
cities$sType <- gsub("Open","O",cities$sType)
cities$sType <- gsub("Industrial","I",cities$sType)
cities$sType <- as.factor(cities$sType)
```

## Make CLR-transformed dataset

```{r clr-transform data, paged.print=FALSE, results='hold'}
cities_clr <- cities
cities_clr[,c("Compact","Open","Lightweight","Industry")] <- 
  clr(cities_clr[,c("Compact","Open","Lightweight","Industry")])
head(cities_clr[,1:8])
```

## Compare correlation matrices

```{r correlation matrix comnparison, fig.height=3.6, fig.width=7.2, out.width="70%", fig.align='center', fig.cap="\\label{corrmats}Correlation matrices for (a) closed, and (b) open (CLR-transformed) variables (code is not shown).", results='hold',echo=FALSE}
par(mar=c(1,5,5,3), mfrow=c(1,2), mgp=c(1.5,0.2,0), xpd=T)
cormat <- round(cor(cities[,c("Compact","Open",
                              "Lightweight","Industry")]),2)
corrplot(cormat, method="ellipse", cl.ratio = 0.25,
         addCoef.col = "black", tl.col = "grey33", cl.pos = "b")
mtext("(a)",3,1,adj=0.1)
cormat <- round(cor(cities_clr[,c("Compact","Open",
                                  "Lightweight","Industry")]),2)
corrplot(cormat, method="ellipse", cl.ratio = 0.25,
         addCoef.col = "black", tl.col = "navy", cl.pos = "b")
mtext("(b)",3,1,adj=0.1)
par(mfrow=c(1,1), mar = c(4,4,1,1))
```

By comparing Fig.1(a) and Fig.1(b), we can already see that addressing closure
with the centered-logratio (CLR) transformation changes the interpretation of
relationships between the variables.

## Some other things to try

- comparing the distributions of untransformed and CLR-transformed pairs of variables using histograms or density plots
- comparing scatter plots of untransformed and CLR-transformed pairs of variables (maybe using the **scatterplotMatrix()** function in the **car** package).

# Principal components analysis

It is quite common to measure many variables in environmental science and other
disciplines. Using various types of *ordination analysis*, we can use the
information contained in multiple variables to create a reduced subset of
variables containing nearly the same amount of information. Ordination methods
are also referred to, for this reason, as 'data reduction' methods and are
commonly used for multivariate analysis.

One of the earliest and most widely used ordination methods for exploration and
dimension-reduction of multivariate data is **principal components analysis**
(PCA). Imagine a dataset with many samples (rows) and n continuous numeric
variables (columns) which contain quantitative information about each sample
such as concentrations, heights, velocities, *etc*. For these *n*
variables/dimensions, the principal component calculation generates *n* new
variables, or principal components, which are each a function of the set of all
the original variables (so each principal component is defined by a weighting or
coefficient for each of the original variables). We may choose to omit some
variables from the analysis if they contain too many missing observations or if
there is another valid reason to doubt their integrity. Since each principal
component is selected to account for successively smaller proportions of the
multiple variance, it is usually the first few principal components which
explain most of the variance and therefore contain the most useful information.
We conventionally visualize this in a 'scree plot' (Fig.2), a kind of bar graph
showing the decrease in variance accounted for by each component.

```{r run and summarise PCA, results='hold'}
#
data0 <- na.omit(cities[,c("Compact","Open","Lightweight","Industry")])
pca_cities_clos <- prcomp(data0, scale. = TRUE)
cat("Variable weightings (rotations) - closed data\n")
pca_cities_clos$rot
cat("...\n\nComponent Variances - Closed data\n")
pca_cities_clos$sdev^2
cat("...\n\nProportions of variance explained by each component",
    "\nCLR-transformed (open) data\n")
round(pca_cities_clos$sdev^2/sum(pca_cities_clos$sdev^2),3)
cat("\n--------------------\nVariable weightings (rotations) - open data\n")

data0 <- na.omit(cities_clr[,c("Compact","Open","Lightweight","Industry")])
pca_cities_open <- prcomp(data0, scale. = TRUE)
pca_cities_open$rot
cat("...\n\nComponent Variances - CLR-transformed (open) data\n")
pca_cities_open$sdev^2
cat("...\n\nProportions of variance explained by each component",
    "\nCLR-transformed (open) data\n")
round(pca_cities_open$sdev^2/sum(pca_cities_open$sdev^2),3)
rm(data0)
```

## Principal Components Analysis output

As well as the component variances, the useful results of principal components
analysis include the variable weightings or 'rotations' for each principal
component. Referring to just the output for **open data** above, PC1 (which
explains the most multivariate variance) has its greatest absolute contributions
from the Lightweight and Industry variables, which act in opposite directions
(remembering that the sign is arbitrary, but that *relative* sign and magnitude
are important). PC2 contains mainly variance from the Compact and Open
variables, PC3 (which explains much less variance) reflects Open/Industry, and
we can ignore PC4 as it has a negligible component variance.

In addition, every individual observation (sample) is a multivariate point, the
observation scores for all samples in each principal component based on the
values of the variables used in PCA for that sample. It is conventional to plot
both of these two types of output in a principal component **biplot**. Before
discussing the biplot, we should note that the sign (positive or negative) of
variable weightings and observation scores (*i.e*. the direction of the axes) is
arbitrary and should not affect our interpretations.

```{r scree plots, echo=FALSE, fig.height=2.5, fig.width=6, message=FALSE, warning=FALSE, out.width='60%', fig.align='center', fig.cap="\\label{screeplots}PCA scree plots for urban land-use data, for: (a) compositionally closed proportions; (b) data corrected for closure using CLR-transformation.", results='hold'}
par(mar = c(3.,3.5,1.2,1), mfrow = c(1,2), mgp=c(2,0.8,0), font.lab=2)
plot(pca_cities_clos, main = "(a) closed", cex.main = 1, col = 5)
axis(1, at=seq(0.8,4.2,l=4), labels=seq(1,4,1), tcl=0.01, col=15, mgp=c(0.5,0.1,0))
mtext("Component", side = 1, line = 1.15, font = 2)
plot(pca_cities_open, main = "(b) open (CLR)", cex.main = 1, col = 4)
axis(1, at=seq(0.8,4.2,l=4), labels=seq(1,4,1), tcl=0.01, col=15, mgp=c(0.5,0.1,0))
mtext("Component", side = 1, line = 1.15, font = 2)
par(mar = c(3.5,3.5,1,1), mfrow = c(1,2), mgp=c(1.7,0.3,0))
```

There are informal 'rules' for deciding which of the principal components 
contain useful information:

1. Components having variance greater than 1 (the 'Kaiser' criterion)
2. Components up to a cumulative proportion of variance explained of 0.8 (80%)

For example, for the open data, the component variances are PC1 1.976, PC2
1.511, PC3 0.513, PC4 1.5E-31 -- so only PC1 and PC2 meet the first criterion.
Similarly, the *cumulative* proportions of variance explained by each component
are PC1 0.494, PC1+PC2 0.872, so by criterion 2 no useful information is
contained in principal components >2. In this example both criteria agree (they
don't always!).

Another part of the PCA output that we need to know about is the set of
*observation scores*, essentially showing where each observation (each city, in
these data) plots in principal component space. We're only interested in PC1 and
PC2, since they are the only components meeting the criteria. The observation
scores for the first rows of the data are in the Table below:

```{r table of PCA obs scores, tab.cap="\\label{PCAObsTabl}PCA observation scores for the first 8 rows of the cities data (closed and open)", results='hold'}
OScores <- data.frame(City=cities_clr$City[1:8],
                      PC1_closed=round(pca_cities_clos$x[1:8,1],3),
                      PC2_closed=round(pca_cities_clos$x[1:8,2],3),
                      PC1_open=round(pca_cities_open$x[1:8,1],3),
                      PC2_open=round(pca_cities_open$x[1:8,2],3))
ft <- flextable(OScores)
ft <- theme_zebra(ft,odd_header = "#D0E0FF", even_header = "#D0E0FF")
ft <- border_outer(ft, border = BorderDk, part = "all")
ft %>% border_inner_v(border=BorderLt, part="header")
```

## References and R Packages

Fox, J. (2022). *RcmdrMisc: R Commander Miscellaneous Functions*. R package
version 2.7-2. <https://CRAN.R-project.org/package=RcmdrMisc>

John Fox and Sanford Weisberg (2019). *An {R} Companion to Applied Regression*
(**car**), Third Edition. Thousand Oaks CA: Sage. URL:
<https://socialsciences.mcmaster.ca/jfox/Books/Companion/>

Garrett, R.G. (2018). *rgr: Applied Geochemistry EDA*. R package version 1.1.15.
<https://CRAN.R-project.org/package=rgr>

Greenacre, M. (2018). *Compositional Data Analysis in Practice*. CRC Press LLC, 
Boca Raton, FL, USA. 

Hu, J., Wang, Y., Taubenböck, H., Zhu, X.X. (2021). Land consumption in cities:
A comparative study across the globe. *Cities*, **113**: 103163,
<https://doi.org/10.1016/j.cities.2021.103163>.

Reimann, C., Filzmoser, P., Garrett, R. G., & Dutter, R. (2008). *Statistical
Data Analysis Explained: Applied Environmental Statistics with R* (First ed.).
John Wiley & Sons, Chichester, UK.