---
documentclass: article
classoption: a4paper
geometry: margin=0.6in
output: 
  pdf_document: 
    fig_caption: TRUE
    number_sections: TRUE
    toc: no
    highlight: tango
  html_document: 
    fig_caption: true
    number_sections: yes
    self_contained: no
  word_document: 
    toc_depth: 2
    fig_caption: yes
fontsize: 12pt
header-includes:
  \usepackage{sourcesanspro}
  \usepackage[T1]{fontenc}
  \renewcommand{\familydefault}{\sfdefault}
  \renewcommand{\thefigure}{2.\arabic{figure}}
  \renewcommand{\thetable}{2.\arabic{table}}
---

\setcounter{section}{1}

```{r load knitr, include=FALSE}
library(flextable)
library(magrittr)
library(png)

set_flextable_defaults(theme_fun = "theme_zebra", 
                       font.size = 9, fonts_ignore = TRUE)
BorderDk <- officer::fp_border(color = "#B5C3DF", style = "solid", width = 1)
BorderLt <- officer::fp_border(color = "#FFFFFF", style = "solid", width = 1)

palette(c("black", "#003087", "#DAAA00", "#8F92C4", "#E5CF7E", 
          "#001D51", "#B7A99F", "#A51890", "#C5003E", "#FDC596", 
          "#AD5D1E", "gray40", "gray85", "#FFFFFF", "transparent"))

addImg <- function(obj, x = NULL, y = NULL, width = NULL, interpolate = TRUE){
  if(is.null(x) | is.null(y) | is.null(width)){stop("Must provide args 'x', 'y', and 'width'")}
  USR <- par()$usr ; PIN <- par()$pin ; DIM <- dim(obj) ; ARp <- DIM[1]/DIM[2]
  WIDi <- width/(USR[2]-USR[1])*PIN[1] ;   HEIi <- WIDi * ARp 
  HEIu <- HEIi/PIN[2]*(USR[4]-USR[3]) 
  rasterImage(image = obj, xleft = x-(width/2), xright = x+(width/2),
            ybottom = y-(HEIu/2), ytop = y+(HEIu/2), interpolate = interpolate)
}
```

```{r page 1 header hide code, fig.height=1.5, fig.width=10, echo=FALSE, out.width="100%", fig.align='right', results='hold'}
logo <- readPNG("UWA logo_text_V_wsL.png")
par(mar = c(0,0,0,0))
layout(matrix(c(1,1,1,1,2),nrow = 1))

plot(c(0,1),c(0,1), axes=F, type="n",xaxt="n", yaxt="n",ann=F)
text(-0.025,0.9, pos = 4, cex = 2.6, font = 2, 
     labels="ENVTM501 Data Analysis in R for Compositional Data")
text(-0.025,0.6, pos = 4, cex = 2.2, 
     labels="Multivariate analyses on cities land-use compositional data")
text(-0.025,0.4, pos = 4, font = 3, cex = 1.4, col = 12,
     labels="Land use proportional area data for 40 cities on Earth")
text(1,0.1, pos = 2, font = 3, family = 'serif', cex = 1.5, col = 2,
     labels="Andrew Rate, School of Agriculture and Environment")
plot(1,1, axes=F, type="n",xaxt="n", yaxt="n",ann=F)
addImg(logo, x = 1.2, y = 1, width = 0.5)
par(mar = c(3.5,3.5,0.5,0.5))
```

```{r load packages etc., message=FALSE, warning=FALSE, include=FALSE, results='hide'}
library(rgr)
library(car)
library(cluster)
library(factoextra)
library(ggplot2)
library(reshape2)
library(ggpubr)
library(flextable)
library(magrittr)

set_flextable_defaults(theme_fun = "theme_zebra", font.size = 9, fonts_ignore = TRUE)
BorderDk <- officer::fp_border(color = "#B5C3DF", style = "solid", width = 1)
BorderLt <- officer::fp_border(color = "#FFFFFF", style = "solid", width = 1)

palette(c("black","blue","green4","red2","purple","darkcyan",
          "firebrick","grey","grey40","white","transparent"))
```

```{r read file and show data, include=FALSE, results='hide'}
cities <- read.csv("cities_Hu_etal_2021.csv", stringsAsFactors = TRUE)
cities$City <- as.character(cities$City)
```

```{r make new Type factor with abbreviated names, include=FALSE, results='hide'}
row.names(cities) <- as.character(cities$City)
cities$sType <- as.character(cities$Type)
cities$sType <- gsub("Compact-Open","CO",cities$sType)
cities$sType <- gsub("Open-Lightweight","OL",cities$sType)
cities$sType <- gsub("Compact","C",cities$sType)
cities$sType <- gsub("Open","O",cities$sType)
cities$sType <- gsub("Industrial","I",cities$sType)
cities$sType <- as.factor(cities$sType)
```

```{r clr-tranform data, include=FALSE, results='hide'}
cities_clr <- cities
cities_clr[,c("Compact","Open","Lightweight","Industry")] <- 
  clr(cities_clr[,c("Compact","Open","Lightweight","Industry")])
names(cities);cat("\n");names(cities_clr)
```

We are using the same cities land-use dataset from the previous session, from Hu
*et al*. (2021).

# K-means clustering

K-means clustering is an unsupervised classification method, which is a type of
machine learning used when you don't know (or don't want to make assumptions
about) any categories or groups. We do have groupings in our data, which are the
factors **Type**, **Global**, and **Region**.

The goal of the K-means clustering algorithm is to find a specified number (*K*)
groups based on the data. The algorithm first requires an estimate of the number
of clusters, and there are several ways to do this. The code below from the
factoextra R package tests different values of *K* and computes the 'total
within sum of squares' (WSS) based on the distance of observations from the
'centroid' (mean) of each cluster, which itself is found by an iterative
procedure. When the decrease in WSS from *K* to *K*+1 is minimal, the algorithm
selects that value of *K*.

\scriptsize

```{r assess optimum clusters for closed and open data, fig.height=3, fig.width=7.2, out.width='75%', fig.align='center', fig.cap="\\label{nClustComp}Estmation of the optimum number of clusters for (a) compositionally closed and (b) open (CLR-transformed) urban land-use data.", results='hold'}
#
require(factoextra)
data0 <- na.omit(cities[,c("Compact","Open","Lightweight","Industry")])
nclus_clos <- fviz_nbclust(data0, kmeans, method = "wss") +
  geom_vline(xintercept = 3, linetype = 2) +
  labs(title="")
#   geom_vline(xintercept = 4, linetype = 2) +
data0 <- na.omit(cities_clr[,c("Compact","Open","Lightweight","Industry")])
nclus_clr <- fviz_nbclust(data0, kmeans, method = "wss") +
  geom_vline(xintercept = 5, linetype = 2) +
  labs(title="")
ggarrange(nclus_clos,nclus_clr,ncol = 2,nrow = 1, 
          labels = c("(a) closed","(b) open (clr)"))
```

\normalsize

We have indicated, in \autoref{nClustComp}, different numbers of clusters for
closed and open data, based on visual identification of a break in the slope of
the WSS *vs*. 'Number of clusters' curve. In the following analyses we will 
assume the same number of clusters (*K* = 4) for both open and closed data.

## Compute K-means clustering for closed data

\scriptsize

```{r kmeans closed, results='hold'}
data0 <- na.omit(cities[,c("sType","Compact","Open","Lightweight","Industry")])
data0[,c("Compact","Open","Lightweight","Industry")] <- 
  scale(data0[,c("Compact","Open","Lightweight","Industry")])
set.seed(123)
cities_clos_kmeans <- kmeans(data0[,2:NCOL(data0)], 4, nstart = 25)
cat("components of output object are:\n")
ls(cities_clos_kmeans)
cat("\nK-means clustering with",length(cities_clos_kmeans$size),
    "clusters of sizes",cities_clos_kmeans$size,"\n\n")
cat("Cluster centers (scaled to z-scores) in K-dimensional space:\n")
cities_clos_kmeans$centers
```
\normalsize

```{r echo=FALSE, tab.cap="Cities in each K-means cluster from analysis of compsitionally closed data.", results='hold'}
outtable <- data.frame(Cluster = seq(1,length(cities_clos_kmeans$size),1),
                       Cities = rep("nil",length(cities_clos_kmeans$size)))
for (i in 1:length(cities_clos_kmeans$size)){
  outtable[i,1] <- paste("Cluster",i)
  outtable[i,2] <- paste(names(which(cities_clos_kmeans$cluster==i)), 
                         collapse = " ")}
ft <- flextable(outtable)
ft <- theme_zebra(ft,odd_header = "#D0E0FF")
ft <- width(ft, j=1:2, width=c(2,5), unit = "in")
ft <- border_inner_h(ft, border=BorderDk, part = "all")
ft <- border_outer(ft, border=BorderDk, part = "all")
ft %>% border_inner_v(border=BorderLt, part="all")
```

The output object from the kmeans function is a list which contains the
information we're interested in: the sum-of-squares between and within clusters
(betweenss, tot.withinss, totss, withinss), the location in *K* dimensions of
the centers of the clusters (centers), the assignment of each observation to a
cluster (cluster), the number of observations in each cluster (size), and the
number of iterations taken to find the solution (iter).

Applying K-means clustering with 4 clusters to the **closed** cities land-use
data results in one larger cluster of 25 cities (3) with three smaller clusters
containing 4-6 cities (1,2, and 4). From the table of cluster centers, we get
some idea that:

- Cluster 1 cities have greater proportions of Lightweight land use
- Cluster 2 cities have greater proportions of Compact land use
- Cluster 3 cities have similar proportions of all land uses
- Cluster 4 cities have greater proportions of Open land use

Interestingly, Lightweight land use does not discriminate between Clusters 2-4,
since the values of cluster centers 2-4 are the same in the Lightweight
dimension.

\scriptsize

## Compute K-means clustering for open data

```{r kmeans open, results='hold'}
data0 <- na.omit(cities_clr[,c("sType","Compact","Open","Lightweight","Industry")])
data0[,c("Compact","Open","Lightweight","Industry")] <- 
  scale(data0[,c("Compact","Open","Lightweight","Industry")])
set.seed(123)
cities_open_kmeans <- kmeans(data0[,2:NCOL(data0)], 4, nstart = 25)
cat("components of output object are:\n")
ls(cities_open_kmeans)
cat("\nK-means clustering with",length(cities_open_kmeans$size),
    "clusters of sizes",cities_open_kmeans$size,"\n\n")
cat("Cluster centers (scaled to z-scores) in K-dimensional space:\n")
cities_open_kmeans$centers
outtable <- data.frame(Cluster = seq(1,length(cities_open_kmeans$size),1),
                       Cities = rep("nil",length(cities_open_kmeans$size)))
```
\normalsize

```{r echo=FALSE, tab.cap="Cities in each K-means cluster from analysis of data with compositional closure removed by centered log ratio transformation.", results='hold'}
for (i in 1:length(cities_open_kmeans$size)){
  outtable[i,1] <- paste("Cluster",i)
  outtable[i,2] <- paste(names(which(cities_open_kmeans$cluster==i)), 
                         collapse = " ")}
ft <- flextable(outtable)
ft <- theme_zebra(ft,odd_header = "#D0E0FF")
ft <- width(ft, j=1:2, width=c(2,5), unit = "in")
ft <- border_outer(ft, border=BorderDk, part = "all")
ft %>% border_inner_v(border=BorderLt, part="all")
```

Applying K-means clustering with 4 clusters to the **open** cities land-use data
results in one larger cluster of 25 cities (Cluster 4 - not the same cities as
the cluster of 25 from analyusis of closed data!). There are three smaller
clusters containing 1-10 cities (1-3). From the table of cluster centers, we get
might conclude that:

- The Cluster 1 city has greater Industry, and lower Compact, land use
- 10 Cluster 2 cities have greater Compact, and lower Open, land uses
- 4 Cluster 3 cities have greater Lightweight, and lower Industry land uses
- 25 Cluster 4 cities have somewhat greater Open land use

An interesting question to ask is whether the clusters are similar to the
categories we already have in the dataset (Type, Global, and Region). The
following plot includes labelling to help us see the relationship of K-means
clusters to the Type category.

\normalsize

To represent clustering in 3 or more dimensions in a plot, we can use principal
components to reduce the number of dimensions, but still retain information from
all the variables. This is done very nicely by the **fviz_cluster()** function
from the R package **factoextra** (Kassambara and Mundt, 2020). The output from
fviz_cluster is a ggplot, so for more efficient presentation and comparison, we
save the ggplot2 output to objects, and plot these together using the
ggarrange() function from the ggpubr R package.

## Plot kmeans clusters for closed and open data

\scriptsize

```{r visualize kmeans clusters compared, fig.align='center', fig.cap="\\label{kmeansComp}K-means cluster plots comparing (a) closed and (b) CLR-transformed (opened) urban land-use data.", fig.height=5, fig.width=10, message=FALSE, warning=FALSE, out.width='100%', results='hold'}
row.names(data0) <- paste0(data0$sType,seq(1,NROW(data0)))
kmeans_viz_clos <- fviz_cluster(cities_clos_kmeans, data = data0[,2:NCOL(data0)],
             palette = c("#800000", "#E7B800", "#FC4E07","purple2"),
             labelsize=10, main = "",
             ellipse.type = "euclid", # Concentration ellipse
             star.plot = TRUE, # Add segments from centroids to items
             repel = F, # if true avoids label overplotting (slow)
             ggtheme = theme_minimal())
kmeans_viz_open <- fviz_cluster(cities_open_kmeans, data = data0[,2:NCOL(data0)],
             palette = c("#800000", "#E7B800", "#FC4E07","purple2"),
             labelsize=10, main = "",
             ellipse.type = "euclid", # Concentration ellipse
             star.plot = TRUE, # Add segments from centroids to items
             repel = F, # if true avoids label overplotting (slow)
             ggtheme = theme_minimal())
ggarrange(kmeans_viz_clos,kmeans_viz_open,ncol = 2,
          labels = c("(a) closed","(b) open (clr)"))
```

\normalsize

From the plots in \autoref{kmeansComp} we can see that, for both the closed and
open cities land use data, K-means clustering did not produce clusters that
overlap convincingly with the city Type category in our data. There is some
differentiation; for example in both cases there is a cluster (cluster 1 for
closed data, cluster 3 for open data) composed only of Open-Lightweight (OL)
cities. It's also clear that removing compositional closure in the data makes a
difference to both the size of the resulting K-means cluster and their
composition (in terms of cities belonging to each cluster). We haven't checked
the relationship of the other categories (Global, Region) to the clusters
obtained, but some editing of the code would answer this question for us!

## References and R Packages

Hu, J., Wang, Y., Taubenböck, H., Zhu, X.X. (2021). Land consumption in cities: A comparative study across the globe. *Cities*, **113**: 103163, <https://doi.org/10.1016/j.cities.2021.103163>.

Kassambara, A. and Mundt, F. (2020). *factoextra: Extract and Visualize the Results of Multivariate Data Analyses*. R package version 1.0.7. <https://CRAN.R-project.org/package=factoextra>

Maechler, M., Rousseeuw, P., Struyf, A., Hubert, M., Hornik, K.(2021). *cluster: Cluster Analysis Basics and Extensions*. R package version 2.1.2. <https://CRAN.R-project.org/package=cluster>

Reimann, C., Filzmoser, P., Garrett, R. G., & Dutter, R. (2008). *Statistical Data Analysis Explained: Applied Environmental Statistics with R* (First ed.). John Wiley & Sons, Chichester, UK.

Venables, W. N. & Ripley, B. D. (2002) *Modern Applied Statistics with S* (**MASS**). Fourth Edition. Springer, New York. ISBN 0-387-95457-0. <http://www.stats.ox.ac.uk/pub/MASS4/>

Wickham, H. (2019). *stringr: Simple, Consistent Wrappers for Common String Operations*. R package version 1.4.0. <https://CRAN.R-project.org/package=stringr>