---
documentclass: article
classoption: a4paper
geometry: margin=0.6in
output: 
  html_document: 
    toc_depth: 2
    fig_caption: yes
    number_sections: yes
    self_contained: no
  pdf_document: 
    fig_caption: TRUE
    number_sections: TRUE
    toc: no
    highlight: tango
  word_document: 
    toc_depth: 2
    fig_caption: yes
fontsize: 12pt
header-includes:
  \usepackage{sourcesanspro}
  \usepackage[T1]{fontenc}
  \renewcommand{\familydefault}{\sfdefault}
  \renewcommand{\thefigure}{0.\arabic{figure}}
  \renewcommand{\thetable}{0.\arabic{table}}
---

\setcounter{section}{-1}

```{r load knitr, include=FALSE}
library(flextable)
library(magrittr)
library(png)

set_flextable_defaults(theme_fun = "theme_zebra", 
                       font.size = 9, fonts_ignore = TRUE)
BorderDk <- officer::fp_border(color = "#B5C3DF", style = "solid", width = 1)
BorderLt <- officer::fp_border(color = "#FFFFFF", style = "solid", width = 1)

palette(c("black", "#003087", "#DAAA00", "#8F92C4", "#E5CF7E", 
          "#001D51", "#B7A99F", "#A51890", "#C5003E", "#FDC596", 
          "#AD5D1E", "gray40", "gray85", "#FFFFFF", "transparent"))

addImg <- function(obj, x = NULL, y = NULL, width = NULL, interpolate = TRUE){
  if(is.null(x) | is.null(y) | is.null(width)){stop("Must provide args 'x', 'y', and 'width'")}
  USR <- par()$usr ; PIN <- par()$pin ; DIM <- dim(obj) ; ARp <- DIM[1]/DIM[2]
  WIDi <- width/(USR[2]-USR[1])*PIN[1] ;   HEIi <- WIDi * ARp 
  HEIu <- HEIi/PIN[2]*(USR[4]-USR[3]) 
  rasterImage(image = obj, xleft = x-(width/2), xright = x+(width/2),
            ybottom = y-(HEIu/2), ytop = y+(HEIu/2), interpolate = interpolate)
}
```

```{r page 1 header hide code, fig.height=1.5, fig.width=10, echo=FALSE, out.width="100%", fig.align='right', results='hold'}
logo <- readPNG("UWA logo_text_V_wsL.png")
par(mar = c(0,0,0,0))
layout(matrix(c(1,1,1,1,2),nrow = 1))

plot(c(0,1),c(0,1), axes=F, type="n",xaxt="n", yaxt="n",ann=F)
text(-0.025,0.9, pos = 4, cex = 2.6, font = 2, 
     labels="ENVTM501 Data Analysis in R for Compositional Data")
text(-0.025,0.6, pos = 4, cex = 2.2, 
     labels="Multivariate analyses on cities land-use compositional data")
text(-0.025,0.4, pos = 4, font = 3, cex = 1.4, col = 12,
     labels="Land use proportional area data for 40 cities on Earth")
text(1,0.1, pos = 2, font = 3, family = 'serif', cex = 1.5, col = 2,
     labels="Andrew Rate, School of Agriculture and Environment")
plot(1,1, axes=F, type="n",xaxt="n", yaxt="n",ann=F)
addImg(logo, x = 1.2, y = 1, width = 0.5)
par(mar = c(3.5,3.5,0.5,0.5))
```

# Introduction

Compositional data, defined as any measurements which are proportions of some
whole, are common in many disciplines, including (Table 0.1):

```{r table of compositional data examples, tab.cap="Examples of compositional data by discipline.", echo=FALSE}
CoDas <- data.frame(Discipline = c("Environmental science","Geography",
                                   "Geology and geochemistry","Biology","Politics"),
                    Examples = c("Sediment or soil texture - proportions of mineral grains in particle size range fractions", "Proportions of land area subject to different land uses","Elemental composition of rocks","Plant species abundance as a proportion of ground cover","Percentage of voters supporting each candidate"))
ft <- flextable(CoDas)
ft <- theme_zebra(ft,odd_header = "#D0E0FF")
ft <- height_all(ft, height=0.2, unit = "in"); ft <- hrule(ft, rule = "exact")
ft <- width(ft, j=1:2, width=c(2,5), unit = "in")
ft <- border_outer(ft, border = BorderDk, part = "all")
ft %>% border_inner_v(border=BorderDk, part="header")
```

Compositional data have mathematical properties which result in misleading
conclusions from traditional statistical analyses. Compositional variables
(including concentrations of substances in water [or air, or rock, or soil, and
so on] or proportions of land surface coverage) are technically part of a
fixed-sum closed set. For example, with data on percent land use over an urban
area, all percentages add up to 100%! As Greenacre (2018) points out, we often 
deal with *subcompositions* rather than compositions, because we do not (or can 
not) measure all possible components of a sample. For example, in environmental 
science, when we measure the substances present in water, we don't usually 
measure the concentration of water itself. Or in politics, we report the 
proportions of voters who voted for each candidate, but not often the proportion 
of citizens who didn't vote.

The issue of subcompositions is not necessarily a problem. For instance, in the 
political example above, the ratio of voters supporting the two leading 
candidates will be the same whether we consider the total eligible electors, or 
only the people who actually voted. As we will see below, a key strategy for 
analysing compositional data is the use of some sort of *ratio*.

If uncorrected, analysis of data having fixed-sum closure can lead to very
misleading conclusions, especially when *relationships* between variables are
being investigated, as in correlation analyses or *multivariate methods* such as
principal component analysis. Closed data require specialised transformations to
remove closure, such as calculation of **centered or additive log ratios**
(Reimann et al. 2008). The centered log-ratio is defined by:

$x_{CLR} = log(\dfrac{x}{geomean(x_{1},...,x_{n})})$, where 

- $x$ is the value being transformed (concentration, proportion);
- $x_{CLR}$ is the centered log-ratio transformed value;
- $geomean$ means the geometric mean of;
- $x_{1}...x_{n}$ are all the $n$ values contributing to the (sub)composition, including the one ($x$) being transformed;
- all the components (parts) of the composition need to be in *identical units* (*e.g*. percent, mg/L, incidence per 10,000, *etc*.).

For example, in a soil texture analysis where the proportions of grain size 
categories gravel (G), sand (S), silt (Z), and clay (C) all add up to 100%, the 
CLR-transformed sand content is:

$S_{CLR} = log(\dfrac{S}{geomean(G,S,Z,C)})$

The example in \autoref{PFe} shows the relationship
between phosphorus (P) and iron (Fe) in soil/sediment materials in an acid
sulfate environment. Without correcting for compositional closure, the P *vs*.
Fe plot implies that P increases as Fe increases. Correcting for compositional
closure, however, suggests the opposite, with P negatively related to Fe! In
this case, if we had used conventional transformations, we might have come to a
very wrong conclusion about the sediment properties affecting phosphorus.

```{r closed open example P-Fe, echo=FALSE, fig.height=3.5, fig.width=7, message=FALSE, warning=FALSE, out.width="60%", fig.align='center', fig.cap="\\label{PFe}Comparison of relationships between P and Fe for (a) compositionally closed concentrations showing a positive relationship and (b) concentration variables corrected for compositional closure using centred log ratios showing a negative relationship. Data from Xu et al. (2018).", results='hold'}
nx.clr <- read.csv("nx_clr.csv", stringsAsFactors = TRUE)
par(mfrow=c(1,2), mar=c(3,3,1.,1), oma=c(0,0,0,0), mgp=c(1.4,0.1,0.0), lend=2, ljoin=1, 
    font.lab=2, cex.axis=1, cex.lab=1.1, cex=1., lwd=1, tcl=0.2)
palette(c("black","goldenrod2","royalblue4","gray92",
          "#EADBA4","#808BD6","gray50","#00083D","#51471D","gray24"))
# closed
plot(nx.clr$Fe.oes/1e3, nx.clr$P.oes, log="xy", pch=c(17,15)[nx.clr$Type],
     col=c(2,3)[nx.clr$Type],
     lwd=2, xlab=expression(bold(paste(Fe[closed]," (mg/kg)"%/%"1000"))),
     ylab=expression(bold(paste(P[closed]," (mg/kg)"))))
mtext(side=2, line=1.4, expression(bold(paste(P[closed]," (mg/kg)"))), cex=1.1)
mtext("(a)", 3, -1.25, adj = 0.05, font = 2, cex = 1.1)
legend("left", bty = "n", legend = c("Dredge spoil","Sediment"),
       pch = c(17,15), col = c(2,3), pt.lwd = c(2,1), pt.cex = c(1,1.1))
# clr
plot(nx.clr$Fe, nx.clr$P, pch=c(17,15)[nx.clr$Type], col=c(2,3)[nx.clr$Type], lwd=2,
     xlab=expression(bold(paste(Fe[open]," (clr-transformed)"))),
     ylab=expression(bold(paste(P[open]," (clr-transformed)"))))
mtext("(b)", 3, -1.25, adj = 0.95, font = 2, cex = 1.1)
```

In this UWA Micro-Credential course, you will learn specialised methods which
allow you to analyse compositional data so that results can be interpreted with
confidence. Students will also gain skills in use of R for relevant multivariate
data analyses.

First we load the packages we need, and set a preferred colour palette.

\scriptsize

```{r load packages etc., message=FALSE, warning=FALSE, results='hide'}
library(stringr)
library(knitr)
library(RcmdrMisc)
library(MASS)
library(rgr)
library(car)
library(cluster)
library(factoextra)
library(ggplot2)
library(reshape2)
library(ggpubr)

palette(c("black", "#003087", "#DAAA00", "#8F92C4", "#E5CF7E", 
          "#001D51", "#B7A99F", "#A51890", "#C5003E", "#FDC596", 
          "#AD5D1E", "gray40", "gray85", "#FFFFFF", "transparent"))
```

\normalsize

## Load the cities data

**Data are from**: Fig. 11 *in* Hu, J., Wang, Y., Taubenböck, H., Zhu, X.X., 2021. Land consumption in cities: A comparative study across the globe. *Cities*, **113**: 103163, <https://doi.org/10.1016/j.cities.2021.103163>. (digitised from figure using <https://automeris.io/WebPlotDigitizer/>)

After reading the data we make a column containing abbreviated codes for the
land-use type factor, using the \texttt{gsub()} function to substitute character
strings.

\small

```{r read file and show data, results='hold'}
cities <- 
  read.csv(paste0("https://raw.githubusercontent.com/Ratey-AtUWA/",
                  "compositional_data/main/cities_Hu_etal_2021.csv"), 
           stringsAsFactors = TRUE)
ft <- flextable(cities)
ft <- add_header_row(ft, values = c("","Land use categories","City categories"),
                     colwidths = c(1,4,3))
ft <- set_caption(ft, caption = "Cities land use data used in this document, showing cities 
      categorized by land-use Type, Global socioeconomic zone, and geographic Region.")
ft <- theme_zebra(ft,odd_header = "#D0E0FF", even_header = "#D0E0FF")
ft <- align(ft,align = "center", part = "header")
ft <- height_all(ft, height=0.39, unit = "cm"); ft <- hrule(ft, rule = "exact")
ft <- width(ft, j=1:8, width=c(2.7,1.5,1.5,1.8,1.5,3,2,3), unit = "cm")
ft <- border_outer(ft, border = BorderDk, part = "all")
ft %>% border_inner_v(border=BorderLt, part="header")
```

\normalsize

The original data categorised the cities by **Type**, based on the land use
categories occupying the greatest proportion(s) of urban area. We have added two
other categories, **Global** identifying each city as being in the socioeconomic
"global south" or "global north" zones, and also assigning each city to a
geographic **Region**, a category having eight levels. All categories are stored
as factors in the R data frame.

\scriptsize

```{r make new Type factor with abbreviated names, results='hide', echo=FALSE}
row.names(cities) <- as.character(cities$City)
cities$sType <- as.character(cities$Type)
cities$sType <- gsub("Compact-Open","CO",cities$sType)
cities$sType <- gsub("Open-Lightweight","OL",cities$sType)
cities$sType <- gsub("Compact","C",cities$sType)
cities$sType <- gsub("Open","O",cities$sType)
cities$sType <- gsub("Industrial","I",cities$sType)
cities$sType <- as.factor(cities$sType)
```

\normalsize

## Make CLR-transformed dataset

\scriptsize

```{r clr-transform data, paged.print=FALSE, results='hold'}
cities_clr <- cities
cities_clr[,c("Compact","Open","Lightweight","Industry")] <- 
  clr(cities_clr[,c("Compact","Open","Lightweight","Industry")])
head(cities_clr[,1:8])
```

\normalsize

## Compare correlation matrices

\scriptsize

```{r correlation matrix comnparison, fig.height=3.6, fig.width=7.2, out.width="70%", fig.align='center', fig.cap="\\label{corrmats}Correlation matrices for (a) closed, and (b) open (CLR-transformed) variables (code is not shown).", results='hold',echo=FALSE}
cormat <- round(cor(cities[,c("Compact","Open",
                              "Lightweight","Industry")]),2)
mltcor <- melt(cormat)
heatmap <- ggplot(mltcor, aes(Var2, Var1, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-1,1), space = "Lab", 
                       name="Pearson\nCorrelation") +
  theme_minimal()+ # minimal theme
  theme(axis.text.x = element_text(angle = 45, vjust = 1, 
                                   size = 12, hjust = 1)) +
  coord_fixed() + 
  geom_text(aes(Var2, Var1, label = value), color = "black", size = 4) +
  theme(axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    panel.grid.major = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank(),
    axis.ticks = element_blank()) +
  guides(fill = guide_colorbar(barwidth = 2, barheight = 7,
                title.position = "top", title.hjust = 0.5))

cormat <- round(cor(cities_clr[,c("Compact","Open",
                                  "Lightweight","Industry")]),2)
mltcor <- melt(cormat)
heatmap_clr <- ggplot(mltcor, aes(Var2, Var1, fill = value))+
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-1,1), space = "Lab", 
                       name="Pearson\nCorrelation") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, 
                                   size = 12, hjust = 1)) +
  coord_fixed() + 
  geom_text(aes(Var2, Var1, label = value), color = "black", size = 4) +
  theme(axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    panel.grid.major = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank(),
    axis.ticks = element_blank()) +
  guides(fill = guide_colorbar(barwidth = 2, barheight = 7,
                       title.position = "top", title.hjust = 0.5))
ggarrange(heatmap,heatmap_clr,labels=c("(a) closed","(b) open"),
          ncol=2,nrow=1)
```

\normalsize

By comparing \autoref{corrmats}(a) and \autoref{corrmats}(b), we can already see
that addressing closure with the centered-logratio (CLR) transformation changes
the interpretation of relationships between the variables.

**Some other things to try**

- comparing the distributions of untransformed and CLR-transformed pairs of variables using histograms or density plots
- comparing scatter plots of untransformed and CLR-transformed pairs of variables (maybe using the scatterplotMatrix() function in the **car** package).

## References and R Packages

Fox, J. (2022). *RcmdrMisc: R Commander Miscellaneous Functions*. R package
version 2.7-2. <https://CRAN.R-project.org/package=RcmdrMisc>

John Fox and Sanford Weisberg (2019). *An {R} Companion to Applied Regression*
(**car**), Third Edition. Thousand Oaks CA: Sage. URL:
<https://socialsciences.mcmaster.ca/jfox/Books/Companion/>

Garrett, R.G. (2018). *rgr: Applied Geochemistry EDA*. R package version 1.1.15.
<https://CRAN.R-project.org/package=rgr>

Greenacre, M. (2018). *Compositional Data Analysis in Practice*. CRC Press LLC, 
Boca Raton, FL, USA. 

Hu, J., Wang, Y., Taubenböck, H., Zhu, X.X. (2021). Land consumption in cities:
A comparative study across the globe. *Cities*, **113**: 103163,
<https://doi.org/10.1016/j.cities.2021.103163>.

Kassambara, A. and Mundt, F. (2020). *factoextra: Extract and Visualize the
Results of Multivariate Data Analyses*. R package version 1.0.7.
<https://CRAN.R-project.org/package=factoextra>

Maechler, M., Rousseeuw, P., Struyf, A., Hubert, M., Hornik, K.(2021). *cluster:
Cluster Analysis Basics and Extensions*. R package version 2.1.2.
<https://CRAN.R-project.org/package=cluster>

Reimann, C., Filzmoser, P., Garrett, R. G., & Dutter, R. (2008). *Statistical
Data Analysis Explained: Applied Environmental Statistics with R* (First ed.).
John Wiley & Sons, Chichester, UK.

Venables, W. N. & Ripley, B. D. (2002) *Modern Applied Statistics with S*
(**MASS**). Fourth Edition. Springer, New York. ISBN 0-387-95457-0.
<http://www.stats.ox.ac.uk/pub/MASS4/>

Wickham, H. (2019). *stringr: Simple, Consistent Wrappers for Common String
Operations*. R package version 1.4.0.
<https://CRAN.R-project.org/package=stringr>

Xu, N., Rate, A. W., & Morgan, B. (2018). From source to sink: Rare-earth
elements trace the legacy of sulfuric dredge spoils on estuarine sediments.
*Science of The Total Environment*, **637-638**, 1537-1549.
<https://doi.org/10.1016/j.scitotenv.2018.04.398>
