---
documentclass: article
classoption: a4paper
geometry: margin=0.6in
output: 
  html_document: 
    toc_depth: 2
    fig_caption: yes
    number_sections: yes
    self_contained: no
  pdf_document: 
    fig_caption: TRUE
    number_sections: TRUE
    toc: no
    highlight: tango
  word_document: 
    toc_depth: 2
    fig_caption: yes
fontsize: 12pt
header-includes:
  \usepackage{sourcesanspro}
  \usepackage[T1]{fontenc}
  \renewcommand{\familydefault}{\sfdefault}
  \renewcommand{\thefigure}{5.\arabic{figure}}
  \renewcommand{\thetable}{5.\arabic{table}}
---

\setcounter{section}{4}

```{r load knitr, include=FALSE}
library(flextable)
library(magrittr)
library(png)

set_flextable_defaults(theme_fun = "theme_zebra", 
                       font.size = 9, fonts_ignore = TRUE)
BorderDk <- officer::fp_border(color = "#B5C3DF", style = "solid", width = 1)
BorderLt <- officer::fp_border(color = "#FFFFFF", style = "solid", width = 1)

palette(c("black", "#003087", "#DAAA00", "#8F92C4", "#E5CF7E", 
          "#001D51", "#B7A99F", "#A51890", "#C5003E", "#FDC596", 
          "#AD5D1E", "gray40", "gray85", "#FFFFFF", "transparent"))

addImg <- function(obj, x = NULL, y = NULL, width = NULL, interpolate = TRUE){
  if(is.null(x) | is.null(y) | is.null(width)){stop("Must provide args 'x', 'y', and 'width'")}
  USR <- par()$usr ; PIN <- par()$pin ; DIM <- dim(obj) ; ARp <- DIM[1]/DIM[2]
  WIDi <- width/(USR[2]-USR[1])*PIN[1] ;   HEIi <- WIDi * ARp 
  HEIu <- HEIi/PIN[2]*(USR[4]-USR[3]) 
  rasterImage(image = obj, xleft = x-(width/2), xright = x+(width/2),
            ybottom = y-(HEIu/2), ytop = y+(HEIu/2), interpolate = interpolate)
}
```

```{r page 1 header hide code, fig.height=1.5, fig.width=10, echo=FALSE, out.width="100%", fig.align='right', results='hold'}
logo <- readPNG("UWA logo_text_V_wsL.png")
par(mar = c(0,0,0,0))
layout(matrix(c(1,1,1,1,2),nrow = 1))

plot(c(0,1),c(0,1), axes=F, type="n",xaxt="n", yaxt="n",ann=F)
text(-0.025,0.9, pos = 4, cex = 2.6, font = 2, 
     labels="ENVTM501 Data Analysis in R for Compositional Data")
text(-0.025,0.6, pos = 4, cex = 2.2, 
     labels="Multivariate analyses on whole-rock major element compositional data")
text(-0.025,0.4, pos = 4, font = 3, cex = 1.4, col = 12,
     labels="Geochemical data for the Yilgarn Craton, Western Australia by J.A. Hallberg")
text(1,0.1, pos = 2, font = 3, family = 'serif', cex = 1.5, col = 2,
     labels="Andrew Rate, School of Agriculture and Environment")
plot(1,1, axes=F, type="n",xaxt="n", yaxt="n",ann=F)
addImg(logo, x = 1.2, y = 1, width = 0.5)
par(mar = c(3.5,3.5,0.5,0.5))
```

```{r load packages etc., message=FALSE, warning=FALSE, include=FALSE, results='hide'}
library(rgr)
library(car)
library(cluster)
library(factoextra)
library(ggplot2)
library(reshape2)
library(ggpubr)
library(flextable)
library(magrittr)
library(RcmdrMisc)

set_flextable_defaults(theme_fun = "theme_zebra", 
                       font.size = 9, fonts_ignore = TRUE)
BorderDk <- officer::fp_border(color = "#B5C3DF", style = "solid", width = 1)
BorderLt <- officer::fp_border(color = "#FFFFFF", style = "solid", width = 1)

UWApal <- c("black", "#003087", "#DAAA00", "#8F92C4", "#E5CF7E", 
          "#001D51", "#B7A99F", "#A51890", "#C5003E", "#FDC596", 
          "#AD5D1E","gray40","gray85","#FFFFFF","transparent"); palette(UWApal)
```

```{r read file and show data, include=FALSE, results='hide'}
Hallberg <- read.csv("Hallberg.csv", stringsAsFactors = TRUE)
```

```{r make new Rock factor with abbreviated names, eval=FALSE, include=FALSE, results='hold'}
# row.names(Hallberg) <- paste0(as.character(Hallberg$Rock),seq(1:NROW(Hallberg)))
# Hallberg$sRock <- as.character(Hallberg$Rock)
# Hallberg$sRock <- gsub("Basaltic Trachyandesite","C",Hallberg$sRock)
# Hallberg$sRock <- gsub("Basaltic Andesite","L",Hallberg$sRock)
# Hallberg$sRock <- gsub("Andesite","A",Hallberg$sRock)
# Hallberg$sRock <- gsub("Trachybasalt","T",Hallberg$sRock)
# Hallberg$sRock <- gsub("Basalt","B",Hallberg$sRock)
# Hallberg$sRock <- gsub("Basanite","N",Hallberg$sRock)
# Hallberg$sRock <- gsub("Phonotephrite","P",Hallberg$sRock)
# Hallberg$sRock <- as.factor(Hallberg$sRock)
```

We are using a curated version of a whole rock major element dataset
from Hallberg
(<https://catalogue.data.wa.gov.au/dataset/hallberg-geochemistry>)

## Use *additive log-ratio* transformation to remove closure

The additive logratio (ALR) transformation calculates the logarithm of
the ratio of variables to one selected variable (which obviously then
can not be used for further data analysis. The variable used as the
denominator is often one that behaves "conservatively", such as
Al<sub>2</sub>O<sub>3</sub> or TiO<sub>2</sub> in geochemistry; see the
equation below (modified from Grunsky, 2010).

$ALR_{i} = log(C_{i}/C_{ref})$ for $(i = 1, ..., N-1)$,  where

- $ALR_{i}$ are the ALR-transformed variables
- $C_{i}$ are the
concentrations of the $i$ elements
- $C_{ref}$ is the concentration of the reference variable (denominator)
- $N$ is the total number of variables.

For multivariate *classification* methods, Campbell et al. (2009) argue
that the ALR (or related isometric log-ratio, ILR) transformation is
more appropriate than the centered log-ratio transformation (CLR) that
we have used so far.

```{r alr-tranform data, message=FALSE, warning=FALSE, include=TRUE, results='hold'}
Hallberg_alr <- Hallberg
Hallberg_alr[,c(11:12,14:24)] <- 
  alr(Hallberg_alr[,11:24], j = 3, ifwarn = FALSE) # recommend ifwarn = TRUE
```

# LDA -- Linear Discriminant Analysis

**Reminder**: To implement LDA in R, we use the **lda()** function in the
**MASS** package (Venables and Ripley, 2002). We specify the variables to be
considered using a formula similar to that used for multiple regression, and we
set the prior (initial) probabilities of an observation being in a particular
category at the actual frequencies at which they occur in the data.

We need scale variables to Z-scores, even if we have applied a log-ratio
transformation (*e.g*. ALR) to remove compositional closure.

## LDA on closed whole rock major element data

We will use LDA to discriminate the rock type, contained in the column
'**Rock**' in the Hallberg dataset. The variables we will use are the major
element oxide contents (except Al<sub>2</sub>O<sub>3</sub> which was our ALR
denominator), SiO<sub>2</sub>, TiO<sub>2</sub>, Fe<sub>2</sub>O<sub>2</sub>,
FeO, MnO, MgO, CaO, Na<sub>2</sub>O, K<sub>2</sub>O, &
P<sub>2</sub>O<sub>5</sub>. We're again excluding LOI, CO<sub>2</sub> and
H<sub>2</sub>O.

\scriptsize

```{r LDA closed whole rock data, results='hold'}
data0 <- Hallberg
data0[,c(11:24)] <- scale(data0[,11:24]) # scale just numeric variables
lda_rock_clos <- lda(formula = Rock ~ SiO2 + TiO2 + Fe2O3 + FeO + MnO + 
                       MgO + CaO + Na2O + K2O + P2O5, 
                    data = data0,
                    prior = as.numeric(summary(Hallberg$Rock))/
                      nrow(Hallberg)) 
print(lda_rock_clos)
```

\scriptsize

## LDA on open (ALR) whole rock major element data

```{r LDA open ALR whole rock data, message=FALSE, warning=FALSE, results='hold'}
data0 <- Hallberg_alr
data0[,11:24] <- scale(data0[,11:24]) # scale just numeric variables
lda_rock_open <- lda(formula = Rock ~ SiO2 + TiO2 + Fe2O3 + FeO + MnO + 
                       MgO + CaO + Na2O + K2O + P2O5, 
                    data = data0,
                    prior = as.numeric(summary(data0$Rock))/nrow(data0)) 
print(lda_rock_open)
```

```{r correl matrix ALR not used in this version, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
cormat <- rcorr.adjust(data0[,11:24], 
                       type="pearson")
cat("\nCorrelation matrix for predictors (not part of LDA output):\n"); print(cormat$R$r, digits = 2)
rm(cormat)
```

\normalsize

The LDA model specified and the prior probabilities are reported back to
us in the output. The *Group means* sub-table in the output contains the
centroids of the categories (5 classes, in our whole rock data) in the
n-dimensional space defined by the n variables used for classification.
The *Coefficients of linear* *discriminants* sub-table essentially
defines the linear discriminant functions that separate our categories,
since each function (LD1, LD2, *etc*.) is a linear combination of all
the variables. Finally, the *Proportion of trace* sub-table gives the
*proportions of between-class variance* that are explained by successive
discriminant functions (*e.g*. for the open Hallberg rock data, LD1
explains 0.717 (72%) and LD2 explains 0.171 (17%) of variance between
Rock type categories).

[**Note**: sometimes we cannot make an LDA model, because the predictors
are collinear (highly correlated). We may be able to fix this by
inspecting a correlation matrix for all the predictor variables, and
removing *one* variable at a time from correlated pairs, then re-running
the LDA procedure. By analogy with multiple linear regression, this
would mean a Pearson's r $\ge$ 0.8.]

## Visualising LDA separation

### LDA histograms

We can plot the separation achieved by each linear discriminant (LD)
function by predicting the classification using the input data, then
using the **ldahist** function (Venables and Ripley 2002). To see the
separation in another LDA dimension, we change the subscript in the
**predClos\$x[,1]** option. Histograms (actually drawn with custom code
to plot side-by side) are shown in \autoref{ldahistComp}.

\scriptsize

```{r lda hist compare, fig.align='center', fig.cap="\\label{ldahistComp}Histograms based on the first linear discriminant function for (a) closed and (b) open (ALR-transformed) whole-rock major element data.", fig.height=14, fig.width=10, out.width="80%", results='hold'}
predClos <- predict(lda_rock_clos, Hallberg[,11:21])
predOpen <- predict(lda_rock_open, Hallberg[,11:21])
LD1c <- data.frame(Rock=as.character(Hallberg$Rock),LD1=predClos$x[,1])
LD1c$Rock <- factor(LD1c$Rock, levels=levels(Hallberg$Rock))
par(mfcol = c(nlevels(LD1c$Rock),2), mar = c(1,2,1,1), oma = c(1,0,1,0), 
    mgp = c(0.75,0.2,0), tcl=0.15)
for (i in 1:nlevels(LD1c$Rock)){
  with(subset(LD1c, subset=LD1c$Rock==levels(LD1c$Rock)[i]),
       hist(LD1, main = "", breaks = pretty(LD1c$LD1, n=20), col=5,
       xlim = c(min(LD1c$LD1, na.rm=T),max(LD1c$LD1, na.rm=T))))
  box()
  mtext(levels(LD1c$Rock)[i],3,-1.55,adj=0.505, cex = 0.85, font = 2, col=14)
  mtext(levels(LD1c$Rock)[i],3,-1.5, cex = 0.85, font = 2, col = 11)
  if(i==1) mtext("(a) Closed data", 3, 0.5, font=2, cex = 1.4, col = 11)
}

LD1o <- data.frame(Rock=as.character(Hallberg$Rock),LD1=predOpen$x[,1])
LD1o$Rock <- factor(LD1o$Rock, levels=levels(Hallberg$Rock))
for (i in 1:nlevels(LD1o$Rock)){
  with(subset(LD1o, subset=LD1o$Rock==levels(LD1o$Rock)[i]),
       hist(LD1, main = "", breaks = pretty(LD1o$LD1, n=20), col=4,
            xlim = c(min(LD1o$LD1, na.rm=T),max(LD1o$LD1, na.rm=T))))
  box()
  mtext(levels(LD1o$Rock)[i],3,-1.55, adj=0.505, cex = 0.85, font = 2, col=14)
  mtext(levels(LD1o$Rock)[i],3,-1.5, cex = 0.85, font = 2, col = 2)
  if(i==1) mtext("(b) Open data", 3, 0.5, font=2, cex = 1.4, col = 2)
}
```

\normalsize

The sets of histograms for closed and open data in \autoref{ldahistComp}
both show some separation of categories, but with overlap. Of course
this only shows the separation in one dimension, and two or more
dimensions may be needed to achieve clear separation. We will make plots
showing more than one LDA dimension later.

### Partition plots

Another potentially useful way of showing separation of groups in LDA is
to use a *partition plot*, accessible using the **partimat()** function
from the klaR R package (Weihs et al.v 2005).

\scriptsize

```{r pplot comparison, fig.align='center', fig.height=5, fig.width=10, message=FALSE, warning=FALSE, fig.cap="\\label{pplotComp}Partition plots for (a) closed and (b) open (ALR-transformed) whole rock major element data, based on the SiO2 and TiO2 contents. Filled symbols are means in each category, with red letters showing apparently mis-classified observations.", results='hold', out.width="100%"}
require(klaR)
par(mfrow = c(1,2),mar = c(3,3,1,1), mgp= c(1.3,0.2,0), tcl=0.2, font.lab=2)
with(Hallberg, 
     drawparti(Rock, SiO2, TiO2, method="lda",
               image.colors = rainbow(8,s=0.2,v=0.85,end=0.8),
               xlab = expression(bold(paste(SiO[2]," (%)"))), 
               ylab = expression(bold(paste(TiO[2]," (%)"))))
     )
mtext("(a)", 3, -1.5, adj=0.05, font=2, cex = 1.2, col=14)
with(Hallberg_alr, 
     drawparti(Rock, SiO2, TiO2, method="lda",
               image.colors = rainbow(8,s=0.2,v=0.85,end=0.8),
               xlab = expression(bold(paste(SiO[2]," (ALR-transformed)"))), 
               ylab = expression(bold(paste(TiO[2]," (ALR-transformed)"))))
     )
mtext("(b)", 3, -1.5, adj=0.05, font=2, cex = 1.2)
```

\normalsize

Partition plots, such as those in \autoref{pplotComp}, are presented for
single pairwise combinations of the variables (in this example
SiO<sub>2</sub> and TiO<sub>2</sub>) used to make the LDA model. We can
make different plots by specifying different variables in the
**drawparti()** function. Each such plot can be considered to be a
different view of the data (which of course has multiple dimensions).
Colored regions delineate each classification area. Any observation that
falls within a region is predicted to be from a specific category, with
apparent mis-classification in a different color (but we usually need
more than two dimensions for correct classification). Each plot also
includes the apparent error rate for that view of the data.

## Scatter Plots resembling biplots

Scatter-plots showing each variable and observation in linear
discriminant dimensions, and grouped by category, are useful for visual
assessment of how well the LDA model separates the observations.

\scriptsize

```{r plot LDA closed, fig.height=8, fig.width=8, fig.align='center', out.width="80%", fig.cap="\\label{LDAclos}Linear discriminant analysis (LDA) plots for closed rock composition data: (a) variable coefficients in LD1-LD2 space, and predictions for observations in (b) LD1-LD2 space; (c) LD1-LD3 space; (d) LS2-LD4 space. Legend in (d) also applies to (b) and (c).", results='hold'}
par(mfrow = c(2,2), mar = c(3.5,3.5,1,1), mgp = c(1.5,0.3,0), tcl = 0.25,
    lend = "square", ljoin = "mitre", cex.main = 0.9, font.lab=2)
plot(lda_rock_clos$scaling[,1], lda_rock_clos$scaling[,2],
     xlim = c(-4.2,1), 
     xlab="Linear Discriminant [1]", ylab="Linear Discriminant [2]", 
     main="(a) Variable Coefficients [LD1, LD2]")
abline(v=0,col="grey",lty=2)
abline(h=0,col="grey",lty=2)
text(lda_rock_clos$scaling[,1],lda_rock_clos$scaling[,2],
     labels=names(Hallberg)[c(11,12,14:21)],
     pos=c(2,2,4),cex=0.95,col="blue2",offset=0.2)
mtext("(a)", 3, -1.5, adj = 0.45, cex = 1.2, font = 2)

ldaPred_rock_clos <- predict(lda_rock_clos)
for(i in 1:NROW(lda_rock_clos$scaling)){
  arrows(0,0,lda_rock_clos$scaling[i,1],lda_rock_clos$scaling[i,2],
         length = 0.1, col = 7)
}
legend("bottomleft",legend=levels(Hallberg$Rock),
       ncol = 1, bty="o", inset=0.015, col=c(1:3,6,8,9,11,12), 
       title="Rock Type in (b) - (d)", box.col=15,
       pch=c(0:7), pt.lwd = 2, pt.cex = 1.5, cex = 1.1)

plot(ldaPred_rock_clos$x[,1], ldaPred_rock_clos$x[,2],
     col=c(1:3,6,8,9,11,12)[Hallberg$Rock],
     pch=c(0:7)[Hallberg$Rock],
     lwd = 2, cex = 1.5, 
  xlab="Linear Discriminant [1]", ylab="Linear Discriminant [2]", 
  main="Predictions for Observations [LD1, LD2]")
abline(v=0,col="grey",lty=2)
abline(h=0,col="grey",lty=2)
# text(ldaPred_rock_clos$x[,1], ldaPred_rock_clos$x[,2], 
#      labels=as.character(Hallberg$Rock), col=c(2,4,6,3,12)[Hallberg$Rock],
#      pos=1, offset=0.15, cex=0.65)
# legend("bottomright",legend=levels(Hallberg$Rock)[6:13],
#        ncol = 2, col=c(6:13), pch=c(5:12), pt.lwd = 2,
#        title=expression(bold("Rock Type")),
#        bty="n", inset=0.01, 
#        pt.cex=c(1.8,1.8,2,2,1.3), cex=0.9)
mtext("(b)", 3, -1.5, adj = 0.95, cex = 1.2, font = 2)

plot(ldaPred_rock_clos$x[,1], ldaPred_rock_clos$x[,3], 
     col=c(1:3,6,8,9,11,12)[Hallberg$Rock],
     pch=c(0:7)[Hallberg$Rock], 
     lwd = 2, cex = 1.5, 
  xlab="Linear Discriminant [1]", ylab="Linear Discriminant [3]", 
  main="Predictions for Observations [LD1, LD3]")
abline(v=0,col="grey",lty=2)
abline(h=0,col="grey",lty=2)
# text(ldaPred_rock_clos$x[,1], ldaPred_rock_clos$x[,3], 
#      labels=Hallberg$Rock, col=c(1:13)[Hallberg$Rock],
#      pos=1, offset=0.15, cex=0.65)
mtext("(c)", 3, -1.5, adj = 0.05, cex = 1.2, font = 2)

plot(ldaPred_rock_clos$x[,2], ldaPred_rock_clos$x[,4],
     col=c(1:3,6,8,9,11,12)[Hallberg$Rock],
     pch=c(0:7)[Hallberg$Rock], 
     lwd = 2, cex = 1.5, 
  xlab="Linear Discriminant [2]", ylab="Linear Discriminant [4]", 
  main="Predictions for Observations [LD2, LD4]")
abline(v=0,col="grey",lty=2)
abline(h=0,col="grey",lty=2)
# text(ldaPred_rock_clos$x[,2], ldaPred_rock_clos$x[,4], 
#      labels=Hallberg$Rock, col=c(2,4,6,3,12)[Hallberg$Rock],
#      pos=1, offset=0.15, cex=0.65)
mtext("(d)", 3, -1.5, adj = 0.55, cex = 1.2, font = 2)
```

```{r plot LDA open, fig.height=8, fig.width=8, out.width="70%", fig.align='center', fig.cap="\\label{LDAopen}Linear discriminant analysis (LDA) plots for open (ALR-transformed) rock composition data: (a) variable coefficients in LD1-LD2 space, and predictions for observations in (b) LD1-LD2 space; (c) LD1-LD3 space; (d) LS2-LD3 space. Legend in (a) applies to plots (b), (c), and (d).", results='hold'}
par(mfrow = c(2,2), mar = c(3.5,3.5,1,1), mgp = c(1.5,0.3,0), tcl = 0.25,
    lend = "square", ljoin = "mitre", cex.main = 0.9, font.lab=2)
plot(lda_rock_open$scaling[,1], lda_rock_open$scaling[,2], 
     xlim = c(-2.8,0.6), ylim = c(-1.3,1.3), 
     xlab="Linear Discriminant [1]", ylab="Linear Discriminant [2]", 
     main="Variable Coefficients [LD1, LD2]")
abline(v=0,col="grey",lty=2); abline(h=0,col="grey",lty=2)
text(lda_rock_open$scaling[,1], lda_rock_open$scaling[,2], 
     labels=names(Hallberg_alr)[11:21],
     pos = c(1,3,2,4), cex = 0.9, col = 2, offset=0.2)
for(i in 1:NROW(lda_rock_open$scaling)){
  arrows(0,0,lda_rock_open$scaling[i,1],lda_rock_open$scaling[i,2],
         length = 0.1, col = 7) }
mtext("(a)", 3, -1.5, adj = 0.95, cex = 1.2, font = 2)
legend("bottomleft", ncol = 1, legend=levels(Hallberg_alr$Rock), 
       col=c(1:3,6,8,9,11,12), pch=c(0:7), pt.lwd = 2,
       bty="o", box.col=15, inset=0.001, x.intersp = 1, 
       title=expression(bold("Rock Type in (b) - (d)")),
       box.lwd=2, pt.cex=1.5, cex=1.1, title.cex = 1.1)

ldaPred_rock_open <- predict(lda_rock_open)

plot(ldaPred_rock_open$x[,1], ldaPred_rock_open$x[,2], 
     col=c(1:3,6,8,9,11,12)[Hallberg_alr$Rock],
     pch=c(0:7)[Hallberg_alr$Rock], lwd=2, 
     cex = 1.5, 
     main="Predictions for Observations [LD1, LD2]", 
     xlab="Linear Discriminant [1]", ylab="Linear Discriminant [2]")
abline(v=0,col="grey",lty=2); abline(h=0,col="grey",lty=2)
# text(ldaPred_rock_open$x[,1], ldaPred_rock_open$x[,2], labels=Hallberg$Rock, 
#      col=c(2,4,6,3,12)[Hallberg$Rock], pos=1, offset=0.15, cex=0.65)
mtext("(b)", 3, -1.5, adj = 0.95, cex = 1.2, font = 2)

plot(ldaPred_rock_open$x[,1], ldaPred_rock_open$x[,3], 
     col=c(1:3,6,8,9,11,12)[Hallberg_alr$Rock],
     pch=c(0:7)[Hallberg_alr$Rock], lwd=2, 
     cex = 1.5, 
     main="Predictions for Observations [LD1, LD3]", 
     xlab="Linear Discriminant [1]", ylab="Linear Discriminant [3]")
abline(v=0,col="grey",lty=2); abline(h=0,col="grey",lty=2)
# text(ldaPred_rock_open$x[,1], ldaPred_rock_open$x[,3], labels=Hallberg$Rock, 
#      col=c(2,4,6,3,12)[Hallberg$Rock], pos=1, offset=0.15, cex=0.65)
mtext("(c)", 3, -1.5, adj = 0.05, cex = 1.2, font = 2)

plot(ldaPred_rock_open$x[,2], ldaPred_rock_open$x[,4], 
     col=c(1:3,6,8,9,11,12)[Hallberg_alr$Rock],
     pch=c(0:7)[Hallberg_alr$Rock], lwd=2, 
     cex = 1.5, 
     main="Predictions for Observations [LD2, LD4]", 
     xlab="Linear Discriminant [2]", ylab="Linear Discriminant [4]")
abline(v=0,col="grey",lty=2); abline(h=0,col="grey",lty=2)
# text(ldaPred_rock_open$x[,2], ldaPred_rock_open$x[,3], labels=Hallberg$Rock, 
#      col=c(2,4,6,3,12)[Hallberg$Rock], pos=1, offset=0.15, cex=0.65)
mtext("(d)", 3, -1.5, adj = 0.95, cex = 1.2, font = 2)
```

```{r reset par to 1x1, message=FALSE, warning=FALSE, include=FALSE, results='hide'}
par(mfrow=c(1,1))
```

\normalsize

From the plots in \autoref{LDAclos} and \autoref{LDAopen}, we can see
that the LDA models obtained do separate observations by the selected
factor. Firstly, however, there is a lot of clustering of the predictor
variables in \autoref{LDAclos}(a), which may relate to spurious
relationships between variables caused by compositional closure. This
clustering is not so pronounced in \autoref{LDAopen}(a), since the
closure has been removed by ALR-transformation.

Out of the combinations of LDA dimensions selected, the best separation
is with LD2 *vs*. LD1, which is not surprising since these dimensions
together account for about 95% of the between-groups variance -- for
both closed and open data. There is a lot of apparent overlap, but we
can not see the true separation in only 2 dimensions. The LDA performed
on open data may result in slightly clearer separation of samples by
Rock category than LDA using closed data, but without a multidimensional
view this is also hard to be sure about.

One way that we can get a better idea about the usefulness of our
classification models is to perform some validation. This involves
'training' the model on a subset of our data, and trying to predict the
category of a different subset using the training model. A later session
will look at some ways we can do this.

## References

Campbell, G. P., Curran, J. M., Miskelly, G. M., Coulson, S., Yaxley, G.
M., Grunsky, E. C., & Simon C. Cox. (2009). Compositional data analysis
for elemental data in forensic science. *Forensic Science
International*, **188**, 81-90.
<https://doi.org/10.1016/j.forsciint.2009.03.018>

Grunsky, E. C. (2010). The interpretation of geochemical survey data.
*Geochemistry: Exploration, Environment, Analysis*, **10**, 27-74.
<https://doi.org/10.1144/1467-7873/09-210>
