---
documentclass: article
classoption: a4paper
geometry: margin=0.6in
output: 
  pdf_document: 
    fig_caption: TRUE
    number_sections: TRUE
    toc: no
    highlight: tango
  html_document: 
    toc_depth: 2
    fig_caption: yes
    number_sections: yes
    self_contained: no
  word_document: 
    toc_depth: 2
    fig_caption: yes
fontsize: 12pt
header-includes:
  \usepackage{sourcesanspro}
  \usepackage[T1]{fontenc}
  \renewcommand{\familydefault}{\sfdefault}
  \renewcommand{\thefigure}{1.\arabic{figure}}
  \renewcommand{\thetable}{1.\arabic{table}}
---

\setcounter{section}{0}

```{r load knitr, include=FALSE}
library(flextable)
library(magrittr)
library(png)

set_flextable_defaults(theme_fun = "theme_zebra", 
                       font.size = 9, fonts_ignore = TRUE)
BorderDk <- officer::fp_border(color = "#B5C3DF", style = "solid", width = 1)
BorderLt <- officer::fp_border(color = "#FFFFFF", style = "solid", width = 1)

palette(c("black", "#003087", "#DAAA00", "#8F92C4", "#E5CF7E", 
          "#001D51", "#B7A99F", "#A51890", "#C5003E", "#FDC596", 
          "#AD5D1E", "gray40", "gray85", "#FFFFFF", "transparent"))

addImg <- function(obj, x = NULL, y = NULL, width = NULL, interpolate = TRUE){
  if(is.null(x) | is.null(y) | is.null(width)){stop("Must provide args 'x', 'y', and 'width'")}
  USR <- par()$usr ; PIN <- par()$pin ; DIM <- dim(obj) ; ARp <- DIM[1]/DIM[2]
  WIDi <- width/(USR[2]-USR[1])*PIN[1] ;   HEIi <- WIDi * ARp 
  HEIu <- HEIi/PIN[2]*(USR[4]-USR[3]) 
  rasterImage(image = obj, xleft = x-(width/2), xright = x+(width/2),
            ybottom = y-(HEIu/2), ytop = y+(HEIu/2), interpolate = interpolate)
}
```

```{r page 1 header hide code, fig.height=1.5, fig.width=10, echo=FALSE, out.width="100%", fig.align='right', results='hold'}
logo <- readPNG("UWA logo_text_V_wsL.png")
par(mar = c(0,0,0,0))
layout(matrix(c(1,1,1,1,2),nrow = 1))

plot(c(0,1),c(0,1), axes=F, type="n",xaxt="n", yaxt="n",ann=F)
text(-0.025,0.9, pos = 4, cex = 2.6, font = 2, 
     labels="ENVTM501 Data Analysis in R for Compositional Data")
text(-0.025,0.6, pos = 4, cex = 2.2, 
     labels="Multivariate analyses on cities land-use compositional data")
text(-0.025,0.4, pos = 4, font = 3, cex = 1.4, col = 12,
     labels="Land use proportional area data for 40 cities on Earth")
text(1,0.1, pos = 2, font = 3, family = 'serif', cex = 1.5, col = 2,
     labels="Andrew Rate, School of Agriculture and Environment")
plot(1,1, axes=F, type="n",xaxt="n", yaxt="n",ann=F)
addImg(logo, x = 1.2, y = 1, width = 0.5)
par(mar = c(3.5,3.5,0.5,0.5))
```

```{r load packages etc., message=FALSE, warning=FALSE, include=FALSE, results='hide'}
library(stringr)
library(knitr)
library(RcmdrMisc)
library(MASS)
library(rgr)
library(car)
library(cluster)
library(factoextra)
library(ggplot2)
library(reshape2)
library(ggpubr)

palette(c("black", "#003087", "#DAAA00", "#8F92C4", "#E5CF7E", 
          "#001D51", "#B7A99F", "#A51890", "#C5003E", "#FDC596", 
          "#AD5D1E", "gray40", "gray85", "#FFFFFF", "transparent"))
```

```{r read file and format data, include=FALSE, results='hold'}
cities <- 
  read.csv(paste0("https://raw.githubusercontent.com/Ratey-AtUWA/",
                  "compositional_data/main/cities_Hu_etal_2021.csv"), 
           stringsAsFactors = TRUE)
row.names(cities) <- as.character(cities$City)
cities$sType <- as.character(cities$Type)
cities$sType <- gsub("Compact-Open","CO",cities$sType)
cities$sType <- gsub("Open-Lightweight","OL",cities$sType)
cities$sType <- gsub("Compact","C",cities$sType)
cities$sType <- gsub("Open","O",cities$sType)
cities$sType <- gsub("Industrial","I",cities$sType)
cities$sType <- as.factor(cities$sType)
```

```{r clr-transform data, include=FALSE, paged.print=FALSE, results='hold'}
cities_clr <- cities
cities_clr[,c("Compact","Open","Lightweight","Industry")] <- 
  clr(cities_clr[,c("Compact","Open","Lightweight","Industry")])
head(cities_clr[,1:8])
```

# Principal components analysis

It is quite common to measure many variables in environmental science and other
disciplines. Using various types of *ordination analysis*, we can use the
information contained in multiple variables to create a reduced subset of
variables containing nearly the same amount of information. Ordination methods
are also referred to, for this reason, as 'data reduction' methods and are
commonly used for multivariate analysis.

One of the earliest and most widely used ordination methods for exploration and
dimension-reduction of multivariate data is **principal components analysis**
(PCA). Imagine a dataset with many samples (rows) and n continuous numeric
variables (columns) which contain quantitative information about each sample
such as concentrations, heights, velocities, *etc*. For these *n*
variables/dimensions, the principal component calculation generates *n* new
variables, or principal components, which are each a function of the set of all
the original variables (so each principal component is defined by a weighting or
coefficient for each of the original variables). We may choose to omit some
variables from the analysis if they contain too many missing observations or if
there is another valid reason to doubt their integrity. Since each principal
component is selected to account for successively smaller proportions of the
multiple variance, it is usually the first few principal components which
explain most of the variance and therefore contain the most useful information.
We conventionally visualize this in a 'scree plot' (\autoref{screeplots}), a
kind of bar graph showing the decrease in variance accounted for by each
component.

\scriptsize

```{r run and summarise PCA, results='hold'}
#
data0 <- na.omit(cities[,c("Compact","Open","Lightweight","Industry")])
pca_cities_clos <- prcomp(data0, scale. = TRUE)
cat("Variable weightings (rotations) - closed data\n")
pca_cities_clos$rot
cat("...\n\nComponent Variances - Closed data\n")
pca_cities_clos$sdev^2
cat("...\n\nProportions of variance explained by each component",
    "\nCLR-transformed (open) data\n")
round(pca_cities_clos$sdev^2/sum(pca_cities_clos$sdev^2),3)
cat("\n--------------------\nVariable weightings (rotations) - open data\n")

data0 <- na.omit(cities_clr[,c("Compact","Open","Lightweight","Industry")])
pca_cities_open <- prcomp(data0, scale. = TRUE)
pca_cities_open$rot
cat("...\n\nComponent Variances - CLR-transformed (open) data\n")
pca_cities_open$sdev^2
cat("...\n\nProportions of variance explained by each component",
    "\nCLR-transformed (open) data\n")
round(pca_cities_open$sdev^2/sum(pca_cities_open$sdev^2),3)
rm(data0)
```

\normalsize

## Principal Components Analysis output

As well as the component variances, the useful results of principal components
analysis include the variable weightings or 'rotations' for each principal
component. Referring to just the output for **open data** above, PC1 (which
explains the most multivariate variance) has its greatest absolute contributions
from the Lightweight and Industry variables, which act in opposite directions
(remembering that the sign is arbitrary, but that *relative* sign and magnitude
are important). PC2 contains mainly variance from the Compact and Open
variables, PC3 (which explains much less variance) reflects Open/Industry, and
we can ignore PC4 as it has a negligible component variance.

In addition, every individual observation (sample) is a multivariate point, the
observation scores for all samples in each principal component based on the
values of the variables used in PCA for that sample. It is conventional to plot
both of these two types of output in a principal component **biplot**, as shown
in \autoref{PCAplots}. Before discussing the biplot, we should note that the
sign (positive or negative) of variable weightings and observation scores
(*i.e*. the direction of the axes) is arbitrary and should not affect our
interpretations.

\scriptsize

```{r scree plots, echo=FALSE, fig.height=2.5, fig.width=6, message=FALSE, warning=FALSE, out.width='60%', fig.align='center', fig.cap="\\label{screeplots}PCA scree plots for urban land-use data, for: (a) compositionally closed proportions; (b) data corrected for closure using CLR-transformation.", results='hold'}
par(mar = c(3.,3.5,1.2,1), mfrow = c(1,2), mgp=c(2,0.8,0), font.lab=2)
plot(pca_cities_clos, main = "(a) closed", cex.main = 1, col = 5)
axis(1, at=seq(0.8,4.2,l=4), labels=seq(1,4,1), tcl=0.01, col=15, mgp=c(0.5,0.1,0))
mtext("Component", side = 1, line = 1.15, font = 2)
plot(pca_cities_open, main = "(b) open (CLR)", cex.main = 1, col = 4)
axis(1, at=seq(0.8,4.2,l=4), labels=seq(1,4,1), tcl=0.01, col=15, mgp=c(0.5,0.1,0))
mtext("Component", side = 1, line = 1.15, font = 2)
par(mar = c(3.5,3.5,1,1), mfrow = c(1,2), mgp=c(1.7,0.3,0))
```

\normalsize

There are informal 'rules' for deciding which of the principal components 
contain useful information:

1. Components having variance greater than 1 (the 'Kaiser' criterion)
2. Components up to a cumulative proportion of variance explained of 0.8 (80%)

For example, for the open data, the component variances are PC1 1.976, PC2
1.511, PC3 0.513, PC4 1.5E-31 -- so only PC1 and PC2 meet the first criterion.
Similarly, the *cumulative* proportions of variance explained by each component
are PC1 0.494, PC1+PC2 0.872, so by criterion 2 no useful information is
contained in principal components >2. In this example both criteria agree (they
don't always!).

Another part of the PCA output that we need to know about is the set of
*observation scores*, essentially showing where each observation (each city, in
these data) plots in principal component space. We're only interested in PC1 and
PC2, since they are the only components meeting the criteria. The observation
scores for the first rows of the data are as follows (\autoref{PCAObsTabl}):

```{r table of PCA obs scores, tab.cap="\\label{PCAObsTabl}PCA observation scores for the first 8 rows of the cities data (closed and open)", results='hold'}
OScores <- data.frame(City=cities_clr$City[1:8],
                      PC1_closed=round(pca_cities_clos$x[1:8,1],3),
                      PC2_closed=round(pca_cities_clos$x[1:8,2],3),
                      PC1_open=round(pca_cities_open$x[1:8,1],3),
                      PC2_open=round(pca_cities_open$x[1:8,2],3))
ft <- flextable(OScores)
ft <- theme_zebra(ft,odd_header = "#D0E0FF", even_header = "#D0E0FF")
ft <- border_outer(ft, border = BorderDk, part = "all")
ft<- border_inner_v(ft, border=BorderLt, part="header"); ft
```

## Principal components biplots

\scriptsize

```{r visualise PCA, fig.height=6, fig.width=12, fig.align='center', fig.cap="\\label{PCAplots}PCA biplots for urban land-use data, with observations categorised by type for: (a) compositionally closed proportions; (b) data corrected for closure using CLR-transformation.", results='hold'}

require(car) # for data ellipses (see Fig 1.4)
par(mfrow=c(1,2), mar = c(3.5,3.5,3.5,3.5), oma = c(0,0,0,0), 
    mgp=c(1.7,0.3,0), tcl = 0.25, font.lab=2,
    lend = "square", ljoin = "mitre")
# choose components and set scaling factor (sf)
v1 <- 1; v2 <- 2; sf <- 0.45

biplot(pca_cities_clos, choices = c(v1,v2), col = c(2,1), cex=c(1,0.0),
       pc.biplot = FALSE, scale = 0.4, arrow.len = 0.08,
       xlim = c(-1.5,1.5), ylim = c(-1.2,1.2),
       xlab = paste0("Scaled PC",v1," Component Loadings"),
       ylab = paste0("Scaled PC",v2," Component Loadings"))
mtext(paste0("Scaled PC",v1," Observation Scores"), 3, 1.6, font = 2)
mtext(paste0("Scaled PC",v2," Observation Scores"), 4, 1.6, font = 2)
mtext("Untransformed data\n(compositionally closed)",
      side = 3, line = -2, font = 2, adj = 0.98)
data0 <- na.omit(cities[,c("Type","Global","Region","Compact",
                           "Open","Lightweight","Industry")])
points(pca_cities_clos$x[,v1]*sf, pca_cities_clos$x[,v2]*sf*1.5,
       pch = c(22,21,24,0,1)[data0$Type],
       lwd = 2, bg = c(2,3,4,5,6)[data0$Type],
       col = c(2,3,8,9,11)[data0$Type],
       cex = c(1.2,1.4,1.2,1.2,1.4)[data0$Type])
dataEllipse(x=pca_cities_clos$x[,v1]*sf, y=pca_cities_clos$x[,v2]*sf*1.5,
            groups = data0$Type, add = TRUE, plot.points = FALSE, 
            levels = c(0.9), center.pch = 3, col = c(2,3,8,9,11),
            lty = 2, lwd = 1, center.cex = 2.4, group.labels = "")
legend("bottomright", bty = "o", inset = 0.03,
       box.col = "gray", box.lwd = 2, bg = 14,
       legend = levels(data0$Type),
       pch = c(22,21,24,0,1), pt.lwd = 2,
       col = c(2,3,8,9,11), pt.bg = c(2,3,4,5,6),
       pt.cex = c(1.2, 1.4, 1.2,1.2,1.4),
       cex = 0.9, y.intersp = 0.9)
mtext("(a)", side = 3, line = -1.5, font = 2, adj = 0.02, cex = 1.25)

sf <- 0.65 # adjust scaling factor for next PCA biplot

biplot(pca_cities_open, choices = c(v1,v2), col = c(2,1), cex=c(0.8,1),
       pc.biplot = FALSE, scale = 0.2, arrow.len = 0.08,
       xlim = c(-1.2,3.2), ylim = c(-3.5,1.7),
       xlab = paste0("Scaled PC",v1," Component Loadings"),
       ylab = paste0("Scaled PC",v2," Component Loadings")) 
mtext(paste0("Scaled PC",v1," Observation Scores"), 3, 1.6, font = 2)
mtext(paste0("Scaled PC",v2," Observation Scores"), 4, 1.6, font = 2)
mtext("CLR\ntransformed\ndata\n(no closure)",
      side = 3, line = -4, font = 2, adj = 0.98)
data0 <- na.omit(cities_clr[,c("Type","Compact","Open","Lightweight","Industry")])
points(pca_cities_open$x[,v1]*sf, pca_cities_open$x[,v2]*sf,
       pch = c(22,21,24,0,1)[data0$Type],
       lwd=2, bg = c(2,3,4,5,6)[data0$Type],
       col = c(2,3,8,9,11)[data0$Type],
       cex = c(1.2,1.4,1.2,1.2,1.4)[data0$Type])
dataEllipse(x=pca_cities_open$x[,v1]*sf, y=pca_cities_open$x[,v2]*sf*1.5,
            groups = data0$Type, add = TRUE, plot.points = FALSE, 
            levels = c(0.9), center.pch = 3, col = c(2,3,8,9,11),
            lty = 2, lwd = 1, center.cex = 2.4, group.labels = "")
legend("bottomright", bty = "o", inset = 0.03,
       box.col = "gray", box.lwd = 2, bg = 14,
       legend = levels(data0$Type),
       pch = c(22,21,24,0,1), pt.lwd = 2,
       col = c(2,3,8,9,11), pt.bg = c(2,3,4,5,6),
       pt.cex = c(1.2, 1.4, 1.2,1.2,1.4),
       cex = 0.9, y.intersp = 0.9)
mtext("(b)", side = 3, line = -1.5, font = 2, adj = 0.02, cex = 1.25)
```

```{r remove temp PCA objects, message=FALSE, warning=FALSE, include=FALSE, echo=FALSE}
rm(list = c("v1","v2","sf","data0"))
```

\normalsize

PCA biplots are useful, because the *variable weightings* group together for
variables (measurements) that are related to one another. For example, in the
biplots in \autoref{PCAplots}, the variables are relative land areas under 4
land-use categories (which have been corrected in \autoref{PCAplots}(b) for
compositional closure using the CLR transformation). These variables are shown
as vectors (arrows) in the biplot of principal components PC1 and PC2, and
variables which are related have vectors of similar length and/or direction. For
example, the variables 'Compact' and 'Industry' plot closely together on biplot
(a), suggesting some relationship between these land use categories. Removing
closure in our data (biplot (b)) seems to remove this relationship, so it is
probably not real!

The other main information we obtain from principal components biplots is from
the *observation scores* (see above). These will plot at locations similar to
their dominant variables: for example, in \autoref{PCAplots}(b), the 'Compact'
cities all plot towards the top of the biplot in the same direction as the
'Compact' variable weighting vector. This suggests that compact cities have
greater proportions of compact land use -- which should not be too surprising!
Note that in the biplot for closed data in \autoref{PCAplots}(a), the separation
of cities is not so clear.

**Some other things to try**

We could also group the observation scores in our biplot by a different factor
in our dataset. For example, it would be interesting to see if we could observe
any separation of cities by geographical region in the principal components
analysis. You might want to practice your **R** coding on this problem!

### References and R Packages

Fox, J. (2022). *RcmdrMisc: R Commander Miscellaneous Functions*. R package
version 2.7-2. <https://CRAN.R-project.org/package=RcmdrMisc>

John Fox and Sanford Weisberg (2019). *An {R} Companion to Applied Regression*
(**car**), Third Edition. Thousand Oaks CA: Sage. URL:
<https://socialsciences.mcmaster.ca/jfox/Books/Companion/>

Garrett, R.G. (2018). *rgr: Applied Geochemistry EDA*. R package version 1.1.15.
<https://CRAN.R-project.org/package=rgr>

Hu, J., Wang, Y., Taubenböck, H., Zhu, X.X. (2021). Land consumption in cities:
A comparative study across the globe. *Cities*, **113**: 103163,
<https://doi.org/10.1016/j.cities.2021.103163>.

Kassambara, A. and Mundt, F. (2020). *factoextra: Extract and Visualize the
Results of Multivariate Data Analyses*. R package version 1.0.7.
<https://CRAN.R-project.org/package=factoextra>

Maechler, M., Rousseeuw, P., Struyf, A., Hubert, M., Hornik, K.(2021). *cluster:
Cluster Analysis Basics and Extensions*. R package version 2.1.2.
<https://CRAN.R-project.org/package=cluster>

Reimann, C., Filzmoser, P., Garrett, R. G., & Dutter, R. (2008). *Statistical
Data Analysis Explained: Applied Environmental Statistics with R* (First ed.).
John Wiley & Sons, Chichester, UK.

Venables, W. N. & Ripley, B. D. (2002) *Modern Applied Statistics with S*
(**MASS**). Fourth Edition. Springer, New York. ISBN 0-387-95457-0.
<http://www.stats.ox.ac.uk/pub/MASS4/>

Wickham, H. (2019). *stringr: Simple, Consistent Wrappers for Common String
Operations*. R package version 1.4.0.
<https://CRAN.R-project.org/package=stringr>

Xu, N., Rate, A. W., & Morgan, B. (2018). From source to sink: Rare-earth
elements trace the legacy of sulfuric dredge spoils on estuarine sediments.
*Science of The Total Environment*, **637-638**, 1537-1549.
<https://doi.org/10.1016/j.scitotenv.2018.04.398>
