---
title: "testing multivariate analyses on landuse data"
output: 
  pdf_document: 
    fig_width: 5
    fig_height: 5
---

# load packages

```{r load packages etc., include=FALSE}
library(stringr)
library(RcmdrMisc)
library(rgr)
library(car)
library(cluster)
library(factoextra)
library(DataExplorer)
library(MASS)

palette(c("black","blue","green4","red2","purple",
          "darkcyan","firebrick","grey","grey40","white",
          "transparent"))
```

# load data

**Data are from**: https://stats.oecd.org/Index.aspx?DataSetCode=LAND_USE#

```{r eval=FALSE, include=FALSE}
# landuse_raw <- read.csv(file="landuse_OECD_2019.csv")
# landuse_raw$COU <- as.factor(landuse_raw$COU)
# landuse_raw$VAR <- as.factor(landuse_raw$VAR)
# str(landuse_raw)
```

# reformat data

this chunk: (i) makes each species a column (ii) replaces zeros with nominally low value to allow CLR transformation (iii) creates a version of data where percent land use areas are centered-log-ratio (CLR) transformed

```{r reformat data, eval=FALSE, include=FALSE}
# remove "eval=FALSE," to actually run this !

# make each species a column
# landuse <- data.frame(Country = levels(landuse_raw$COU), CountName = rep(NA,nlevels(landuse_raw$COU)))
# row.names(landuse) <- as.character(landuse$Country)
# UsesList <- levels(landuse_raw$VAR)

# for (i in 1:length(UsesList)){
#   vname <- UsesList[i]
#   landuse[,vname] <- rep(0, NROW(landuse))
#   rowz <- which(as.character(landuse_raw$VAR) == UsesList[i])
#   set <- landuse_raw[rowz, c("COU","Country", "Landuse")]
#   for (j in 1:NROW(set)){
#     landuse[as.character(set[j,1]),vname] <- set[j,3]
#     landuse[as.character(set[j,1]),"CountName"] <- set[j,2]
#   }
# }
# rm(list = c("i","j","vname","set"))

# for (i in 3:6){
#   landuse[which(landuse[,i] == 0),i] <- rep(0.01, 
#                                   length(which(landuse[,i] == 0)))
# }
# rm(i)
landuse <- read.csv(file = "landuse2.csv")
landuse$Region_coarse <- as.factor(landuse$Region_coarse)
landuse$Region <- as.factor(landuse$Region)
landuse$Climate <- as.factor(landuse$Climate)
landuse$Koppen <- as.factor(landuse$Koppen)
landuse$Koppen2 <- as.factor(landuse$Koppen2)
```

# make CLR-transformed dataset
```{r centered-logratio data}
require(rgr)
landuse_clr <- landuse
landuse_clr[,c("Ara","For","Mea","Oth")] <- 
  clr(landuse_clr[,c("Ara","For","Mea","Oth")])
names(landuse);names(landuse_clr)
```

# make logit-transformed dataset
```{r logit data, paged.print=FALSE}
require(car) # note rgr:: also had logit function!
landuse_logit <- landuse
landuse_logit[,c("Ara","For","Mea","Oth")] <- car::logit(landuse_logit[,c("Ara","For","Mea","Oth")]/100)
# landuse_logit[,c("Ara","For","Mea","Oth")] <- car::logit(landuse_logit[,5:8])
cat("\n")
head(landuse); cat("\n")
head(landuse_logit)
```

# correlation matrices
```{r correlation matrix untransformed, fig.height=4, fig.width=3.3}
require(DataExplorer)
plot_correlation(landuse[,c("Ara","For","Mea","Oth")],
                 cor_args = list("use" = "pairwise.complete.obs"))
```

# correlation matrices (CLR)
```{r correlation matrix clr, fig.height=4, fig.width=3.3}
require(DataExplorer)
plot_correlation(landuse_clr[,c("Ara","For","Mea","Oth")],
                 cor_args = list("use" = "pairwise.complete.obs"))
```

# correlation matrices (logit)
```{r correlation matrix clr, fig.height=4, fig.width=3.3}
require(DataExplorer)
plot_correlation(landuse_logit[,c("Ara","For","Mea","Oth")],
                 cor_args = list("use" = "pairwise.complete.obs"))
```

# principal components analysis

```{r run and summarise PCA}
data0 <- na.omit(landuse[,c("Ara","For","Mea","Oth")])
pca_landuse_clos <- prcomp(data0, scale. = TRUE)
pca_landuse_clos$rot
cat("...\n\nComponent Variances\n")
pca_landuse_clos$sdev^2
cat("\n--------------------\n")

data0 <- na.omit(landuse_clr[,c("Ara","For","Mea","Oth")])
pca_landuse_open <- prcomp(data0, scale. = TRUE)
pca_landuse_open$rot
cat("...\n\nComponent Variances\n")
pca_landuse_open$sdev^2
cat("\n--------------------\n")
rm(data0)
```

```{r visualise PCA, fig.height=10, fig.width=20}
require(car)
palette(c("black","blue","green4","red2","purple",
          "darkcyan","firebrick","grey","grey40","white",
          "transparent"))
par(mfrow=c(1,2), mar = c(3.5,3.5,3.5,3.5), oma = c(0,0,0,0), 
    mgp=c(1.7,0.3,0), tcl = 0.25, font.lab=2,
    lend = "square", ljoin = "mitre")
# choose components and set scaling factor (sf)
v1 <- 1
v2 <- 2
sf <- 0.8

biplot(pca_landuse_clos, choices = c(v1,v2), col = c(11,9), cex=c(1,0.65),
       pc.biplot = FALSE, scale = 0.4, arrow.len = 0.08,
       xlab = paste0("Scaled PC",v1," Component Loadings"),
       ylab = paste0("Scaled PC",v2," Component Loadings"))
mtext(paste0("Scaled PC",v1," Observation Scores"), 3, 1.6, font = 2)
mtext(paste0("Scaled PC",v2," Observation Scores"), 4, 1.6, font = 2)
mtext("Untransformed data\n(compositionally closed)",
      side = 3, line = -2, font = 2, adj = 0.98)
text(pca_landuse_clos$x[,v1]*sf, pca_landuse_open$x[,v2]*sf, labels=landuse$Koppen2, 
     col=rainbow(5,v=0.7)[landuse$Koppen2], cex = 0.8)

sf <- 0.8
biplot(pca_landuse_open, choices = c(v1,v2), col = c(11,9), cex=c(1,0.65),
       pc.biplot = FALSE, scale = 0.5, arrow.len = 0.08, 
       xlab = paste0("Scaled PC",v1," Component Loadings"),
       ylab = paste0("Scaled PC",v2," Component Loadings"))
text(pca_landuse_open$x[,v1]*sf, pca_landuse_open$x[,v2]*sf, labels=landuse_clr$Koppen2, 
     col=rainbow(5,v=0.7)[landuse_clr$Koppen2], cex = 0.8)
mtext(paste0("Scaled PC",v1," Observation Scores"), 3, 1.6, font = 2)
mtext(paste0("Scaled PC",v2," Observation Scores"), 4, 1.6, font = 2)
mtext("CLR\ntransformed\ndata\n(no closure)",
      side = 3, line = -4, font = 2, adj = 0.02)
rm(list = c("v1","v2","sf"))
```

# K-means clustering

```{r assess optimum clusters for closed data, fig.height=3, fig.width=4}
require(factoextra)
# data0 <- na.omit(landuse[,c("Ara","For","Mea","Oth")])
fviz_nbclust(scale(landuse[,c("Ara","For","Mea","Oth")]), kmeans, method = "wss") +
  geom_vline(xintercept = 4, linetype = 2)
```

```{r assess optimum clusters for CLR data, fig.height=3, fig.width=4}
require(factoextra)
# data0 <- na.omit(landuse_clr[,c("Ara","For","Mea","Oth")])
fviz_nbclust(scale(landuse_clr[,c("Ara","For","Mea","Oth")]), kmeans, method = "wss") +
  geom_vline(xintercept = 4, linetype = 2)
```

# compute K-means closed

```{r compute K-means closed}
# data0 <- na.omit(landuse[,c("Ara","For","Mea","Oth")])
set.seed(123)
landuse_clos_kmeans <- kmeans(scale(landuse[,c("Ara","For","Mea","Oth")]), 5, nstart = 25)
cat("K-means clustering with",length(landuse_clos_kmeans$size),"clusters of sizes",landuse_clos_kmeans$size,"\n")
cat("\ncomponents of output object are:\n")
ls(landuse_clos_kmeans)
kc_rslt <- as.data.frame(cbind(as.character(landuse$Koppen2), landuse_clos_kmeans$cluster))
colnames(kc_rslt) <- c("Koppen","Cluster")
output <- vector(mode = "list", length = max(kc_rslt$Cluster))
for(i in 1:max(kc_rslt$Cluster)) {
  output[[i]] <- subset(kc_rslt, subset = kc_rslt$Cluster == i)
  cat("\n======== Cluster",i,"========\n")
  print(table(as.factor(output[[i]]$Koppen)))
}
rm(list=c("kc_rslt","output"))
```

# plot kmeans clusters closed

```{r visualize kmeans clusters closed, fig.height=9, fig.width=9, message=FALSE, warning=FALSE}
data0 <- na.omit(landuse[,c("Koppen2","Ara","For","Mea","Oth")])
data0[,c("Ara","For","Mea","Oth")] <- scale(data0[,c("Ara","For","Mea","Oth")])
row.names(data0) <- paste0(data0$Koppen2, seq(1, nrow(data0)))
fviz_cluster(landuse_clos_kmeans, data = data0[,c("Ara","For","Mea","Oth")],
             palette = c("#00AFBB", "#E7B800", "#FC4E07","purple2","grey60"),
             labelsize=9,
             ellipse.type = "euclid", # Concentration ellipse
             star.plot = TRUE, # Add segments from centroids to items
             repel = F, # if true avoids label overplotting (slow)
             ggtheme = theme_minimal()
             )
```

# compute K-means open

```{r kmeans open}
data0 <- scale(na.omit(landuse_clr[,c("Ara","For","Mea","Oth")]))
# row.names(data0) <- landuse_clr$CountName
set.seed(42)
landuse_open_kmeans <- kmeans(data0[,c("Ara","For","Mea","Oth")], 5, nstart = 25)
cat("K-means clustering with",length(landuse_open_kmeans$size),"clusters of sizes",landuse_open_kmeans$size,"\n")
cat("\ncomponents of output object are:\n")
ls(landuse_open_kmeans)
kc_rslt <- as.data.frame(cbind(as.character(landuse_clr$Koppen2), landuse_open_kmeans$cluster))
colnames(kc_rslt) <- c("Koppen","Cluster")
output <- vector(mode = "list", length = max(kc_rslt$Cluster))
for(i in 1:max(kc_rslt$Cluster)) {
  output[[i]] <- subset(kc_rslt, subset = kc_rslt$Cluster == i)
  cat("\n======== Cluster",i,"========\n")
  print(table(as.factor(output[[i]]$Koppen)))
}
rm(list=c("kc_rslt","output"))
```

# plot kmeans clusters open

```{r visualize kmeans clusters open, fig.height=10, fig.width=9, message=FALSE, warning=FALSE}
row.names(data0) <- paste0(landuse$Koppen2, seq(1,nrow(data0)))
fviz_cluster(landuse_open_kmeans, data = data0[,c("Ara","For","Mea","Oth")],
             palette = c("#00AFBB", "#E7B800", "#FC4E07","purple2","grey60"),
             labelsize=9,
             ellipse.type = "euclid", # Concentration ellipse
             star.plot = TRUE, # Add segments from centroids to items
             repel = F, # if true avoids label overplotting (slow)
             ggtheme = theme_minimal()
             )
```

# compute K-means logit

```{r kmeans logit}
data0 <- scale(na.omit(landuse_logit[,c("Ara","For","Mea","Oth")]))
row.names(data0) <- paste0(landuse_logit$Koppen,seq(1,nrow(data0)))
set.seed(42)
landuse_logit_kmeans <- kmeans(data0[,c("Ara","For","Mea","Oth")], 5, nstart = 25)
cat("K-means clustering with",length(landuse_logit_kmeans$size),"clusters of sizes",landuse_logit_kmeans$size,"\n")
cat("\ncomponents of output object are:\n")
ls(landuse_logit_kmeans)
kc_rslt <- as.data.frame(cbind(as.character(landuse$Koppen), landuse_logit_kmeans$cluster))
colnames(kc_rslt) <- c("Koppen","Cluster")
output <- vector(mode = "list", length = max(kc_rslt$Cluster))
for(i in 1:max(kc_rslt$Cluster)) {
  output[[i]] <- subset(kc_rslt, subset = kc_rslt$Cluster == i)
  cat("\n======== Cluster",i,"========\n")
  print(table(as.factor(output[[i]]$Koppen)))
}
rm(list=c("kc_rslt","output"))
```

# plot kmeans clusters logit

```{r visualize kmeans clusters logit, fig.height=8, fig.width=9.5, message=FALSE, warning=FALSE}
row.names(data0) <- paste0(landuse_logit$Koppen,seq(1,nrow(data0)))
fviz_cluster(landuse_open_kmeans, data = data0[,c("Ara","For","Mea","Oth")],
             palette = c("#00AFBB", "#E7B800", "#FC4E07","purple2","grey60"),
             labelsize=9,
             ellipse.type = "euclid", # Concentration ellipse
             star.plot = TRUE, # Add segments from centroids to items
             repel = F, # if true avoids label overplotting (slow)
             ggtheme = theme_minimal()
             )
```

# hierarchical clustering

## create dissimilarity (distance) matrix

```{r distance matrix closed, message=FALSE, warning=FALSE}
dataHC <- na.omit(landuse[,c("Koppen2","Ara","For","Mea","Oth")])
row.names(dataHC) <- paste0(as.character(dataHC$Koppen2),seq(1,nrow(dataHC)))
dataHC$Koppen2 <- NULL
landuse_clos_diss <- get_dist(dataHC, method = "euclidean")
as.matrix(landuse_clos_diss)[1:8, 1:8]
```

## perform hierarchical clustering

```{r hierarchical clustering closed, fig.height=4.5, fig.width=15, message=FALSE, warning=FALSE}
landuse_clos_hc <- hclust(landuse_clos_diss, method = "average")
require(factoextra)
fviz_dend(landuse_clos_hc, cex = 0.5)
```

## assess cluster tree

```{r verify cluster tree closed, results="hold"}
landuse_clos_coph <- cophenetic(landuse_clos_hc)
cat("Correlation r =",cor(landuse_clos_diss,landuse_clos_coph), "when hclust method is",landuse_clos_hc$method)
cat("\nRule-of-thumb: cluster tree represents actual distance matrix accurately enough if r>0.75\n")
```

## cut dendrogram into different groups

```{r cut dendrogram closed, paged.print=FALSE}
landuse_clos_grp <- cutree(landuse_clos_hc, k = 5)
results <- data.frame(Country = names(landuse_clos_grp),
                      Cluster = as.numeric(landuse_clos_grp))
results <- results[order(results$Cluster, results$Country),] 
output <- vector(mode = "list", length = max(results$Cluster))
for(i in 1:max(results$Cluster)) {
  output[[i]] <- subset(results, subset = results$Cluster==i)
  cat("\nCluster",i,"\n")
  print(output[[i]]$Country)
}
```

## plot dendrogram with cuts

```{r plot cut dendrogram closed, fig.height=4.5, fig.width=12, message=FALSE, warning=FALSE}
fviz_dend(landuse_clos_hc, k = 5, # Cut in five groups
          cex = 0.5, # label size
          k_colors = c("red3", "blue2", "purple","sienna","grey40"),
          color_labels_by_k = TRUE, # color labels by groups
          rect = TRUE # Add rectangle around groups
          )
```

# hierarchical clustering (CLR-open data)

## create dissimilarity (distance) matrix

```{r distance matrix open, message=FALSE, warning=FALSE}
dataHC <- na.omit(landuse_clr[,c("Koppen2","Ara","For","Mea","Oth")])
row.names(dataHC) <- paste0(as.character(dataHC$Koppen2),seq(1,nrow(dataHC)))
dataHC$Koppen2 <- NULL
landuse_open_diss <- get_dist(dataHC, method = "euclidean")
as.matrix(landuse_open_diss)[1:8, 1:8]
```

## perform hierarchical clustering

```{r hierarchical clustering open, fig.height=4.5, fig.width=12, message=FALSE, warning=FALSE}
landuse_open_hc <- hclust(landuse_open_diss, method = "average")
require(factoextra)
fviz_dend(landuse_open_hc, cex = 0.5)
```

## assess cluster tree

```{r verify cluster tree open}
landuse_open_coph <- cophenetic(landuse_open_hc)
cat("Correlation r =",cor(landuse_open_diss,landuse_open_coph), "when hclust method is",landuse_open_hc$method)
cat("\nRule-of-thumb: cluster tree represents actual distance matrix accurately enough if r>0.75\n")
```

## cut dendrogram into different groups

```{r cut dendrogram open}
landuse_open_grp <- cutree(landuse_open_hc, k = 5)
results <- data.frame(Country = names(landuse_open_grp),
                      Cluster = as.numeric(landuse_open_grp))
results <- results[order(results$Cluster, results$Country),] 
output <- vector(mode = "list", length = max(results$Cluster))
for(i in 1:max(results$Cluster)) {
  output[[i]] <- subset(results, subset = results$Cluster==i)
  cat("\nCluster",i,"\n")
  print(output[[i]]$Country)
}
table(landuse_open_grp)
```

## plot dendrogram with cuts

```{r plot cut dendrogram open, fig.height=4.5, fig.width=12, message=FALSE, warning=FALSE}
fviz_dend(landuse_open_hc, k = 5, # Cut into 5 groups
          cex = 0.5, # label size
          k_colors = c("black","red3", "blue2", "purple","darkcyan"),
          color_labels_by_k = TRUE, # color labels by groups
          rect = TRUE # Add rectangle around groups
)
```

# LDA linear discriminant analysis

```{r LDA closed cities data}
data0 <- landuse
data0[,c("Ara","For","Mea","Oth")] <- scale(data0[,c("Ara","For","Mea","Oth")])
lda_landuse_clos <- lda(formula = Koppen2 ~ Ara + For + Mea + Oth, 
                    data=data0,
                    prior=as.numeric(summary(landuse$Koppen2))/
                      nrow(landuse)) 
print(lda_landuse_clos)
```

```{r LDA open cities data}
data0 <- landuse_clr
data0[,c("Ara","For","Mea","Oth")] <- scale(data0[,c("Ara","For","Mea","Oth")])
lda_landuse_open <- lda(formula = Koppen2 ~ Ara + For + Mea + Oth, 
                    data=data0,
                    prior=as.numeric(summary(landuse$Koppen2))/
                      nrow(landuse)) 
print(lda_landuse_open)
```

```{r plot LDA closed, fig.height=5, fig.width=10}
par(mfrow = c(1,2), mar = c(3.5,3.5,1,1), mgp = c(1.5,0.3,0), tcl = 0.25,
    lend = "square", ljoin = "mitre", cex.main = 0.9, font.lab=2)
plot(lda_landuse_clos$scaling[,1], lda_landuse_clos$scaling[,2], 
     xlim = c(-0.5,12), ylim=c(-14,0.5),
     xlab="Linear Discriminant [1]", ylab="Linear Discriminant [2]", main="(a) Variable Coefficients")
abline(v=0,col="grey",lty=2)
abline(h=0,col="grey",lty=2)
text(lda_landuse_clos$scaling[,1],lda_landuse_clos$scaling[,2],
     labels=c("Compact", "Open", "Lightweight","Industry"),
     pos=1,cex=0.95,col="blue2",offset=0.2)

ldaPred_landuse_clos <- predict(lda_landuse_clos)
for(i in 1:NROW(lda_landuse_clos$scaling)){
  arrows(0,0,lda_landuse_clos$scaling[i,1],lda_landuse_clos$scaling[i,2],
         length = 0.15, col = 8)
}
plot(ldaPred_landuse_clos$x[,1], ldaPred_landuse_clos$x[,2], 
     col=c(2,"#00666680",8,"#99660080","darkred","tomato")[landuse$Koppen2], 
     pch=c(0,15,1,19,2,3)[landuse$Koppen2], 
     ylim = c(-3,3),
  lwd=c(2,1,2,1,2,2)[landuse$Koppen2],
  cex=c(1.8,1.8,2,2,2)[landuse$Koppen2], 
  xlab="Linear Discriminant [1]", ylab="Linear Discriminant [2]", 
  main="(b) Predictions for Observations")
abline(v=0,col="grey",lty=2)
abline(h=0,col="grey",lty=2)
text(ldaPred_landuse_clos$x[,1], ldaPred_landuse_clos$x[,2], 
     labels=landuse$Koppen2, col=c(2,"blue4","#991100","#663300",8)[landuse$Koppen2],
     pos=1, offset=0.15, cex=0.65)
legend("bottomleft",legend=levels(landuse$Koppen2),
       col=c(2,"#00666680",8,"#99660080","darkred","tomato"),
       pch=c(0,15,1,19,2,3), pt.lwd=c(2,1,2,1,2,2),
       title="City Koppen",bty="o", box.col="grey90", 
       box.lwd=2, inset=0.02, 
       pt.cex=c(1.8,1.8,2,2,2), cex=0.85)

```

```{r compare models with reality, paged.print=FALSE}
ldaPred_landuse_clos <- predict(lda_landuse_clos)
closComp <- as.data.frame(cbind(as.character(landuse$Koppen2), 
                                as.character(ldaPred_landuse_clos$class)))
colnames(closComp) <- c("=-Actual-=","=-Predicted-=")
closComp$test <- as.character(landuse$Koppen2) == as.character(ldaPred_landuse_clos$class)
k = length(which(closComp$test == TRUE))
cat("Predictions by LDA using closed data:",k,"out of",NROW(landuse_clr),"=",paste0(signif(100*k/NROW(landuse),4),"% correct\n"))
closComp

ldaPred_landuse_open <- predict(lda_landuse_open)
openComp <- as.data.frame(cbind(as.character(landuse_clr$Koppen2), as.character(ldaPred_landuse_open$class)))
colnames(openComp) <- c("=-Actual-=","=-Predicted-=")
openComp$test <- as.character(landuse_clr$Koppen2) == as.character(ldaPred_landuse_open$class)
k = length(which(openComp$test == TRUE))
cat("\nPredictions by LDA using open data:",k,"out of",NROW(landuse_clr),"=",paste0(signif(100*k/NROW(landuse_clr),4),"% correct\n"))
openComp
```

```{r LDA train and predict closed, message=FALSE, warning=FALSE, paged.print=FALSE}
library(MASS)
library(RcmdrMisc)
library(knitr)
library(beepr)
### MATCH this to read file below
# sink(file="landuse_type_pred.csv", type="output")
n0 <- 100 # number of iterations
ftrain <- 0.75 # proportion of observations in training set
results <- data.frame(
    Rep = rep(NA, n0),
    matches = rep(NA, n0),
    non_matches = rep(NA, n0),
    success = rep(NA, n0))
# train <- sample(1:NROW(landuse), round(NROW(landuse)-5,0))
train <- sample(1:NROW(landuse), round(NROW(landuse) * ftrain,0))
lda.landuse.train <- lda(formula = Koppen2 ~  Ara + For + Mea + Oth, 
                    data = landuse[train,],
                    prior = as.numeric(summary(landuse$Koppen2[train]))/
                      nrow(landuse[train,])) 
lda.landuse.pred <- predict(lda.landuse.train, landuse[-train,])
# make vector of individual category non-matches
matchByClass <- data.frame(Matches = rep(0,nlevels(landuse$Koppen2))) 
rownames(matchByClass) <- levels(lda.landuse.pred$class)
e0 <- 0
# colnames(matchByClass) <- c("Matches")
for (i in 1:n0) {
  # train <- sample(1:NROW(landuse), round(NROW(landuse)-5,0))
  train <- sample(1:NROW(landuse), round(NROW(landuse) * ftrain,0))
  if (is.na(match(NA,tapply(landuse[train,]$Ara, landuse[train,]$Koppen2, sd, na.rm=T))) == TRUE) {
    lda.landuse.train <- lda(formula = Koppen2 ~ Ara + For + Mea + Oth, 
                          data = landuse[train,],
                          prior=as.numeric(summary(landuse$Koppen2[train]))/
                        nrow(landuse[train,])) 
    lda.landuse.pred <- predict(lda.landuse.train, landuse[-train,])
    e0 <- e0 + 1
  }

  k=0 # number of non-matches
  m0 <- as.matrix(rep(0,5)) # vector of individual category non-matches
  rownames(m0) <- levels(lda.landuse.pred$class)
  rownames(matchByClass) <- levels(lda.landuse.pred$class)
  colnames(matchByClass) <- c("Matches")
  for (jM in 1:NROW(landuse[-train,])) {
    # cat("medium loop #",jM,"\n")
    for (jS in 1:nlevels(lda.landuse.pred$class)) {
      if((lda.landuse.pred$class[jM] == levels(lda.landuse.pred$class)[jS]) & 
         (landuse$Koppen2[-train][jM] == levels(lda.landuse.pred$class)[jS]) ) 
        m0[jS] = m0[jS] + 1
      else  m0[jS] = m0[jS] 
      # cat("small loop iteration #",jS,"; matching",
      #     levels(lda.landuse.pred$class)[jS],
      #     "; matches =",m0,"\n")
    }
    k = sum(m0)
    # cat("medium loop iteration ",jM,"; matches = ",k,"\n", sep="")
    # if(jM==NROW(landuse[-train,])) 
  }
  # cat("GIANT LOOP #",i,"\n")
  matchByClass <- matchByClass + m0
  # output to results data frame: iteration, matches, non-matches, proportion matched
  results[i,] <- c(i, k, NROW(landuse[-train,])-k, signif(k/NROW(landuse[-train,]),3))
  # cbind(lda.landuse.pred$class,landuse$Koppen2[-train])
}
matchByClass$Actual <- round(1000*as.numeric(summary(landuse$Koppen2))*
                               (NROW(landuse[-train,])/NROW(landuse)),0)
matchByClass$Proportion <- matchByClass$Matches/matchByClass$Actual

# sink() # close output file
### make sure read file is SAME AS SINK
beep(sound = 10)

# results <- read.csv("landuse_type_pred.csv")
{cat("[Based on", n0, "random subsets of dataset to",
    "train LDA model to\npredict remaining observations]\n")
cat("Number of obs. in random subsets =",NROW(train),
    " (predicting",NROW(landuse)-NROW(train),"samples)\n")
print(numSummary(results[,2:4], statistics=c("mean","sd"))$table)
ns0 <- numSummary(results$success)
t0 <- t.test(results$success)
cat(rep("-\u2013-",24),
    "\nStat. summary for 'success':\nMean = ",round(ns0$table[1],4),
    ", sd = ",round(ns0$table[2],4),", 95% confidence interval = (",
    signif(t0$conf.int[1],3),", ",signif(t0$conf.int[2],4),") (after ",i," reps)\n", sep="")
cat(n0-e0,"iterations failed due to random sampling missing a group\n\n")
print(matchByClass, digits=3)}
rm(list = c("n0","ftrain","i","e0","jS","jM","k","m0","matchByClass","results","train"))
```

