---
title: "testing cluster analysis"
output: 
  pdf_document: 
    fig_width: 5
    fig_height: 5
---

**Reference**: Kassambara, Alboukadel (2017) *Practical Guide To Cluster Analysis in R* https://xsliulab.github.io/Workshop/week10/r-cluster-book.pdf

# load packages

```{r load packages etc., include=FALSE}
library(stringr)
library(RcmdrMisc)
library(rgr)
library(car)
library(cluster)
library(factoextra)
library(DataExplorer)

palette(c("black","blue","green4","red2","purple",
          "darkcyan","firebrick","grey","grey40","white",
          "transparent"))
```

# load data

**Data are from**: Matonis, Megan; Binkley, Daniel (2018), “Data for: Not just about the trees: Key role of mosaic-meadows in restoration of ponderosa pine ecosystems”, *Mendeley Data*, V1, https://doi.org/10.17632/4yf6ck7hjx.1

**based on this article**: Matonis, M.S., Binkley, D., 2018. Not just about the trees: Key role of mosaic-meadows in restoration of ponderosa pine ecosystems. *Forest Ecology and Management*, **411**: 120-131, https://doi.org/10.1016/j.foreco.2018.01.019

**Sites are** Heil Valley Ranch (HL), Long John (LJ), Messenger Gulch (MG), Phantom Creek (PC), Manitou Experimental Forest (ME), UncMesas (UM).

```{r load data, eval=FALSE, include=FALSE}
# remove "eval=FALSE," to actually run this !
meadow <- read.csv(file="MatonisBinkley_meadow.csv")
meadow$QuadratID <- as.factor(meadow$QuadratID)
meadow$Life.form <- as.factor(meadow$Life.form)
names(meadow)[4] <- "Cover_pct"
with(meadow, table(QuadratID))
meadow$meadow <- as.factor(str_sub(meadow$QuadratID,4,5))
meadow$rep <- as.factor(str_trunc(as.character(meadow$QuadratID),
                                  2,side="left",ellipsis = ""))

meadow$SiteID <- as.factor(str_trunc(as.character(meadow$QuadratID),
                                     2, side="right", ellipsis = ""))
str(meadow)
```

# reformat data

this chunk: (i) makes each species a column (ii) replaces zeros with nominally low value to allow CLR transformation (iii) creates a version of data where percent species abundances are centered-log-ratio (CLR) transformed

```{r reformat data, eval=FALSE, include=FALSE}
# remove "eval=FALSE," to actually run this !

# make each species a column
matonis <- data.frame(SiteMeadQuad = levels(meadow$QuadratID))
row.names(matonis) <- as.character(matonis$SiteMeadQuad)
meadow$SpeciesFac <- as.factor(meadow$Species)
speclist <- levels(meadow$SpeciesFac)

for (i in 1:length(speclist)){
  vname <- str_replace_all(speclist[i], " ", "_")
  matonis[,vname] <- rep(0, NROW(matonis))
  rowz <- which(as.character(meadow$Species) == speclist[i])
  set <- meadow[rowz, c("QuadratID", "Cover_pct")]
  for (j in 1:NROW(set)){
    matonis[as.character(set[j,1]),vname] <- set[j,2]
  }
}
rm(list = c("i","j","vname","set"))
# calc bare soil by difference and replace zeros
matonis$Nothing <- NULL
matonis$Bare_soil <- 100-rowSums(matonis[,2:150])
negz <- which(matonis$Bare_soil < 0)
matonis$Bare_soil[negz] <- rep(0.01, length(negz)); rm(negz)
matonis$SumSpp <- rowSums(matonis[,2:151])

for (i in 2:151){
  matonis[which(matonis[,i] == 0),i] <- rep(0.001, 
                                  length(which(matonis[,i] == 0)))
}

matonis$site <- as.factor(str_trunc(matonis$SiteMeadQuad, 2, 
                                    side="right", ellipsis = ""))
matonis$meadow <- as.factor(str_sub(matonis$SiteMeadQuad, 4, 5))
matonis$rep <- as.factor(str_trunc(matonis$SiteMeadQuad, 2, 
                                   side="left", ellipsis = ""))
# CLR-transform percent cover
matonCLR <- matonis
matonCLR[,2:150] <- clr(matonCLR[,2:150])
```

# principal components analysis
```{r}

```

```{r run and summarise PCA}
pca_closed <- prcomp(matonis[,2:151], scale. = TRUE)
pca_closed$rot[1:20,1:5]
cat("...\n\nComponent Variances\n")
pca_closed$sdev[1:10]^2
cat("\n--------------------\n")

pca_open <- prcomp(matonCLR[,2:151], scale. = TRUE)
pca_open$rot[1:20,1:5]
cat("...\n\nComponent Variances\n")
pca_open$sdev[1:10]^2
cat("\n--------------------\n")
```

```{r visualise PCA, fig.height=6, fig.width=12}
require(car)
palette(c("black","blue","green4","red2","purple",
          "darkcyan","firebrick","grey","grey40","white",
          "transparent"))
par(mfrow=c(1,2), mar = c(3.5,3.5,3.5,3.5), oma = c(0,0,0,0), 
    mgp=c(1.7,0.3,0), tcl = 0.25, font.lab=2,
    lend = "square", ljoin = "mitre")
# choose components and set scaling factor (sf)
v1 <- 1
v2 <- 3
sf <- .1
# truncate species names to first 3 letters of Genus & species
specnames <- names(matonis[,2:151])
specshort <- paste0(str_trunc(specnames,3,"right",""),
                    str_to_title(str_sub(specnames,
                        str_locate(specnames,"_")[,1]+1,
                        str_locate(specnames,"_")[,1]+3)))
specshort[62] <- "Forb"
specshort[150] <- "Soil"

biplot(pca_closed, choices = c(v1,v2), col = c(11,9), cex=c(1,0.65),
       pc.biplot = FALSE, scale = 0.4, arrow.len = 0.08,
       xlim = c(-1.5,1.5), ylim = c(-1.2,1.2), 
       xlab = paste0("Scaled PC",v1," Component Loadings"),
       ylab = paste0("Scaled PC",v2," Component Loadings"),
       ylabs = specshort)
mtext(paste0("Scaled PC",v1," Observation Scores"), 3, 1.6, font = 2)
mtext(paste0("Scaled PC",v2," Observation Scores"), 4, 1.6, font = 2)
mtext("Untransformed data\n(compositionally closed)",
      side = 3, line = -2, font = 2)
# points(pca_closed$x[,v1]*sf, pca_closed$x[,v2]*sf*1.5,
#        pch = c(22,21,24,1,0,2)[matonis$site],
#        bg = c(2,3,4,5,6,7)[matonis$site],
#        col = c(2,3,4,5,6,7)[matonis$site],
#        cex = c(1.2,1.4,1.2,1.4,1.2,1.2)[matonis$site],
#        lwd = c(1,1,1,2,2,2)[matonis$site])
# dataEllipse(x=pca_closed$x[,v1]*sf, y=pca_closed$x[,v2]*sf*1.5, 
#             groups = matonis$site, add = TRUE,
#             plot.points = FALSE, levels = c(0.95),
#             center.pch = 3, col = c(2,3,4,5,6,7), 
#             lty = 2, lwd = 1, center.cex = 2.4)
# symbols(0,0, circles = 0.5, fg = "grey", 
#         add = TRUE, inches = FALSE)
# legend("bottomright", bty = "o", inset = 0.03,
#        title = "Site",
#        box.col = "gray", box.lwd = 2, bg = 10,
#        legend = levels(matonis$site),
#        pch = c(22,21,24,1,0,2),
#        col = c(2,3,4,5,6,7), pt.bg = c(2,3,4,5,6,7),
#        pt.cex = c(1.2, 1.4, 1.2,1.4,1.2,1.2),
#        pt.lwd = c(1,1,1,2,2,2),
#        cex = 1.1, y.intersp = 0.9)

# v1 <- 1
# v2 <- 3
# sf <- .1
specnames <- names(matonCLR[,2:151])
specshort <- paste0(str_trunc(specnames,3,"right",""),
                    str_to_title(str_sub(specnames,
                        str_locate(specnames,"_")[,1]+1,
                        str_locate(specnames,"_")[,1]+3)))
specshort[62] <- "Forb"
specshort[150] <- "Soil"

biplot(pca_open, choices = c(v1,v2), col = c(11,9), cex=c(1,0.65),
       pc.biplot = FALSE, scale = 0.5, arrow.len = 0.08,
             xlim = c(-1.25,1.25), ylim = c(-0.75,1.75), 
xlab = paste0("Scaled PC",v1," Component Loadings"),
       ylab = paste0("Scaled PC",v2," Component Loadings"),
       ylabs = specshort)
mtext(paste0("Scaled PC",v1," Observation Scores"), 3, 1.6, font = 2)
mtext(paste0("Scaled PC",v2," Observation Scores"), 4, 1.6, font = 2)
mtext("Centred log ratio transformed data\n(no compositional closure)",
      side = 3, line = -2, font = 2)
# points(pca_open$x[,v1]*sf, pca_open$x[,v2]*sf*1.5,
#        pch = c(22,21,24,1,0,2)[matonCLR$site],
#        bg = c(2,3,4,5,6,7)[matonCLR$site],
#        col = c(2,3,4,5,6,7)[matonCLR$site],
#        cex = c(1.2,1.4,1.2,1.4,1.2,1.2)[matonCLR$site],
#        lwd = c(1,1,1,2,2,2)[matonCLR$site])
# dataEllipse(x=pca_open$x[,v1]*sf, y=pca_open$x[,v2]*sf*1.5, 
#             groups = matonCLR$site, add = TRUE,
#             plot.points = FALSE, levels = c(0.95),
#             center.pch = 3, col = c(2,3,4,5,6,7), 
#             lty = 2, lwd = 1, center.cex = 2.4)
# symbols(0,0, circles = 0.5, fg = "grey", 
#         add = TRUE, inches = FALSE)
# legend("bottomright", bty = "o", inset = 0.03,
#        title = "Site",
#        box.col = "gray", box.lwd = 2, bg = 10,
#        legend = levels(matonCLR$site),
#        pch = c(22,21,24,1,0,2),
#        col = c(2,3,4,5,6,7), pt.bg = c(2,3,4,5,6,7),
#        pt.cex = c(1.2, 1.4, 1.2,1.4,1.2,1.2),
#        pt.lwd = c(1,1,1,2,2,2),
#        cex = 1.1, y.intersp = 0.9)

```

# CLUSTER ANALYSIS

## hierarchical clustering

```{r fig.height=10, fig.width=8, message=TRUE, warning=FALSE}
clusClos <- droplevels(matonis[1:50,])
row.names(clusClos) <- clusClos$SiteMeadQuad
# make similarity matrix
clusClos_sim <- dist(clusClos, method = "euclidean")
# use similarity matrix to calculate clusters
maton_clus_clos <- hclust(clusClos_sim, method = "ward.D")
par(mfrow=c(3,1), mar = c(5,3.5,1,1), mgp = c(1.7,0.3,0), tcl = 0.25, font.lab = 2)
plot(maton_clus_clos, ylab = "Euclidean Distance",
     cex=0.75, col = c(1,2,3))
maton_den_clos <- as.dendrogram(maton_clus_clos)
plot(maton_den_clos, ylab = "Euclidean Distance")
plot(cut(maton_den_clos, h=50)$lower[[2]], ylab = "Euclidean Distance", 
     main="just cluster 2 at cutoff=50")
par(mfrow=c(1,1), mar = c(3.5,3.5,1,1))
```

```{r fig.height=8, fig.width=9, message=FALSE, warning=FALSE}
clusOpen <- droplevels(matonCLR[1:50,])
row.names(clusOpen) <- clusOpen$SiteMeadQuad
clusOpen_sim <- dist(clusOpen, method = "euclidean")
maton_clus_open <- hclust(clusOpen_sim, method = "ward.D")
# par(mfrow=c(1,1), mar = c(3.5,3.5,1,1))
# plot(maton_clus_open, ylab = "Euclidean Distance",
#      cex=0.5, col = c(1,2,3))
maton_den_open <- as.dendrogram(maton_clus_open)
par(mfrow=c(2,1), mar = c(5,3,1,1), mgp = c(1.7,0.3,0), tcl = 0.25, font.lab = 2)
plot(maton_den_open, ylab = "Euclidean Distance", 
     main="Complete Dendrogram", cex.main = 1)
kut = 42
abline(h=kut, col = "grey", lty = 2)
plot(cut(maton_den_open, h=kut)$upper, ylab = "Euclidean Distance", 
     main=paste0("Dendrogram above cutoff=",kut," (for instance)"), cex.main = 1)
par(mfrow=c(1,1), mar = c(3.5,3.5,1,1)); rm(kut)
```

```{r fig.height=8, fig.width=12}
# compare closed and open ####
par(mfrow=c(2,1), mar = c(4.5,3.5,1,1))
plot(maton_den_clos, ylab = "Euclidean Distance")
mtext("Closed data", side = 3, line = -1.5, adj = 0.95, font=2, col = 2)
plot(maton_den_open, ylab = "Euclidean Distance")
mtext("Open (CLR) data", side = 3, line = -1.5, adj = 0.95, font=2, col = 5)
par(mfrow=c(1,1))
```

# K-means clustering

using Heil Valley Ranch "HL" subset of data from Matonis, Megan; Binkley, Daniel (2018), “Data for: Not just about the trees: Key role of mosaic-meadows in restoration of ponderosa pine ecosystems”, *Mendeley Data*, V1, https://doi.org/10.17632/4yf6ck7hjx.1

```{r read and prepare data, include=FALSE}
MatonHL_clos <- read.csv(file="clusClos.csv", stringsAsFactors = TRUE)
row.names(MatonHL_clos) <- MatonHL_clos$MeadQuad
MatonHL_clos[,2:49] <- scale(MatonHL_clos[,2:49])

# use CLR transform to open data...
require(rgr)
MatonHL_open <- read.csv(file="clusClos.csv", stringsAsFactors = TRUE)
row.names(MatonHL_open) <- MatonHL_open$MeadQuad
MatonHL_open[,2:49] <- clr(MatonHL_open[,2:49])
# ...and scale the transformed variables
MatonHL_open[,2:49] <- scale(MatonHL_open[,2:49])
```

```{r assess optimum clusters for closed data}
require(factoextra)
fviz_nbclust(MatonHL_clos[,2:49], kmeans, method = "wss") +
  geom_vline(xintercept = 4, linetype = 2)
```

```{r assess optimum clusters for open (CLR) data}
require(factoextra)
fviz_nbclust(MatonHL_open[,2:49], kmeans, method = "wss") +
  geom_vline(xintercept = 4, linetype = 2)
```

# compute K-means

```{r}
set.seed(123)
MatonHL_clos_kmeans <- kmeans(MatonHL_clos[,2:49], 4, nstart = 25)
cat("K-means clustering with",length(MatonHL_clos_kmeans$size),"clusters of sizes",MatonHL_clos_kmeans$size,"\n")
cat("\ncomponents of output object are:\n")
ls(MatonHL_clos_kmeans)
```

# plot kmeans clusters

```{r visualize kmeans clusters, fig.height=6, fig.width=6, message=FALSE, warning=FALSE}
fviz_cluster(MatonHL_clos_kmeans, data = MatonHL_clos[,2:49],
             palette = c("#2E9FDF", "#00AFBB", "#E7B800", "#FC4E07"),
             labelsize=10,
             ellipse.type = "euclid", # Concentration ellipse
             star.plot = TRUE, # Add segments from centroids to items
             repel = F, # if true avoids label overplotting (slow)
             ggtheme = theme_minimal()
             )
```

```{r}
set.seed(123)
MatonHL_open_kmeans <- kmeans(MatonHL_open[,2:49], 4, nstart = 25)
cat("K-means clustering with",length(MatonHL_open_kmeans$size),"clusters of sizes",MatonHL_open_kmeans$size,"\n")
cat("\ncomponents of output object are:\n")
ls(MatonHL_open_kmeans)
```

```{r visualize kmeans clusters, fig.height=7, fig.width=7, message=FALSE, warning=FALSE}
fviz_cluster(MatonHL_open_kmeans, data = MatonHL_open[,2:49],
             palette = c("#2E9FDF", "#00AFBB", "#E7B800", "#FC4E07"),
             labelsize=10,
             ellipse.type = "euclid", # Concentration ellipse
             star.plot = TRUE, # Add segments from centroids to items
             repel = F, # if true avoids label overplotting (slow)
             ggtheme = theme_minimal()
             )
```

# hierarchical clustering

## create dissimilarity (distance) matrix

```{r distance matrix closed}
MatonHL_clos_diss <- get_dist(MatonHL_clos[,2:49], method = "kendall")
as.matrix(MatonHL_clos_diss)[1:8, 1:8]
```

## perform hierarchical clustering

```{r hierarchical clustering closed, message=FALSE, warning=FALSE}
MatonHL_clos_hc <- hclust(MatonHL_clos_diss, method = "ward.D2")
require(factoextra)
fviz_dend(MatonHL_clos_hc, cex = 0.5)
```

## assess cluster tree

```{r verify cluster tree closed}
MatonHL_clos_coph <- cophenetic(MatonHL_clos_hc)
cor(MatonHL_clos_diss,MatonHL_clos_coph)
cat("\nRule-of-thumb: cluster tree represents actual distance matrix accurately enough if r>0.75\n")
```

## cut dendrogram into different groups

```{r cut dendrogram closed}
MatonHL_clos_grp <- cutree(MatonHL_clos_hc, k = 4)
MatonHL_clos_grp ; cat("\n")
table(MatonHL_clos_grp)
```

## plot dendrogram with cuts

```{r plot cut dendrogram closed, message=FALSE, warning=FALSE}
fviz_dend(MatonHL_clos_hc, k = 4, # Cut in four groups
          cex = 0.5, # label size
          k_colors = c("red3", "darkgreen", "blue2", "purple"),
          color_labels_by_k = TRUE, # color labels by groups
          rect = TRUE # Add rectangle around groups
          )
```

## create dissimilarity (distance) matrix (open CLR)

```{r distance matrix openCLR}
MatonHL_open_diss <- get_dist(MatonHL_open[,2:49], method = "kendall")
as.matrix(MatonHL_open_diss)[1:8, 1:8]
```

## perform hierarchical clustering (open CLR)

```{r hierarchical clustering openCLR, message=FALSE, warning=FALSE}
MatonHL_open_hc <- hclust(MatonHL_open_diss, method = "ward.D2")
require(factoextra)
fviz_dend(MatonHL_open_hc, cex = 0.5)
```

## assess cluster tree (open CLR)

```{r verify cluster tree openCLR}
MatonHL_open_coph <- cophenetic(MatonHL_open_hc)
cor(MatonHL_open_diss,MatonHL_open_coph)
cat("\nRule-of-thumb: cluster tree represents actual distance matrix accurately enough if r>0.75\n")
```

## try another linkage method for clustering

```{r}
MatonHL_open_hc <- hclust(MatonHL_open_diss, method = "average")
MatonHL_open_coph <- cophenetic(MatonHL_open_hc)
cor(MatonHL_open_diss,MatonHL_open_coph)
cat("\nRule-of-thumb: cluster tree represents actual distance matrix accurately enough if r>0.75\n")
```

## cut dendrogram into different groups (open CLR)

```{r cut dendrogram openCLR}
MatonHL_open_grp <- cutree(MatonHL_open_hc, k = 3)
table(MatonHL_open_grp)
```

## plot dendrogram with cuts (open CLR)

```{r plot cut dendrogram openCLR, fig.height=5, fig.width=9, message=FALSE, warning=FALSE}
fviz_dend(MatonHL_open_hc, k = 4, # Cut in four groups
          cex = 0.5, # label size
          k_colors = rainbow(4, v=0.6, start = 0, end = 0.75),
          color_labels_by_k = TRUE, # color labels by groups
          rect = TRUE # Add rectangle around groups
)
```

# correlation matrices
```{r fig.height=10, fig.width=10}
require(DataExplorer)
plot_correlation(MatonHL_clos[,2:49], cor_args = list("use" = "pairwise.complete.obs"))
```

# correlation matrices (CLR)
```{r correlation matrix clr, fig.height=10, fig.width=10}
require(DataExplorer)
plot_correlation(MatonHL_open[,2:49], cor_args = list("use" = "pairwise.complete.obs"))
```
