---
title: "testing multivariate analyses on SLCVR data"
subtitle: "SLCVR refers to Smith's Lake and Charles Veryard Reserves, Perth, Western Australia"
documentclass: article
classoption: a4paper
geometry: margin=0.787in
output: 
  pdf_document: 
  highlight: espresso
  fig_height: "50%"
  fig_width: "50%"
  fig_caption: TRUE
  number_sections: TRUE
  toc: TRUE
---

# load packages

```{r load packages etc., include=FALSE}
library(stringr)
library(RcmdrMisc)
library(rgr)
library(car)
library(cluster)
library(factoextra)
library(DataExplorer)

palette(c("black","blue","green4","red2","purple",
          "darkcyan","firebrick","grey","grey40","white",
          "transparent"))
```

# load data

**Data are from**: Rate, A.W., 2021. Smith's Lake and Charles Veryard Reserves 2017 multielement soil, sediment, street dust, *PANGAEA*, doi:10.1594/PANGAEA.935591 https://doi.org/10.1594/PANGAEA.935591.

```{r}
slcvr <- read.csv("SLCVR_2021.csv", stringsAsFactors = TRUE)
# str(slcvr)
slcvr$Code <- as.character(slcvr$Code)
slcvr$Sample.ID <- as.character(slcvr$Sample.ID)
slcvr$Group <- as.factor(slcvr$Group)
```

# make CLR-transformed dataset
```{r}
slcvr_clr <- slcvr
temp <- clr(slcvr_clr[,c("pH","Al","As","Ba","Ca","Ce","Cr","Cu","Fe","Gd","K","La",
                                      "Mg","Mn","Mo","Na","Ni","P","Pb","S","Sr","V","Y","Zn")])
rowsNA <- as.numeric(attributes(temp)$na.action)
slcvr_clr <- cbind(slcvr[-rowsNA,1:13], temp)
numSummary(slcvr_clr[,c("pH","Al","As","Ba","Ca","Ce","Cr","Cu","Fe","Gd","K","La",
                                      "Mg","Mn","Mo","Na","Ni","P","Pb","S","Sr","V","Y","Zn")])
names(slcvr);names(slcvr_clr)
```

# correlation matrices
```{r fig.height=7, fig.width=7}
require(DataExplorer)
plot_correlation(slcvr[,c("pH","EC","Al","As","Ba","Ca","Ce","Cr","Cu","Fe","Gd","K","La",
                          "Mg","Mn","Mo","Na","Ni","P","Pb","S","Sr","V","Y","Zn")],
                 cor_args = list("use" = "pairwise.complete.obs"))
```

# correlation matrices (CLR)
```{r correlation matrix clr, fig.height=7, fig.width=7}
require(DataExplorer)
plot_correlation(slcvr_clr[,c("pH","EC","Al","As","Ba","Ca","Ce","Cr","Cu","Fe","Gd","K","La",
                              "Mg","Mn","Mo","Na","Ni","P","Pb","S","Sr","V","Y","Zn")],
                 cor_args = list("use" = "pairwise.complete.obs"))
```

# principal components analysis

```{r run and summarise PCA}
data0 <- na.omit(slcvr[,c("pH","Al","As","Ba","Ca","Ce","Cr","Cu","Fe","Gd","K","La",
                              "Mg","Mn","Mo","Na","Ni","P","Pb","S","Sr","V","Y","Zn")])
pca_slcvr_clos <- prcomp(data0, scale. = TRUE)
pca_slcvr_clos$rot[,1:6]
cat("...\n\nComponent Variances\n")
pca_slcvr_clos$sdev[1:10]^2
cat("\n--------------------\n")

data0 <- na.omit(slcvr_clr[,c("pH","Al","As","Ba","Ca","Ce","Cr","Cu","Fe","Gd","K","La",
                              "Mg","Mn","Mo","Na","Ni","P","Pb","S","Sr","V","Y","Zn")])
pca_slcvr_open <- prcomp(data0, scale. = TRUE)
pca_slcvr_open$rot[,1:6]
cat("...\n\nComponent Variances\n")
pca_slcvr_open$sdev[1:10]^2
cat("\n--------------------\n")
rm(data0)
```

```{r visualise PCA, fig.height=6, fig.width=12}
require(car)
palette(c("black","blue","green4","red2","purple",
          "darkcyan","firebrick","grey","grey40","white",
          "transparent"))
par(mfrow=c(1,2), mar = c(3.5,3.5,3.5,3.5), oma = c(0,0,0,0), 
    mgp=c(1.7,0.3,0), tcl = 0.25, font.lab=2,
    lend = "square", ljoin = "mitre")
# choose components and set scaling factor (sf)
v1 <- 1
v2 <- 2
sf <- 0.12

biplot(pca_slcvr_clos, choices = c(v1,v2), col = c(11,9), cex=c(1,0.65),
       pc.biplot = FALSE, scale = 0.4, arrow.len = 0.08,
       xlab = paste0("Scaled PC",v1," Component Loadings"),
       ylab = paste0("Scaled PC",v2," Component Loadings"))
mtext(paste0("Scaled PC",v1," Observation Scores"), 3, 1.6, font = 2)
mtext(paste0("Scaled PC",v2," Observation Scores"), 4, 1.6, font = 2)
mtext("Untransformed data\n(compositionally closed)",
      side = 3, line = -2, font = 2, adj = 0.98)
data0 <- na.omit(slcvr[,c("Type","pH","Al","As","Ba","Ca","Ce","Cr","Cu","Fe","Gd","K","La",
                              "Mg","Mn","Mo","Na","Ni","P","Pb","S","Sr","V","Y","Zn")])
points(pca_slcvr_clos$x[,v1]*sf, pca_slcvr_clos$x[,v2]*sf*1.5,
       pch = c(22,21,24)[data0$Type],
       bg = c(2,3,4)[data0$Type],
       col = c(2,3,4)[data0$Type],
       cex = c(1.2,1.4,1.2)[data0$Type])
dataEllipse(x=pca_slcvr_clos$x[,v1]*sf, y=pca_slcvr_clos$x[,v2]*sf*1.5,
            groups = data0$Type, add = TRUE,
            plot.points = FALSE, levels = c(0.9),
            center.pch = 3, col = c(2,3,4,5,6,7),
            lty = 2, lwd = 1, center.cex = 2.4)
symbols(0,0, circles = 0.5, fg = "grey",
        add = TRUE, inches = FALSE)
legend("bottomleft", bty = "o", inset = 0.03,
       title = "Type",
       box.col = "gray", box.lwd = 2, bg = 10,
       legend = levels(data0$Type),
       pch = c(22,21,24),
       col = c(2,3,4), pt.bg = c(2,3,4),
       pt.cex = c(1.2, 1.4, 1.2),
       cex = 1.1, y.intersp = 0.9)

# v1 <- 1
# v2 <- 3
sf <- 0.15

biplot(pca_slcvr_open, choices = c(v1,v2), col = c(11,9), cex=c(1,0.65),
       pc.biplot = FALSE, scale = 0.5, arrow.len = 0.08,
xlab = paste0("Scaled PC",v1," Component Loadings"),
       ylab = paste0("Scaled PC",v2," Component Loadings"))
mtext(paste0("Scaled PC",v1," Observation Scores"), 3, 1.6, font = 2)
mtext(paste0("Scaled PC",v2," Observation Scores"), 4, 1.6, font = 2)
mtext("CLR transformed data\n(no closure)",
      side = 3, line = -2, font = 2, adj = 0.02)
data0 <- na.omit(slcvr_clr[,c("Type","pH","Al","As","Ba","Ca","Ce","Cr","Cu","Fe","Gd","K","La",
                              "Mg","Mn","Mo","Na","Ni","P","Pb","S","Sr","V","Y","Zn")])
points(pca_slcvr_open$x[,v1]*sf, pca_slcvr_open$x[,v2]*sf*1.5,
       pch = c(22,21,24)[data0$Type],
       bg = c(2,3,4)[data0$Type],
       col = c(2,3,4)[data0$Type],
       cex = c(1.2,1.4,1.2)[data0$Type])
dataEllipse(x=pca_slcvr_open$x[,v1]*sf, y=pca_slcvr_open$x[,v2]*sf*1.5,
            groups = data0$Type, add = TRUE,
            plot.points = FALSE, levels = c(0.9),
            center.pch = 3, col = c(2,3,4,5,6,7),
            lty = 2, lwd = 1, center.cex = 2.4)
symbols(0,0, circles = 0.5, fg = "grey",
        add = TRUE, inches = FALSE)
legend("bottomleft", bty = "o", inset = 0.03,
       title = "Type",
       box.col = "gray", box.lwd = 2, bg = 10,
       legend = levels(data0$Type),
       pch = c(22,21,24),
       col = c(2,3,4), pt.bg = c(2,3,4),
       pt.cex = c(1.2, 1.4, 1.2),
       cex = 1.1, y.intersp = 0.9)
rm(list = c("v1","v2","sf","data0"))
```


# K-means clustering

```{r assess optimum clusters for closed data, fig.height=3.5, fig.width=5}
require(factoextra)
data0 <- na.omit(slcvr[,c("pH","Al","As","Ba","Ca","Ce","Cr","Cu","Fe","Gd","K","La",
                              "Mg","Mn","Mo","Na","Ni","P","Pb","S","Sr","V","Y","Zn")])
fviz_nbclust(data0, kmeans, method = "wss") +
  geom_vline(xintercept = 4, linetype = 2)
```

```{r assess optimum clusters for CLR data, fig.height=3.5, fig.width=5}
require(factoextra)
data0 <- na.omit(slcvr_clr[,c("pH","Al","As","Ba","Ca","Ce","Cr","Cu","Fe","Gd","K","La",
                              "Mg","Mn","Mo","Na","Ni","P","Pb","S","Sr","V","Y","Zn")])
fviz_nbclust(data0, kmeans, method = "wss") +
  geom_vline(xintercept = 4, linetype = 2)
```

# compute K-means closed

```{r}
data0 <- na.omit(slcvr[,c("Type","pH","Al","As","Ba","Ca","Ce","Cr","Cu","Fe","Gd","K","La",
                              "Mg","Mn","Mo","Na","Ni","P","Pb","S","Sr","V","Y","Zn")])
set.seed(123)
slcvr_clos_kmeans <- kmeans(data0[,2:NCOL(data0)], 3, nstart = 25)
cat("K-means clustering with",length(slcvr_clos_kmeans$size),"clusters of sizes",slcvr_clos_kmeans$size,"\n")
cat("\ncomponents of output object are:\n")
ls(slcvr_clos_kmeans)
```

# plot kmeans clusters closed

```{r visualize kmeans clusters, fig.height=6, fig.width=6, message=FALSE, warning=FALSE}
row.names(data0) <- paste0(data0$Type,seq(1,NROW(data0)))
fviz_cluster(slcvr_clos_kmeans, data = data0[,2:NCOL(data0)],
             palette = c("#00AFBB", "#E7B800", "#FC4E07"),
             labelsize=10,
             ellipse.type = "euclid", # Concentration ellipse
             star.plot = TRUE, # Add segments from centroids to items
             repel = F, # if true avoids label overplotting (slow)
             ggtheme = theme_minimal()
             )
```

# compute K-means open

```{r}
data0 <- na.omit(slcvr_clr[,c("Type","pH","Al","As","Ba","Ca","Ce","Cr","Cu","Fe","Gd","K","La",
                              "Mg","Mn","Mo","Na","Ni","P","Pb","S","Sr","V","Y","Zn")])
set.seed(123)
slcvr_open_kmeans <- kmeans(data0[,2:NCOL(data0)], 3, nstart = 25)
cat("K-means clustering with",length(slcvr_open_kmeans$size),"clusters of sizes",slcvr_open_kmeans$size,"\n")
cat("\ncomponents of output object are:\n")
ls(slcvr_open_kmeans)
```

# plot kmeans clusters open

```{r visualize kmeans clusters, fig.height=6, fig.width=6, message=FALSE, warning=FALSE}
row.names(data0) <- paste0(data0$Type,seq(1,NROW(data0)))
fviz_cluster(slcvr_open_kmeans, data = data0[,2:NCOL(data0)],
             palette = c("#00AFBB", "#E7B800", "#FC4E07"),
             labelsize=10,
             ellipse.type = "euclid", # Concentration ellipse
             star.plot = TRUE, # Add segments from centroids to items
             repel = F, # if true avoids label overplotting (slow)
             ggtheme = theme_minimal()
             )
```

# hierarchical clustering

## create dissimilarity (distance) matrix

```{r distance matrix closed}
dataHC <- na.omit(slcvr[,c("Type","pH","Al","As","Ba","Ca","Ce","Cr","Cu","Fe","Gd","K","La",
                              "Mg","Mn","Mo","Na","Ni","P","Pb","S","Sr","V","Y","Zn")])
row.names(dataHC) <- paste0(dataHC$Type, seq(1,NROW(dataHC)))
dataHC$Type <- NULL
slcvr_clos_diss <- get_dist(dataHC, method = "euclidean")
as.matrix(slcvr_clos_diss)[1:8, 1:8]
```

## perform hierarchical clustering

```{r hierarchical clustering closed, fig.height=4.5, fig.width=10, message=FALSE, warning=FALSE}
slcvr_clos_hc <- hclust(slcvr_clos_diss, method = "ward.D2")
require(factoextra)
fviz_dend(slcvr_clos_hc, cex = 0.5)
```

## assess cluster tree

```{r verify cluster tree closed}
slcvr_clos_coph <- cophenetic(slcvr_clos_hc)
cor(slcvr_clos_diss,slcvr_clos_coph)
cat("\nRule-of-thumb: cluster tree represents actual distance matrix accurately enough if r>0.75\n")
```

## cut dendrogram into different groups

```{r cut dendrogram closed}
slcvr_clos_grp <- cutree(slcvr_clos_hc, k = 3)
slcvr_clos_grp ; cat("\n")
table(slcvr_clos_grp)
```

## plot dendrogram with cuts

```{r plot cut dendrogram closed, fig.height=4.5, fig.width=10, message=FALSE, warning=FALSE}
fviz_dend(slcvr_clos_hc, k = 3, # Cut in four groups
          cex = 0.5, # label size
          k_colors = c("red3", "blue2", "purple"),
          color_labels_by_k = TRUE, # color labels by groups
          rect = TRUE # Add rectangle around groups
          )
```

# hierarchical clustering (CLR-open data)

## create dissimilarity (distance) matrix

```{r distance matrix open}
dataHC <- na.omit(slcvr_clr[,c("Type","pH","Al","As","Ba","Ca","Ce","Cr","Cu","Fe","Gd","K","La",
                           "Mg","Mn","Mo","Na","Ni","P","Pb","S","Sr","V","Y","Zn")])
row.names(dataHC) <- paste0(dataHC$Type, seq(1,NROW(dataHC)))
dataHC$Type <- NULL
slcvr_open_diss <- get_dist(dataHC, method = "euclidean")
as.matrix(slcvr_open_diss)[1:8, 1:8]
```

## perform hierarchical clustering

```{r hierarchical clustering open, fig.height=4.5, fig.width=10, message=FALSE, warning=FALSE}
slcvr_open_hc <- hclust(slcvr_open_diss, method = "ward.D2")
require(factoextra)
fviz_dend(slcvr_open_hc, cex = 0.5)
```

## assess cluster tree

```{r verify cluster tree open}
slcvr_open_coph <- cophenetic(slcvr_open_hc)
cor(slcvr_open_diss,slcvr_open_coph)
cat("\nRule-of-thumb: cluster tree represents actual distance matrix accurately enough if r>0.75\n")
```

## cut dendrogram into different groups

```{r cut dendrogram open}
slcvr_open_grp <- cutree(slcvr_open_hc, k = 3)
slcvr_open_grp ; cat("\n")
table(slcvr_open_grp)
```

## plot dendrogram with cuts

```{r plot cut dendrogram open, fig.height=4.5, fig.width=10, message=FALSE, warning=FALSE}
fviz_dend(slcvr_open_hc, k = 3, # Cut in four groups
          cex = 0.5, # label size
          k_colors = c("red3", "blue2", "purple"),
          color_labels_by_k = TRUE, # color labels by groups
          rect = TRUE # Add rectangle around groups
)
```

